# Survival of the fraudulent: Cross-national patterns of manipulation and protest outcome
\chaptermark{Survival of the fraudulent}

<!--
Setup:

1) Electoral manipulation is understood to produce legitimacy costs, which can lead to protest
2) Protest is costly to governments, and so they may restrain themselves
  a) This is especially true if manipulation is an indicator of regime weakness
3) But there are lots of reasons to think that electoral manipulation indicates strength (though it may still entail legitimacy costs)
4) If that is the case, we should expect to see cross national patterns where manipulation yields protest (legitimacy costs) but not risk to incumbents (more manipulation = more staying power despite risk)

-->


```{r setup, include = FALSE}
library(tidyverse)
library(countrycode)
library(readxl)
library(lme4)
library(interplot)
library(lubridate)
library(summarytools)
library(stargazer)
library(ggeffects)
#library(glmmTMB)
library(modelsummary)
library(huxtable)
library(simPH)
library(broom)
library(broom.helpers)
library(kableExtra)
library(gt)

nelda2.sub <- read.csv(here::here("Crossnational_data", "nelda2_sub_vdem15_2025-05-29.csv"))


nelda2.sub <- nelda2.sub %>% mutate(legit_no_proc = (v2exl_legitperf + v2exl_legitlead + v2exl_legitideol))
nelda2.sub <- nelda2.sub %>% mutate(legit_total = (v2exl_legitperf + v2exl_legitlead + v2exl_legitideol + v2exl_legitratio))



##Naming nelda variables

nelda2.sub <- nelda2.sub %>% rename(boycott = nelda14)
nelda2.sub <- nelda2.sub %>% mutate(boycott = ifelse(boycott == "yes", 1, ifelse(boycott == "no", 0,
                                                                                 NA)))

##Creating economic crisis variable

nelda2.sub <- nelda2.sub %>% mutate(econ.crisis = ifelse(is.na(e_miinflat) == T | is.na(gdpgrowth.1lag) == T, NA, ifelse(e_miinflat > 15 | gdpgrowth.1lag <= -0.012, 1, 0)))
nelda2.sub <- nelda2.sub %>% mutate(econ.crisis = as.factor(econ.crisis))

nelda2.sub <- nelda2.sub %>% mutate(gdp.crisis = ifelse(is.na(gdpgrowth.1lag) == T, NA, ifelse(gdpgrowth.1lag <= -0.012, 1, 0)))
nelda2.sub <- nelda2.sub %>% mutate(inflat.crisis = ifelse(is.na(e_miinflat) == T, NA, ifelse(e_miinflat > 15, 1, 0)))

nelda2.sub <- nelda2.sub %>% mutate(grievance.index = ifelse(is.na(gdppc.1lag) == T, NA, gdppc.1lag*v2xeg_eqdr))



## Removing cases where inc. loses
#nelda2.sub <- nelda2.sub %>% filter(inc.lose == 0)  #See how this affects things

##Intmon as a factor
nelda2.sub <- nelda2.sub %>% mutate(v2elintmon = as.factor(v2elintmon))


nelda2.sub <- nelda2.sub %>% mutate(electoral.dem = ifelse(v2x_regime == 2, 1, 0))

#Square of physint

nelda2.sub <- nelda2.sub %>% mutate(physintlag.sq = physinteg.1lag^2)

## Categorical variable for incumbent status

nelda2.sub <- nelda2.sub %>% mutate(incumb.status = as.factor(ifelse(inc.termlimit == "yes" & chosen.successor == 1, "Chosen successor",
                                                           ifelse(inc.termlimit == "yes" & chosen.successor == 0, "Contested exec. election", 
                                                                  ifelse(inc.termlimit == "no", "Executive eligible", "Legislative election"))))) 
nelda2.sub <- nelda2.sub %>% mutate(incumb.status = relevel(incumb.status, ref = "Legislative election"))


#Coding as factors

nelda2.sub <- nelda2.sub %>% mutate(presidential = factor(presidential))
#nelda2.sub <- nelda2.sub %>% mutate(inc.termlimit.num = factor(inc.termlimit.num))
nelda2.sub <- nelda2.sub %>% mutate(rentierstate = factor(rentierstate))
nelda2.sub <- nelda2.sub %>% mutate(taxstate = factor(taxstate))
nelda2.sub <- nelda2.sub %>% mutate(unscheduled_election = factor(unscheduled_election))
nelda2.sub <- nelda2.sub %>% mutate(inc.lose = factor(inc.lose))
nelda2.sub <- nelda2.sub %>% mutate(econ.crisis.nelda.1lag = factor(econ.crisis.nelda.1lag))


#Successful protest = new elections or new incumbent

nelda2.sub <- nelda2.sub %>% mutate(protest.success = ifelse(inc.replaced.prot == 1 | new.elections.held == 1, 1, 0))



# Replacing NA with 0 for term limits variable in known parliamentary countries in legislative elections

nelda2.sub <- nelda2.sub %>% mutate(inc.termlimit.num = ifelse(is.na(inc.termlimit.num) == T & v2ex_elechog == 0, 0, inc.termlimit.num))

## Replacing NA with 0 for protest size

#nelda2.sub <- nelda2.sub %>% mutate(max_size = ifelse(is.na(max_size)==T & n.events.post == 0, 0, max_size))

## Fixing missing values of winner.share.upd for liberal democracies

nelda2.sub <- nelda2.sub %>% mutate(winner.share.upd = ifelse(is.na(winner.share.upd)==T & v2x_regime == 3, winner.margin, winner.share.upd))

## Set max_size zero for elections with no protest

#nelda2.sub <- nelda2.sub %>% mutate(max_size = ifelse(n.events.post == 0, 0, max_size))

## Excluding liberal democracies

nelda2.sub <- nelda2.sub %>% filter(v2x_regime == 1 | v2x_regime == 2 ) #Includes electoral authoritarian, and e-dems with 

## Remove outlier
nelda2.sub <- nelda2.sub %>% filter(d.events.num <= 100)

#Protest cases only

protest.cases <- nelda2.sub %>% filter(any.ecav.protest == 1)

## Coding median size as 0 when n.events = 0

nelda2.sub <- nelda2.sub %>% mutate(median.participants = ifelse(n.events.post == 0, 0, median.participants))

## Adding other dimensions of manipulation

nelda2.sub <- nelda2.sub %>% mutate(v2elintim.inv = v2elintim * -1)
nelda2.sub <- nelda2.sub %>% mutate(v2elvotbuy.inv = v2elvotbuy * -1)
nelda2.sub <- nelda2.sub %>% mutate(emb_control = v2elembaut * -1)


nelda2.sub <- nelda2.sub %>% mutate(illegal_man_index = v2elintim.inv + v2elvotbuy.inv + v2elirreg.inv)


## All manipulation index

nelda2.sub <- nelda2.sub %>% mutate(manipulation_index_add = v2elintim.inv + v2elirreg.inv + v2elvotbuy.inv)

rescale_to_zero <- function(x) {
  x - min(x, na.rm = TRUE)
}

nelda2.sub <- nelda2.sub %>% mutate(elintim.inv.zeroscale = rescale_to_zero(v2elintim.inv))
nelda2.sub <- nelda2.sub %>% mutate(elirreg.inv.zeroscale = rescale_to_zero(v2elirreg.inv))
nelda2.sub <- nelda2.sub %>% mutate(elvotbuy.inv.zeroscale = rescale_to_zero(v2elvotbuy.inv))


nelda2.sub <- nelda2.sub %>% mutate(manipulation_index_mult = (elintim.inv.zeroscale*elvotbuy.inv.zeroscale)+elirreg.inv.zeroscale)

## For descriptive statistics
nelda2.sub.auth <- nelda2.sub %>% filter(v2x_regime == "1")
```


Beginning with the 'Bulldozer Revolution' in Serbia, and ending with the 2009 electoral protests in Moldova, a series of mass protest events succeeded in ousting incumbent governments across the post-communist region. These 'Color Revolutions' (though only one, Ukraine's Orange Revolution, is technically named after a color) often captivated public attention, with dramatic scenes of sustained mass protest, sometimes in the face of harsh repression. They also drew considerable attention of scholars of democracy and authoritarianism, coming as they did in the second decade after the fall of the Soviet Union brought about a 'third wave' of democracy. 

Whether driven by effective opposition strategy [@bunce2010] or structural factors [@LucanWay2008;@hale_democracy_2006], these contentious elections appeared to highlight a significant risk that incumbents accept when they manipulate elections. Though far from certain, egregious manipulation could result in mass protests that are costly to the regime, regardless of their outcome. In the worst case, as in Georgia (2003), Ukraine (2004), Kyrgyzstan (2005) and Moldova (2009), incumbents may find themselves ousted. Even in the case of a 'failed revolution,' as in Russia (2011-2012) or Belarus (2020-2021) incumbents may find themselves paying considerable resource costs, experiencing international repercussions, making concessions to the opposition, and/or reallocating the distribution of rents to elites and the security services in order to survive protest outbreaks.

Events like these play a central role in theories about election manipulation and how it might be curbed — with broader implications for the stability of both democratic and authoritarian regimes.
A classic view, influential in formal modeling, holds that when an incumbent resorts to election manipulation, it signals their unpopularity. That lack of public support, in turn, makes it harder for them to suppress a protest movement if the opposition decides to mobilize. Models in this tradition predict that some incumbents need to cheat to win reliably, but — fearing the backlash of costly protests — choose instead to take their chances with a relatively clean election. This framework carries an optimistic message: stronger opposition movements, meaning those that are better organized, better funded, and more popular, are more capable of deterring incumbents from rigging elections in the first place. That dynamic, in turn, makes democratic transition and consolidation more likely. I call this the regime-weakness model of manipulation.

A competing view has emerged in recent years. Developed most prominently by Simpser (2013), it argues that election manipulation is actually a costly signal of the incumbent's power and resources. By manipulating elections, incumbents demonstrate their financial muscle, patronage networks, organizational capacity, and ability to bend or break the law. On this account, incumbents who manipulate elections effectively are not revealing weakness through unpopularity — they are displaying alternative sources of strength that can offset low public approval, and that can equally be deployed to suppress protest movements after the election. I call this the regime-strength model.

At first glance, testing these competing models might seem straightforward. The regime-weakness model generally predicts that more manipulation — holding the incumbent's true popularity constant — makes it more likely that opposition groups will launch a post-election protest (Luo and Rozenas 2017; Little 2012; Schedler 2002; Beaulieu 2014). The most telling signal of incumbent weakness is a low vote-share achieved through heavy manipulation: by relying heavily on 'cheap-talk' manipulation and still posting a poor result, the incumbent demonstrates that its popularity is very low and that its chances of defeating a protest movement are slim. The regime-strength model, by contrast, predicts the opposite (Simpser 2013): by demonstrating resourcefulness and impunity, more manipulation should make protest less likely (holding vote-share constant). Under this model, the conditions most favorable for protest initiation are a low incumbent vote-share combined with a low level of manipulation — a combination that signals both weak natural popularity and limited resources.

There are two problems with this neat picture, however. First, the empirical evidence on the relationship between manipulation and protest has been mixed, revealing the limits of existing theory (Ananyev and Poyker 2022; Daxecker 2012; Hafner-Burton, Hyde, and Jablonski 2016; Hyde and Marinov 2014; Lankina and Skovoroda 2017). Second, the expected relationship between manipulation and protest under either model is complicated by a third framework: a more complex, grievance-based model of electoral protest. This view, which has substantial empirical support, holds that electoral protests are driven primarily by citizen grievances — not by the narrowly strategic calculations about the odds of toppling the government that the regime-weakness model assumes. Under the grievance framework, a heavily manipulated election can serve as a focal point for collective action when citizens harbor widespread discontent toward the incumbent (Tucker 2007). Combined with opposition elites' various incentives to stage or join protests (e.g. generating publicity, waging factional battles, costly signaling in bargaining for policy change, etc.), this means that mass electoral protests may occur even when they have little chance of unseating a strong incumbent.

This theoretical tangle makes it difficult to distinguish between the regime-strength and regime-weakness models when looking at protest initiation as a dependent variable. Depending on which model of protest behavior is correct, the two models of manipulation can generate identical predictions. If protest movements are instrumentally rational and manipulation signals weakness, we would expect the probability of protest to rise with the level of manipulation. But if manipulation instead indicates regime strength, we might reach the same expectation — provided that mass protests are driven by grievance rather than instrumental calculation. In that model, protest initiation and extensive manipulation are correlated, but the relationship is confounded by the incumbent's low popularity. A strong-but-unpopular incumbent must manipulate heavily to win the election, while protests are likely to occur due to that same unpopularity (not because the opposition is exploiting a signal of weakness). Because a single outcome is consistent with two different theories, we are left at an impasse if protest initiation is the variable of interest.

 

To move the debate forward, I model electoral protest *duration* as a dependent variable, in addition to protest occurrence. The regime-weakness and regime-strength models produce distinct predictions with regard to protest duration, unlike with regard to protest initiation. Once protests begin, the 'strength' of an incumbent (that is, its likelihood of defeating the opposition movement) is no longer latent. Instead, the incumbent's response and its survival (or not) reveal the incumbents' true strength relative to the opposition. The more quickly a government is able to restore order, the greater the underlying (unobserved) distribution of power can be assumed to favor the incumbent. By contrast, the longer the government struggles to put down the protest movement, the less favorable the distribution of power.


Using data from `r nrow(nelda2.sub.auth)` elections across `r length(unique(nelda2.sub.auth$country))` electoral authoritarian countries, from 1989 to 2012, the primary analysis in this chapter shows that governments that engage in more manipulation are likely to put down protests more quickly than governments that produce little manipulation, especially when incumbents claim small vote-shares. This result is consistent with a regime-strength model of electoral manipulation, in which the same incumbents that are best equipped to rig elections are also best equipped to survey extra-electoral challenges. The results also show that protest is most likely to be initiated when manipulation is severe but incumbent vote-share is low. Additional analyses make use of a fuller dataset which includes electoral democratic country-years, bringing the total number of cases to `r nrow(nelda2.sub)`.  

In combination, this pattern is consistent with a world in which extensive election manipulation is indicative of incumbent resilience, while protest initiation is often not motivated by narrowly strategic goals. That is, protesters may take to the streets even when they have rationally low expectations of unseating the incumbent,  in pursuit of broader goals and/or to express their anger and sense of grievance. In light of this pattern, it is likely that the role of electoral protest as a deterrent to manipulation has been overemphasized in prior theories. 

This interpretation offers a unified explanation for several secular patterns: the infrequency of electoral protests [@Brancati2016], the limited success of such protests in toppling incumbent regimes, and the tendency of governments to report implausibly large victory margins in heavily rigged elections. For theories of electoral manipulation, it implies that more attention should be given to alternative deterrents to manipulation (such as the role of risk to low-level actors who implement manipulation, as explored later in this volume). Ultimately, this perspective points to a more pessimistic view of democratization through elections in authoritarian contexts, and for the logic of 'self-enforcing democracy.'




## Theory: Developing the regime-strength model of electoral manipulation

\sectionmark{Theory}

Under what circumstances should protest movements be expected to last longer in authoritarian regimes? Authoritarian incumbents have strong incentives to keep protests small and short-lived. Generally speaking, by definition, authoritarian regimes have a smaller winning coalition and selectorate than democracies do, and allocate more private and club goods to elites as a result [@BuenodeMesquita2005]. Despite this, they can appear superficially popular, with most citizens falsifying their preferences and demonstrating support for the regime, tacitly or otherwise. The appearance of even a moderate number of protesters publicly signalling their disapproval of the government can trigger a cascade of dissent in short order, as other citizens stop falsifying their preferences as well [@Kuran1991]. Given time, persistent protests can lead elites and the security services to update their prior beliefs about the strength of the incumbent, prompting defections [@svolik_politics_2012;@Chenoweth_Stephan2011]. Protest movements can also challenge the regime's legitimacy [@keremouglu2022thin], eroding individual compliance in other areas. To prevent all these negative outcomes, incumbents try to prevent mass protest from occurring, and to quickly quell those protest movements that do arise.

Assuming the incumbent remains in office, stronger protest movements will tend to last longer than weaker ones. Ordinary protesters are more likely to remain engaged when they are more aggrieved [@gurrWhyMenRebel2015], when they are angrier, and when they observe many other people expressing their discontent; extensive public grievance can thus be a source of durability for protest movements. The longer-lasting and larger protest movements are, the more protesters may become socially networked, enabling easier subsequent mobilization [@schussmanProcessProtestAccounting2005;@sheldonIdentityCompetitionKey2000], increasing organizational resilience and capacity . When group goals appear more achievable---in part because the protest movement is large, well-organized, and well-resourced, and the incumbent appears off balance---individuals we will be more willing to participate in protest [@Klandermans1997] and to become regulars [@saundersExplainingDifferentialProtest2012].  Strategic protest entrepreneurs, who are using protest as a form of costly bargaining against the incumbent, will aim to demonstrate the resilience and popularity of their movements in order to present a stronger negotiating position [@robertson_protest_2010;@Beaulieu2014a]. As a result, they often seek to maintain high levels of mobilization even in the face of repression and ordinary citizens' waning interest over time [@tillyContentiousPolitics2015]. On average, then, I expect that authoritarian and proto-authoritarian incumbents will seek to demobilize protest, and protest movement leaders will seek to maintain (or even increase) high levels of mobilization. Which side prevails, and how quickly, depends on the relative strength of each.

Simpser's foundational argument in the regime-strength model holds that manipulated elections can send costly signals of the incumbent's "capacities, resources, [and] dispositions" that will help it retain power in the future [@simpser2013, p. 80]\textemdash that is, of its political strength. Additional research focusing on the principal-agent dynamics inherent in election manipulation reinforces the idea that manipulation carries such costly signals, such that more extensive electoral manipulation indicates greater regime strength. This research trend emphasizes the degree to which political principals require the cooperation of large numbers of front-line agents in order to generate extensive election fraud, agents who may shirk if their perceived interests are not aligned with the incumbent [@rundlett2016deliver].

Electoral manipulation has been shown to be more extensive under conditions that minimize that minimize principal-agent problems (i.e. those that make low-level agents more likely to cooperate with the incumbent). These include: greater local popularity for the incumbent [@rundlett2016deliver], greater incumbent control over patronage resources [@Harvey2019;@hale_patronal_2014], and lower risk of punishment for pro-regime lawbreaking [@Harvey2022]. As discussed below, these capacities and resources are also independently useful for surviving post-election protest. Consequently, if an incumbent is able to manufacture a large numbers of votes, it is also likely to have the resource capacity, control over bureaucratic institutions, and perhaps deep wells of popularity in parts of the country that can serve as independent sources of regime durability, even if the incumbent's nationwide margin of victory is low. Lastly, extensive manipulation also indicates that large number of agents assess that the incumbent will remain in power in the future [@simpser2013], which itself becomes a self-reinforcing equilibrium [@hale_patronal_2014].

These resources and capacities improve the incumbent's ability to respond to a nascent protest movement through repression or concessions.^[Toleration, another possibility, is discussed later.] [@tillyMobilizationRevolution1978;@lipskyProtest1968;@Hummel2019;@raslerConcessionsRepression1996;@Bishara2015;@Frantz2014]. Concessions may be aimed at protesting parties themselves [@reuter2015;@mooreRepressionDissentSubstitution2000;@Hafner-Burton2016;@Piven1979] or the more moderate public, in order to keep them on the sidelines of the conflict [@Hummel2019]. In either case, large-scale election manipulation indicates control over considerable patronage resources [@Harvey2019;@hale_patronal_2014;@greene_why_2007], which in turn indicates the capacity to engage in such payments (or withhold them from clients in exchange for passive support [@Rosenfeld2017]).   

Ruling parties may also lean on repression, calling upon the security services to disperse protests, arrest leaders, and perhaps employ harsher methods including disappearance, torture, and execution [@Escriba_Folch2013;@Davenport2007]. Though such crackdowns risk provoking a backlash, they are aimed at demobilizing the opposition by increasing the costs of participation in overt dissent [@DeVogel2022]. The ability to protect regime agents from any legal consequences for carrying out orders, a condition which helps enable illegal election manipulation [@Harvey2022], is also an asset when asking law enforcement agents to carry out possibly violent repression [@MAGALONI_RODRIGUEZ_2020;@stavro_police_2023;@abouharbFactoJudicialIndependence2013]. Additionally, more consolidated control over patronage resources can help incentivize elite unity [@Reuter2019;@hale_patronal_2014;@Arriola_2009], which can enable repression by the security forces [@hendrix_2017;@Frugé_2019;@Albrecht_Ohl_2016].

In addition, as Gehlbach and Simpser [-@gehlbach2015electoral] argue, effective electoral manipulation can induce regime agents to exert effort on behalf of the incumbent. Elite defections, which can presage real challenges for the incumbent during a crisis, are also rarer in less democratic regions, where election manipulation is likely more severe [@Reuter2019]. Just as election manipulation has been held to solve a coordination problem for regime opponents, it may also solve a coordination problem for regime agents: when successful, a campaign to manufacture votes signals to agents that other agents are likely to stick with the incumbent due to its revealed capacities [@hale_patronal_2014]. 


Putting it all together: this theory assumes that authoritarian incumbents generally prefer order to unrest, and that protesters\textemdash who under either a narrow or broadly strategic view are aiming for some kind of concession from the incumbent\textemdash want to demonstrate the popularity of their cause through large, sustained protests. Electoral protests are a form of competition between the incumbent and opposition actors, in which both sides are trying to demonstrate their strength in order to win concessions from\textemdash or defeat outright\textemdash the other [@Beaulieu2014a]. The outcome of this contest reveals the underlying strengths of the participants. Consequently, governments that quickly defeat protest movements can be considered stronger than those that take weeks or months to prevail, which in turn can be considered stronger than those governments which collapse in the face of mass protest.  Incumbents with the resources and capacities needed to produce widespread election manipulation can draw on those same tools to buy off, repress, or otherwise quash post-election protest. This yields Hypothesis 2.

> H1: Election manipulation is negatively associated with protest movement survival, with the size of this effect increasing with incumbent vote-share.

Second, this analysis will also confirm that electoral manipulation is associated with increasing likelihood of protest initiation, especially when the incumbent vote-share is low. The proposed mechanism, however, is different from that articled in regime-weakness models. In those models, a high fraud / low incumbent vote-share is the ideal signal of regime weakness. It alerts the opposition that the incumbent's true popularity is unusually low (since fraud is cheap talk), and thus that the incumbent is unlikely to survive a protest movement. Protest initiation is thus strategically rational in a a narrow sense, in regime-weakness models of manipulation. 

This argument holds that widespread public grievances lead to low natural popularity for the incumbent, prompting the use of electoral manipulation to offset this disadvantage. Protests are more likely to occur both because of these underlying grievances [@Beaulieu2014a; @Brancati2016] and because a manipulated—or "stolen"—election serves as a coordination signal for collective action [@tucker2007]. As an incumbent’s genuine vote share rises, holding the level of manipulation constant, protest becomes less likely due to diminished public discontent. Conversely, for a constant level of popularity, an increase in manipulation raises the perceived credibility of election theft, thereby making protest more likely among disaffected citizens.

Here, the argument is that widespread grievances result in low natural popularity for the incumbent; at election time, the incumbent must compensate for this low popularity by manipulating the election. Protest initiation is made more likely both by the population's grievances alone [@Beaulieu2014a;@Brancati2016], and by the coordination device provided by the 'stolen' election [@tucker2007]. As natural vote-share increases for a fixed level of manipulation, the likelihood of protest initiation should decline due to reduced grievances; for a fixed level of popularity, increasing manipulation should increase the likelihood of protest by raising the plausibility of a stolen election among dissatisfied citizens.

> H2: Increasing election fraud makes protes initiationt more likely, with the size of this effect decreasing with incumbent-party vote-share. 

The core of the argument is as follows: incumbents generally aim to manipulate elections as extensively as possible to reap both the direct and indirect benefits of doing so [@simpser2013]. Achieving this requires overcoming principal-agent problems---by tightening control over patronage networks and insulating agents from potential consequences. The very resources that enable large-scale manipulation—such as genuine popularity, control over state resources, and dominance of the legal system---also equip incumbents to suppress or dismantle protest movements when they arise. Consequently, more extensive manipulation should correlate with shorter durations of anti-incumbent protests.

One potential challenge to this theory is the possibility that incumbents might choose to tolerate protests rather than suppress or co-opt them. However, this concern is relatively limited for the cases considered here. Toleration is most likely when protests pose minimal costs to the regime and make narrow or localized demands [@kleinDynamicsPoliticalProtests2018]. National elections, by contrast, are highly salient, and contesting their legitimacy poses a direct threat to incumbents' survival in office. In such contexts, tolerance is unlikely. What may appear as toleration is more accurately described as attrition---efforts by the incumbent to raise the costs of participation and gradually wear down protest movements without resorting to overt repression [@yuenNeitherRepressionConcession2017]. Thus, even without direct crackdowns or concessions, we should expect incumbents to take measures to undermine and disperse post-election protests. 

Testing both H1 and H2 will shed light on the validity of the regime-weakness model of electoral manipulation. If H2 is supported but H1 is not, this would suggest that manipulation reflects incumbent vulnerability, with protest movements exploiting this weakness. Alternatively, if both hypotheses are supported, it would point toward a regime-strength interpretation, in which manipulation reflects incumbent capacity and protest serves broader purposes.


This argument\textemdash that successful, large-scale election manipulation draws on the same resources a state may use to suppress protest\textemdash turns the deterrent logic on its head. Instead of signalling weakness, large-scale fraud indicates a well of resilience: control over resources, impunity for pro-regime lawbreaking, and loyal agents. As a result, governments that build and deploy an electoral manipulation apparatus are simultaneously inoculating themselves against the expected cost of protest.

<!-- CJH note: Do I need to write more about why protesters stay in the streets? If protest is not instrumental, they might go back to their homes for non-instrumental reasons. I.e. they've had their say once or twice, and have vented their anger. Consider also Slantchev 2004 as an analogy (wars::protest movements -- both involve commitment problems and uncertainty; last longer when both sides are equally matched.)   -->


## Data and methods



These hypotheses are tested using data from three datasets: NELDA [@hyde_2012_marinov], V-Dem version 15  [@VDemv15], and the Electoral Contention and Violence dataset [@Daxecker2019]. The unit of observation for this study is an election period as recorded in NELDA. This framework prevents selecting on the dependent variable, by capturing both protested and non-protested elections. I include in the sample authoritarian regimes which hold multiparty national elections (as well as, in some analyses) electoral democracies. This excludes fully closed autocracies (where no national elections are held de jure), as well as liberal democracies. The specific variable used is 'v2x_regime'; I limit the sample in the primary analysis to those regime-years coded 1 (electoral authoritarian) for the election year. Secondary analyses also include country-years coded 2 (electoral democracy). This sample includes a range of electoral authoritarian cases, including those that might be labeled as 'hegemonic' and 'competitive' [@donno2013elections].^[According to the V-Dem codebook, the coding scheme includes cases where multiparty elections were manipulated to the extent that "irregularities...affected the outcome of the election," and ones in which it is "hard to determine whether the irregularities affected the outcome or not." Only when irregularities definitively did not affect the outcome would election-years be categorized as electoral democracy, and excluded from the main analysis.] For instance, the dataset includes election years from Egypt and Belarus, but also from Ukraine, and Mexico in the 1990s. Regarding electoral democracies, the dataset includes cases like Bulgaria, Brazil, or Senegal in the 1990s and 2000s, and India in the 1990s.

Although the theoretical mechanisms discussed here may also operate in liberal democracies, their influence likely diminishes as democratization progresses. As states move closer to liberal democracy, patronage networks tend to weaken, and adherence to the rule of law increases—developments that inherently make it harder for incumbents to manipulate elections and easier for challengers to mobilize electoral protest. Moreover, in more liberal-democratic contexts, citizens and opposition actors have a broader range of institutional tools beyond protest to hold executives accountable. Independent judiciaries and stronger constraints on executive power have been linked to improved electoral integrity [@Harvey2022; @Bishop2016a]. It is in electoral authoritarian regimes—where such institutional checks are weakest—that the deterrence logic associated with the regime-weakness model is most applicable.

Additionally, from a methodological standpoint, protest duration becomes a less reliable indicator of regime strength in more liberal democratic environments, where protests are more routine and pose less of an immediate threat to incumbents. Therefore, the decision to exclude liberal democracies from the analysis aligns both with the theoretical focus and with the goal of conducting a more rigorous test of the regime-weakness model. The case selection strategy here, excluding liberal democracies, is thus one that both fits the theory and makes for a stronger test of the regime-weakness model. In total, there are `r nrow(nelda2.sub)` elections across `r length(unique(nelda2.sub$country))`. The date range is limited by the availability of data on protest from ECAV, which ranges from 1989 to 2012.

For models of protest-wave duration, as discussed in more detail below, I include only observations where a positive number of protests occurred. Only `r nrow(protest.cases)` fall into this more limited dataset. In eleven of these cases, according to NELDA, protesters succeeded in ousting the government for forcing new elections.^[These cases are: Bangladesh (1996), Peru (2000), Serbia (2000), Georgia (2003), Ukraine (2004), Azerbaijan (2005), Kyrgyzstan (2005), Haiti (2006), and Cote d'Ivoire (2000 and 2010).]


### Dependent variables and modeling strategies



ECAV records protests and other events related to an election, ranging from six months before to three months after the vote (or three months after the final round of a multi-round election). To qualify, an action must be contentious; rallies in support of the incumbent are excluded. I only include events labeled as protests, occupations, or blockades, excluding acts like shootings, bombings, and coups. If any post-election protests are recorded in the ECAV data for a particular election, the binary variable *protest initiation* is marked as 1. ECAV data is in event-day format; to capture *protest duration* I take the number of days between the last protest recorded in ECAV and the first, for each election.^[The duration of a protest wave with only one protest is taken to be one day.] For the survival models, protest movements that were ultimately defeated by the incumbent are considered 'dead' after the last day of protest. Protest movements that prevailed in unseating the incumbent are included, and are marked as 'alive.' 


The large majority of elections experience no post-election protest: in `r nelda2.sub %>% filter(n.events.post == 0) %>% nrow()` of the `r nelda2.sub %>% nrow()` elections, no protests are recorded. Figure \@ref(fig:elex-hist) reports the distribution of election fraud across the dataset, as well as protest duration. As the figure shows, while election fraud is common among this set of regimes, protests are rare, and, when they do occur, are usually short in duration. These distributions\textemdash extensive election fraud combined with large numbers of non-protest\textemdash provides some early indication that the regime-weakness model has limited support. Still, approximately one-quarter of elections experienced some protest, and so protest initiation is modeled using standard logistic regression. Protest wave duration is modeled using Cox proportional hazards models.



```{r, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}

p.hist.events <- ggplot(nelda2.sub, aes(x=n.events.post)) + geom_histogram(bins = 90) + xlab("Number of protest events") + theme_bw()

p.hist.duration <- ggplot(nelda2.sub, aes(x=d.events.num)) + geom_histogram(bins=100) + xlab("Protest wave duration") + theme_bw()


p.hist.fraud <- ggplot(nelda2.sub, aes(x = v2elirreg.inv)) + geom_histogram(bins=100) + xlab("Election fraud") + theme_bw()


 ggplot(nelda2.sub, aes(y = v2elirreg.inv, x= jitter(n.events.post))) + geom_point(aes(color = factor(inc.fav.cancel.protests))) + xlab("n.events.post (jittered)") + theme_bw()
 
#  ggplot(nelda2.sub, aes(y = v2elirreg.inv, x= max_size)) + geom_point(aes(color = factor(inc.fav.cancel.protests))) + xlab("max size") + theme_bw()
```


```{r elex-hist, fig.cap="Histograms of election fraud and protest duration", fig.height=4, echo=FALSE, message=FALSE, warning = FALSE}
plot.combined.hist <- cowplot::plot_grid(p.hist.fraud, p.hist.duration, ncol=1)
plot.combined.hist
```



```{r, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE, eval=FALSE}
ggplot(nelda2.sub, aes(x = v2elirreg.inv, fill = as.factor(v2x_regime))) +
  geom_histogram(alpha = 0.6, position = "identity", bins = 80) +
  scale_fill_manual(values = c("1" = "#FF6B6B", 
                               "2" = "#4ECDC4")) +
  labs(title = "Overlapping Histogram of Two Groups",
       x = "Values",
       y = "Count",
       fill = "Group") +
  theme_minimal() +
  theme(legend.position = "top")
```


### Explanatory variables

To capture the severity of *election fraud*, I use the V-Dem measure of intentional election irregularities, a category which includes "double IDs, intentional lack of voting materials, ballot-stuffing, misreporting of votes, and false collation of votes." It is important to note that this measure captures illegal election-day manipulation broadly, and is not confined simply to falsification. 

I operationalize electoral manipulation using this variable, rather than alternative measures of election integrity, because fraud of this kind is thought to be most likely to trigger protest [@Linebarger2020]---it is often incumbents' tool of last resort [@Sjoberg2016], may not mobilize real voters [@Harvey2020], and can be more damaging to legitimacy than pre-election manipulation [@birch2011;@Szakonyi2021]. This measure of election integrity should thus serve as a strong test of the regime-weakness model: if manipulation indicates incumbent weakness, it will be most likely to do so when employed in this way. It also matches the conception of election manipulation used in much of the relevant formal literature.^[For example, election manipulation is treated as post-hoc adjustment of the results in Little et al [-@Little2015], Little [-@Little2012], Magaloni [-@magaloni2010], and Fearon [-@fearon_self-enforcing_2011], among others.] 



### Control variables

Since vote-share data is given in V-Dem for the winning party in an election, without recording whether that winner was the incumbent or the challenger, it is necessary to include controls for incumbent defeat in the models. From the point of view of those considering anti-incumbent protest, for instance, a large opposition vote-share conveys very different information when the opposition is declared the loser compared to when it wins outright. Consequently, in models of protest initiation, I condition for whether the incumbent party lost, based on coding in NELDA. In the binary models of Hypothesis 2, I control for incumbent loss as a standalone variable or in interaction with the main explanatory variables.


For the models of protest duration, I exclude all cases where the incumbent lost the election outright. This limits the observations to those that fit the population of interest: protest following an election in which the incumbent claimed victory, possibly through reliance on election fraud.^[Excluding cases where incumbents stepped down after losing an election avoids introducing measurement error in the dependent variable, due to a quirk of the ECAV dataset, which codes the 'side' of the protest (pro- or anti-government) as missing during the period between an election and the swearing in of a new government in cases where the incumbent loses. Including these cases thus runs the risk of conflating pro- and anti-incumbent protest, along with cases where protesters goals are genuinely unknown. However, excluding them also influences the interpretation of the results from these models. Threats to inference from this section are evaluated in more detail in the discussion section.]  

The models incorporate several potential confounding variables, in addition to the *election fraud* variable described above. The *lagged GDP growth rate*, taken from V-Dem, is used to capture economic grievances: lower GDP growth rates, especially negative ones, should be expected to increase citizens' sense of grievance. The winning party's *vote-share* is also taken from the V-Dem dataset.^[This variable has considerable missingness, which I reduced in part by updating missing values in V-Dem using data from Wikipedia (when links to official sources were available), and Psephos.]

As identified in prior research, other factors that may influence both fraud and protest are included as controls variables. The repressive capacity of the government [@Davenport2019] is measured using the *physical integrity index* from V-Dem (v2x_clphy), and its square term. Indicators of government *control over civil society* (the 'v2cseeorgs' variable from V-Dem), de facto *judicial independence* [@Linzer2015] and the available of *alternative sources of information* ('v2x_freexp_altinf' from V-Dem) are also employed. Since urban areas can make electoral manipulation more difficult [@ziblatt2009;@larreguy2016parties] while also making protest more likely [@Wallace2013a], I also include data on *urbanization* taken from the United Nations [@DepartmentofEconomicandSocialAffairsPopulationDivision2019]. All models include *legislative constraints* on the executive, a V-Dem variable which captures the extent to which the legislative body in the country acts independently in practice to limit executive power.  

In addition, the literature on the diffusion of electoral protest holds that successful electoral protest in one country increases the likelihood of electoral protest in similar countries nearby [@Bunce2010;@Beissinger_2007]. Such a dynamic could plausibly increase the likelihood of protest in a nearby country, while also reducing the incumbent's use of fraud. I control for this using each country's geopolitical region as coded in Teorell et al [-@qog_teorell], in conjunction with binary variables indicating a successful protest movement from NELDA. A 'successful protest' is recorded if a protest movement resulted in either a new election being run or the incumbent stepping down. The *regional diffusion* dummy variable takes on a value of one for all countries in the same region as a successful protest, for elections within one calendar year of the successfully overturned election.^[For example, the variable takes on a positive value for elections held in the post-communist region, within one year of the September 24, 2000 election in Serbia that sparked the 'Bulldozer Revolution'.] 


Several other binary control variables are also included. First, a variable from NELDA is used to indicate *presidential* elections, since these may include both higher levels of manipulation [@simpser2013] and a greater risk of protest. Another NELDA variable indicates if the incumbent executive is not running due to *term limits* [@hale_patronal_2014]. Since strong non-democratic governments may be able to adjust the electoral calendar strategically (to affecting both election manipulation and protest risk), I include a variable from NELDA that indicates *off-schedule elections*.   

<!-- I include dummy variables that provide a categorical evaluation of state capacity. These are taken from the ordinal version of the V-Dem variable "state source of fiscal revenue" (v2stfisccap). If the V-Dem data codes a state as not capable of raising revenue or primarily relying on loans and foreign aid, I code it as a *weak state*. If the state is described as primarily relying on resource rents and expropriation, I code it is a *rentier state*. States that rely primarily on taxes on property or economic transactions are coded as *tax states*. This distinction is relevant since access to natural resource rents may both make protest less likely and repression more effective [@Girod2018]. -->

In the models of electoral protest initiation, I include the *number of pre-election protests* counted in the ECAV dataset, which is likely to be an important possible confounder; more numerous pre-election protest may affect the level of election fraud (as the deterrence logic predicts) as well as being directly tied to the likelihood of post-election protest. Finally, in the models of protest duration, I also include a binary  measure that captures whether any protest in the wave was *large*, using ECAV data. ECAV data codes protest size as a categorical variable, ranging from 1 (< 10 participants) to 5 (> 10,000). If any single protest event was categorized as a '5' on this scale, the *any large protest* variable takes on a value of 1.  This is important as a control for the possibility that incumbents may tolerate smaller protests.

Table \@ref(tab:summary-table) provides summary statistics for these variables.

```{r, include=F}
nelda2.sub <- nelda2.sub %>% mutate(any_large = ifelse(is.na(median.participants)==T, 0, ifelse(median.participants > 4, 1, 0)))

incl.vars <- nelda2.sub %>% dplyr::select(#'Judicial independence' = jucon.1lag, 
                                            'Legislative constraints' = legcon.1lag, 
                                            'Alternative information' = altinf.1lag,
                                          #'Liberal democracy index' = libdem.1lag,
                                            'GDP growth rate' = gdpgrowth.1lag, 
                                            'Civil society openness' = osorg.1lag, 
                                            #'Public sector corruption' = pubseccorrup.lag,
                                           
                                             'Election fraud' = v2elirreg.inv, 
                                            'Presidential election' = presidential,
                                            'Incumbent term-limited' = inc.termlimit.num,
                                            'Physical integrity index' = physinteg.1lag,
                                            'Urbanization' = urban.pct, 
                                           
                                            'Winning vote-share' = winner.share.upd, 
                                            'Number of pre-election protests' = n.events.pre,
                                            'Off-schedule election' = unscheduled_election,
                                          'Regional protest diffusion' = regional_success_lag,
                                          'Large protest' = any_large
                                       )
#incl.vars <- incl.vars %>% mutate(v2elintmon = as.numeric(v2elintmon))

incl.vars <- incl.vars %>% mutate(`Presidential election` = as.numeric(`Presidential election`))
#incl.vars <- incl.vars %>% mutate(`Rentier state` = as.numeric(`Rentier state`))
#incl.vars <- incl.vars %>% mutate(`Tax state` = as.numeric(`Tax state`))
incl.vars <- incl.vars %>% mutate(`Off-schedule election` = as.numeric(`Off-schedule election`))



```




```{r summary-table, echo=FALSE, message=FALSE, warning=FALSE}
library(kableExtra)
datasummary_skim(incl.vars, output = "kableExtra", title = "Summary statistics and variable distributions", histogram = F)  %>%
    kable_styling(font_size = 9, full_width = FALSE)
```





```{r, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}

p.hist.events <- ggplot(nelda2.sub, aes(x=n.events.post)) + geom_histogram(bins = 90) + xlab("Number of protest events") + theme_bw()

p.hist.duration <- ggplot(nelda2.sub, aes(x=d.events.num)) + geom_histogram(bins=100) + xlab("Protest wave duration") + theme_bw()


p.hist.fraud <- ggplot(nelda2.sub, aes(x = v2elirreg.inv)) + geom_histogram(bins=100) + xlab("Election fraud") + theme_bw()


 ggplot(nelda2.sub, aes(y = v2elirreg.inv, x= jitter(n.events.post))) + geom_point(aes(color = factor(inc.fav.cancel.protests))) + xlab("n.events.post (jittered)") + theme_bw()
 
#  ggplot(nelda2.sub, aes(y = v2elirreg.inv, x= max_size)) + geom_point(aes(color = factor(inc.fav.cancel.protests))) + xlab("max size") + theme_bw()
```



## Results




### Fraud and the duration of protest

<!-- All models exclude "successful protests," which either overturned an election or ousted an incumbent. Including such cases (and coding them as "alive" rather than "dead" in the survival model) does not meaningfully change the results; it would, however, complicate the interpretation of the models. Excluding successful protests allows us to interpret every 'death' of a protest movement as a defeat for the protesters and a victory for the incumbent. -->

Table \@ref(tab:cox-models-tab) reports the results of Cox proportional hazard models of protest duration. In these models, positive coefficients indicate factors that are associated with shorter-lived protests; those with a negative sign indicate factors associated with greater likelihood of protest-movement survival. Models 4 and 5 do not include controls (other than *presidential election*, which is included due to the common mechanical difference in winning vote-shares between presidential and parliamentary elections), while the remaining models include a fuller set of controls. Models 1 and 3 do not account for potential right-censoring, while Models 2, 4, and 5 do. Accounting for right-censoring helps address a threat to inference in this design that derives from the 90-day post-election window considered by ECAV. It is possible that some protest movements considered 'dead' in the baseline Models 1 and 3 are, in fact, 'alive,' if subsequent protests occurred after 90 days and are unobserved. To address this, observations are assumed to be right-censored if a protest occurred within three-weeks of the 90-day cutoff in Models 2, 4, and 5. This amounts to 15 of the 156 observations in the sample. Model 5 includes right censoring as well as a control variable for the occurrence of at least one very large protest. Lastly, in order to interpret these findings, it is important to recall that these models include only cases where incumbents claimed victory and electoral protest occurred.

To aid in the interpretation of the interaction effect in Model 4 (with right-censoring), Figure \@ref(fig:margin-dur-fig) shows the predicted risk score As shown in the figure, for all incumbent vote-shares near and below fifty percent, large-scale fraud significantly increases the likelihood that a protest movement will end in defeat. This result also supports the regime-strength model, and shows that only as vote-share increases above fifty percent\textemdash creating a stronger signal of popularity, another independent source of regime strength\textemdash does a government in a low-fraud election have the same likelihood of defeating a protest movement at a given time. In effect, either high genuine popularity *or* the ability to produce high levels of fraud enables governments to more quickly end protests.

The results of the models are roughly consistent whether they incorporate censoring or not. Model 5 adds the a control for large protests, as a measure of the level of danger faced by the incumbent. Including this control variable does not meaningfully change the results; the interaction coefficient retains its direction and statistical significance. 



```{r, include=FALSE}
library(survival)
library(ggsurvfit)



## With non protests = insta-death (status 1 at time zero)
## And protest success == survival
### This doesn't work conceptually, since a happy populace (no protest) will be counted as a defeated protest movement
#nelda2.sub <- nelda2.sub %>% mutate(surv_status = ifelse(is.na(d.events.num) == F, 1, NA))
#nelda2.sub <- nelda2.sub %>% mutate(surv_status = ifelse(inc.replaced.prot == 1 | new.elections.held == 1, 0, surv_status))

#nelda2.sub <- nelda2.sub %>% mutate(surv_status_censor_assume = ifelse(is.na(d.events.num) == T, NA,  
#                                                                       ifelse(d.events.num >= 69, 0, 1)))

## With non-protests excluded
## And protest success == survival
nelda2.sub <- nelda2.sub %>% mutate(surv_status = ifelse(d.events.num > 0, 1, NA))
nelda2.sub <- nelda2.sub %>% mutate(surv_status = ifelse(inc.replaced.prot == 1 | new.elections.held == 1, 0, surv_status))
nelda2.sub <- nelda2.sub %>% mutate(surv_status_censor_assume = ifelse(d.events.num == 0, NA, 
                                                                       ifelse(d.events.num >= 69, 0, 1)))










cox1 <- coxph(Surv(d.events.num, surv_status) ~ v2elirreg.inv*winner.share.upd  +  presidential  + v2elirreg.inv , data = nelda2.sub %>% filter(inc.lose == 0  & v2x_regime == 1))
summary(cox1)



cox1.cen <- coxph(Surv(d.events.num, surv_status_censor_assume) ~  v2elirreg.inv*winner.share.upd  +  presidential, data = nelda2.sub %>% filter(inc.lose == 0  & v2x_regime == 1))
summary(cox1.cen)





#cox1t2 <- coxph(Surv(d.events.num, surv_status) ~ legcon.1lag + altinf.1lag  +  osorg.1lag + gdpgrowth.1lag + natlpartyorg.lag + winner.share.upd + presidential + inc.termlimit.num  +  physinteg.1lag + physintlag.sq + urban.pct + rentierstate + taxstate + unscheduled_election +  regional_success_lag + n.events.post + v2elirreg.inv + tt(n.events.post), data = nelda2.sub %>% filter(inc.lose == "0") %>% filter(d.events.num > 0 & inc.replaced.prot == 0), tt = function(x, t, ...)  x* nsk(t, knots=c(10, 70, 90),
#Boundary.knots = FALSE))
#summary(cox1t2)









cox2 <- coxph(Surv(d.events.num, surv_status) ~  legcon.1lag + gdpgrowth.1lag +  + v2elirreg.inv*winner.share.upd  + presidential + inc.termlimit.num  +  physinteg.1lag  + urban.pct  + unscheduled_election + regional_success_lag + n.events.pre , data = nelda2.sub %>% filter(inc.lose == "0"  & v2x_regime == 1))
summary(cox2)
p.duration.margin <- sjPlot::plot_model(cox2, type = "pred", terms = c("winner.share.upd", "v2elirreg.inv [-.53, 2.45]")) #) #This is the 5th and 95th quantiles
p.duration.margin <- p.duration.margin +  labs(x = "Incumbent vote-share", y = "Hazard ratio", title = "Hazard ratio for protest movement end", color = "Election fraud") + theme_bw() + coord_cartesian(ylim = c(0, 8))


cox2.cen <- coxph(Surv(d.events.num, surv_status_censor_assume) ~  legcon.1lag + gdpgrowth.1lag +  + v2elirreg.inv*winner.share.upd  + presidential + inc.termlimit.num  +  physinteg.1lag  + urban.pct  + unscheduled_election + regional_success_lag + n.events.pre, data = nelda2.sub %>% filter(inc.lose == "0"  & v2x_regime == 1))
summary(cox2.cen)

p.duration.margin.cen <- sjPlot::plot_model(cox2.cen, type = "pred", terms = c("winner.share.upd", "v2elirreg.inv [-.23, 2.58]"))
p.duration.margin.cen <- p.duration.margin.cen +  labs(x = "Incumbent vote-share", y = "Hazard ratio", title = "Hazard ratio for protest movement end", color = "Election fraud") + theme_bw() + coord_cartesian(ylim = c(0, 10))




cox3.cen <-  coxph(Surv(d.events.num, surv_status_censor_assume) ~  legcon.1lag +  gdpgrowth.1lag +  + v2elirreg.inv*winner.share.upd  + presidential + inc.termlimit.num  +  physinteg.1lag  + urban.pct  + unscheduled_election +  any_large, data = nelda2.sub %>% filter(inc.lose == "0" & v2x_regime == 1))
summary(cox3.cen)
p.duration.margin.cen3 <- sjPlot::plot_model(cox3.cen, type = "pred", terms = c("winner.share.upd", "v2elirreg.inv [-.53, 2.45]")) + coord_cartesian(ylim = c(0, 10))


```



```{r km-fig-auth, fig.cap="Kaplan-Meier curves for protest duration across levels of fraud", fig.height= 5, echo=F, message=F, fig.pos='H'}
nelda2.sub.auth <- nelda2.sub %>% filter(v2x_regime == 1) %>% mutate(fraud_level = ifelse(v2elirreg.inv >= (1.09 + .97), "High",
                                                            ifelse(v2elirreg.inv <= (1.09 - .97), "Low", "Moderate")))
nelda2.sub.auth.km <- nelda2.sub.auth %>% filter(inc.lose == "0") %>% filter(d.events.num > 0 & inc.replaced.prot == 0)


p.km.auth <- survfit2(Surv(d.events.num, surv_status) ~ fraud_level, data = nelda2.sub.auth.km %>% filter(d.events.num < 300)) %>% 
  ggsurvfit(linetype_aes = T) +
#   add_confidence_interval() +
  labs(
    x = "Days",
    y = "Overall survival probability"
  )

p.km.auth
```




The relationship between the level of fraud and the survival of protest movements can also be demonstrated non-parametrically using Kaplan-Meier curves for varying levels of fraud. In Figure \@ref(fig:km-fig-auth), observations are divided into three groups. High fraud observations are those where *election fraud* is one standard deviation above the median or more, low fraud observations are those one standard deviation below the median or less, with moderate levels in between. The figure shows a clear difference between the low-fraud group and the other two. Low-fraud cases are much more likely to see their protest movements survive past the first day than their counterparts. For much of the range of the data, they remain roughly twice as likely to survive at any given date than protest movements in medium- or high-fraud cases. This is further evidence in that high levels of fraud is an indication of regime resilience, as it is the low-fraud cases where governments struggle to put an end to protest movements on average. 






```{r cox-models-tab, echo=FALSE, warning=FALSE}
models3 <- list()
models3[['Model 1']] <- cox1
models3[['Model 2']] <- cox1.cen
models3[['Model 3']] <- cox2
models3[['Model 4']] <- cox2.cen
models3[['Model 5']] <- cox3.cen





modelsummary::modelsummary(models3, gof_omit = "AIC|BIC|Log", longtable=F, stars = T, title = 'Table 4: Cox proportional hazard models of protest duration', coef_map = 
               c("(Intercept)" = "(Intercept)", "presidential1" = "Presidential election", "lji.lag" = "Judical independence (lag)", "legcon.1lag" = "Leg. constraints (lag)", 
                 "gdpgrowth.1lag" = "GDP growth rate (lag)", "altinf.1lag" = "Alternative info. (lag)", 
                 "osorg.1lag" = "Civil soc. openness (lag)", 
                 "natlpartyorg.lag" = "National party organization (lag)", 
                 "inc.termlimit.num" = "Incumbent term-limited", "libdem.1lag" = "Liberal democracy index (lag)", 
                 "urban.pct" = "Urbanization", "rentierstate" = "Rentier state", "taxstate" = "Tax state",
                 "n.events.post" = "N. of post-election protests", "median.participants.mm.pre" = "Median size of pre-election protests", "unscheduled_election" = "Off-schedule election",
                 "rentierstate1" = "Rentier state", "taxstate1" = "Tax state", "n.events.pre" = "Number of pre-election protests", "unscheduled_election1" = "Off-schedule election", "physinteg.1lag" = "Phys. integrity index (lag)",
                 "physintlag.sq" = "Phys. int. lag squared", "regional_success_lag" = "Regional protest diffusion", 
                 "median.participants" = "Median protest size",
                 "any_large" = "Large protest (> 10,000)",
                                  "winner.share.upd" = "Incumbent vote-share",
                 "v2elirreg.inv" = "Election fraud",
"v2elirreg.inv:winner.share.upd" = "Election fraud:Inc. vote-share"
                 ))
```




```{r margin-dur-fig, fig.cap="Fraud, winning vote-share, and duration of protest", fig.height=5, echo=FALSE, message=FALSE, fig.pos='H'}
p.duration.margin.cen
```





### Fraud and the likelihood of protest



```{r, echo=FALSE, message=FALSE, warning=FALSE}
nelda2.sub <- nelda2.sub %>% mutate(any_protest = ifelse(d.events.num > 0, 1, 0))
#nelda2.sub <- nelda2.sub %>% mutate(no_size = ifelse(max_size == 0, 1, 0))


model.binary_noint <- glm(any_protest ~ inc.lose + v2elirreg.inv + winner.share.upd, data = nelda2.sub %>% filter(v2x_regime == "1"), family = binomial(link = "logit"))

model.binary.base <-  glm(any_protest ~ winner.share.upd*v2elirreg.inv*inc.lose + presidential, data = nelda2.sub %>% filter(v2x_regime == "1"), family = binomial(link = "logit"))

model.binary.base.auth <-  glm(any_protest ~ winner.share.upd*v2elirreg.inv*inc.lose + presidential, data = nelda2.sub %>% filter(v2x_regime == 1), family = binomial(link = "logit"))

model.binary.full <- glm(any_protest ~ lji.lag + legcon.1lag + gdpgrowth.1lag + altinf.1lag  +  + osorg.1lag   +    winner.share.upd*v2elirreg.inv*inc.lose   + presidential + inc.termlimit.num  +  physinteg.1lag + physintlag.sq + urban.pct + n.events.pre + rentierstate + taxstate + unscheduled_election + regional_success_lag, family = binomial(link = "logit"), data = nelda2.sub %>% filter(v2x_regime == "1"))



plot_binary_margin_typ_n <- sjPlot::plot_model(model.binary.full, type = "pred", terms = c("winner.share.upd", "v2elirreg.inv [.16, 2.09]", "inc.lose"), show.data = T)
plot_binary_margin_typ_n <- plot_binary_margin_typ_n +  labs(x = "Winner's vote-share", y = "Predicted probability", title = "Predicted probability of protest", color = "Election fraud") + theme_bw()
```



```{r logit-table, echo=FALSE}
models2 <- list()
models2[['Model 6']] <- model.binary_noint
models2[['Model 7']] <- model.binary.base
models2[['Model 8']] <- model.binary.full
#models2[['Model 6']] <- model.margin.binary_noc

reg_table <- modelsummary(models2, output = "gt",  stars = T, title = 'Table 3: Logit models of protest occurrence', gof_omit = c("Log|RMSE"), coef_map = c("(Intercept)" = "(Intercept)",
             "inc.lose1" = "Incumbent loss",
             "v2elirreg.inv" = "Election fraud",
             "winner.share.upd" = "Winner vote-share",
"winner.share.upd:v2elirreg.inv" = "Winner vote-share:Election fraud",
"winner.share.upd:inc.lose1" = "Winner vote-share:Inc. loss",
"v2elirreg.inv:inc.lose1" = "Election fraud:Inc. loss",
"winner.share.upd:v2elirreg.inv:inc.lose1" = "Election fraud:Winner vote-share:Inc. loss",
  "lji.lag" = "Judical independence (lag)",                                                                 "legcon.1lag" = "Leg. constraints (lag)",                                                                 "gdpgrowth.1lag" = "GDP growth rate (lag)",                                                                   "altinf.1lag" = "Alternative info. (lag)", 
                 "osorg.1lag" = "Civil soc. openness (lag)",
                 "inc.termlimit.num" = "Incumbent term-limited", 
                 "libdem.1lag" = "Liberal democracy index (lag)", 
                 "urban.pct" = "Urbanization", 
                 "rentierstate1" = "Rentier state", 
                 "taxstate1" = "Tax state", 
                "presidential1" = "Presidential",
                 "unscheduled_election1" = "Off-schedule election",
                 "n.events.pre" = "Number of pre-election protests", 
                 "physinteg.1lag" = "Phys. integrity index (lag)",
                 "physintlag.sq" = "Phys. int. lag squared", 
                 "regional_success_lag" = "Regional protest diffusion",

                 "median.participants" = "Median protest participants",
 
                "v2elirreg.inv:winner.share.upd"= "Election fraud:Winner vote-share")
                 ) 

reg_table |> tab_options(
  latex.use_longtable = TRUE, 
  data_row.padding = px(5)
  ) 
```






Table \@ref(tab:logit-table) shows the results for Hypothesis 2. Model 6 is a base model, where incumbent loss, winner vote-share, and *election fraud* are standalone variables. In keeping with the general (though inconsistent) findings in the prior literature, there is a significant positive relationship between fraud severity and the likelihood of protest initiation. The winner's claimed margin of victory is negatively associated with protest initiation. Model 7 includes the three-way interaction between *election fraud*, incumbent loss, and winning vote-share, with no controls other than a dummy variable for *presidential elections*. This allows for the largest possible sample size, and shows that the coefficient signs are consistent with those of the full Model 8. 


Model 8 is the main model, which includes the full set of controls. Given the complexity of the model, Figure \@ref(fig:margin-prob-fig) illustrates the effects of election fraud on protest risk across various levels of winning vote-share and incumbent loss^[All other variables constant at their means (continuous variables) or modes (categorical variables).]. The values for election fraud represent one standard deviation above and below the mean value. 

The direction of the effect supports Hypothesis 2. When an incumbent claims victory (left panel) and their vote-share is low, electoral protest is significantly likelier when fraud is elevated than when the election is relatively clean. Furthermore, protest against winning incumbents is significantly likelier when fraud is high and vote-shares are low, compared to an election with an equivalent level of fraud but high claimed vote-shares. As noted above, such a finding could be consistent with either the regime-weakness or regime-strength models of manipulation, under different conceptualizations of protest. To distinguish between the two, we turn to the models of protest duration.



```{r margin-prob-fig, fig.cap="Fraud, winning vote-share, and probability of protest initiation", fig.height=5, echo=FALSE, message=FALSE, fig.pos='H'}
plot_binary_margin_typ_n
```





## Additional analysis: structural and protest-based correlates of fraud
\sectionmark{Protest risk and fraud}

The final section of this chapter tests the implication that incumbent should respond to greater protest risk by *increasing* the level of electoral manipulation (conditional on their structural ability to do so). Testing for this possibility simultaneously helps confirm the theory that strong incumbents are not fearful of mass protest, and helps rule out selection effects that might undermine the general findings above.^[The classic regime-weakness models predict that some incumbents who will voluntarily avoid electoral manipulation and accept a possible loss at the ballot box, to avoid costly protest. If this is the case, it could introduce selection bias into the sample for the duration analyses, by sorting only the strongest incumbents into the sample that faces protest.] Here, I use a similar approach to take two different cuts at the data. First, I model the likelihood of post-election protest in each case, using only pre-election variables. I then use this propensity score for protest to model the level of electoral manipulation that in fact occurs. In the second approach, I model the predicted average level of manipulation in each case, and use the residuals from that model as a measure of over- or under-production of manipulation relative to baseline. I then model those residuals as a function of pre-election predictors of protest. In both approaches, the outcome is the same: a greater risk of protest is associated with a *higher* production of manipulation, not lower.

Together, these two approaches provide mutually reinforcing evidence. The propensity score approach establishes that anticipated protest is a positive predictor of manipulation; the residuals approach confirms that this relationship holds even after accounting for the structural conditions that make manipulation possible in the first place. Both findings are consistent with the theory that strong incumbents are not deterred by the prospect of mass protest — indeed, they treat it as a reason to consolidate their advantage further. This directly complements the main analysis: if incumbents feared protest, they would pull back on manipulation to avoid provoking it; the fact that they do the opposite suggests they are confident in their ability to weather any backlash.


### Predicted and residual levels of fraud

First, let us consider how indicators of protest risk shape deviations from the expected level of electoral manipulation in each case. To do so, I use all the observations in the dataset to estimate the predicted level of election fraud for each case, using only structural variables (i.e. professionalization of the bureaucracy, electoral management body autonomy, urbanization, etc.) in an OLS model. Factors associated with classic protest deterrence, like the number of pre-election protests, are excluded. This should produce an estimate of the structural level of manipulation the incumbent could produce in the absence of strategic deterrence.

The residuals from this first model are then used as the dependent variable in a second model. These residuals capture the extent to which an incumbent manipulated more or less than their baseline level would predict — that is, whether they over- or under-produced manipulation. If the deterrence interpretation is correct, indicators of protest risk should be negatively associated with these residuals: incumbents facing greater protest risk should pull back on manipulation, producing less than expected. The regime-strength logic predicts the opposite — that incumbents facing riskier environments will work harder to signal strength, and so will manipulate more than their baseline predicts.

Table \@ref(tab:fraud-loss) shows the results. Model 5 shows how structural factors are predictors of the level of observed manipulation. We observe some expected results; for instance, more autonomous EMBs and more professionalized bureaucracies are associated with lower fraud. Model 6 shows that indicators of protest risk\textemdash the number of pre-election protests, and the existence of a recent successful protest movement in the region\textemdash are not associated with lower fraud. In fact, just the opposite: incumbents appear to manipulate more heavily under riskier conditions. This is a key finding, and is evidence against the classic deterrent approach; it also suggests that weaker governments are not voluntarily dialing back manipulation to avoid protest, mitigating concerns about selection effects in the main models. 


```{r, include=F, echo=FALSE, message=FALSE, warning=F, eval=FALSE}
## Anova
anova_dat <- nelda2.sub %>% dplyr::select(inc.replaced.prot, v2elirreg.inv,  jucon.1lag, legcon.1lag, gdpgrowth.1lag,  osorg.1lag, natlpartyorg.lag,  urban.pct, n.events.pre, state.owned.1lag, patronage.appts.1lag, v2elembaut)

# boxplots and t-tests for the 4 variables at once
for (i in 2:ncol(anova_dat)) { # variables to compare are variables 2 to end
  boxplot(anova_dat[, i] ~ anova_dat$inc.replaced.prot, # draw boxplots by group
    ylab = names(anova_dat[i]), # rename y-axis with variable's name
    xlab = "Incumbent replaced after protest"
  )
  print(paste(c("ANOVA for variable ", (names(anova_dat[i])))))
  print(t.test(anova_dat[, i] ~ anova_dat$inc.replaced.prot)) # print results of t-test
}

aov_output <- aov(state.owned.1lag ~ inc.replaced.prot, data = nelda2.sub)
summary(aov_output)
```


```{r, echo=FALSE, message=FALSE, include=FALSE, warning=FALSE}
m.fraud.pred_base <- lm(v2elirreg.inv ~ n.events.pre + regional_success_lag, data = nelda2.sub)

m.fraud.pred <- lm(v2elirreg.inv ~ jucon.1lag +  gdpgrowth.1lag + presidential   +   urban.pct + rentierstate + taxstate + physinteg.1lag + physintlag.sq + state.owned.1lag + patronage.appts.1lag  + v2elembaut, data = nelda2.sub)

#summary(m.fraud.pred)

nelda2.sub.fraudpred <- nelda2.sub %>% mutate(.rownames = rownames(nelda2.sub))

new_frame <- augment(m.fraud.pred)
new_frame <- new_frame %>% dplyr::select(.fitted, .rownames, .resid)

nelda2.sub.fraudpred <- nelda2.sub.fraudpred %>% left_join(new_frame, by = ".rownames")

model.lose_reg <- glm(inc.lose ~ .fitted + regional_success_lag , data = nelda2.sub.fraudpred, family = binomial(link="logit"))
#summary(model.lose_reg)

model.lose_events <- glm(inc.lose ~ .fitted + n.events.pre , data = nelda2.sub.fraudpred, family = binomial(link="logit"))
#summary(model.lose_events)

model.fraud.resid <- lm(.resid ~ n.events.pre +  regional_success_lag , data = nelda2.sub.fraudpred)
#summary(model.fraud.resid)

model.lose_resid <- glm(inc.lose ~  .resid  , data = nelda2.sub.fraudpred, family = binomial(link="logit"))
#summary(model.lose_resid)

model.protest_resid <- glm(any_protest ~  .resid  , data = nelda2.sub.fraudpred, family = binomial(link="logit"))
#summary(model.protest_resid)
```

```{r fraud-loss, echo=FALSE}
models2 <- list()
models2[['Model 5']] <- m.fraud.pred
models2[['Model 6']] <- model.fraud.resid


row <- data.frame(cbind("", "Fraud", "Fraud residual"))
attr(row, 'position') <- c(0)

modelsummary(models2, output = "kableExtra",  stars = T, title = 'Models of manipulation and residuals (OLS) and incumbent loss (logit)', coef_omit = "as.factor", gof_omit = "RMSE|Log.", coef_rename = 
               c("jucon.1lag" = "Judical independence (lag)", "legcon.1lag" = "Leg. constraints (lag)", 
                 "gdpgrowth.1lag" = "GDP growth rate (lag)", "altinf.1lag" = "Alternative info. (lag)", 
                 "osorg.1lag" = "Civil soc. openness (lag)", "econ.crisis.nelda.1lag1" = "Economic crisis", 
                 "natlpartyorg.lag" = "National party organization (lag)", "presidential1" = "Presidential election",
                 "inc.termlimit.num" = "Incumbent term-limited", "libdem.1lag" = "Liberal democracy index (lag)", "v2elirreg.inv" = "Election fraud",
                 "winner.share.upd" = "Incumbent vote-share",
                 "urban.pct" = "Urbanization", "rentierstate" = "Rentier state", "taxstate" = "Tax state",
                 "n.events.massmob.pre" = "Number of pre-election protests", "median.participants.mm.pre" = "Median size of pre-election protests", "unscheduled_election" = "Off-schedule election",
                 "rentierstate1" = "Rentier state", "taxstate1" = "Tax state", "n.events.pre" = "Number of pre-election protests", "unscheduled_election1" = "Off-schedule election", "physinteg.1lag" = "Phys. integrity index (lag)",
                 "physintlag.sq" = "Phys. int. lag squared", "regional_success_lag" = "Regional protest diffusion", "antisys.1lag" = "Anti-system CS activity",
                 "state.owned.1lag" = "Private ownership (lag)",
                 "patronage.appts.1lag" = "Professional bureaucracy (lag)",
                 "v2elembaut" = "EMB autonomy",
                 ".fitted" = "Expected fraud level"
                 ), add_rows = row) %>% kable_styling(latex_options = "HOLD_position")

## Figure out how to Add a row beneath the model numbers, with the DV
```




### Protest risk and fraud level

Examining the data from the other direction produces similar results. In this approach, a logit model of any post-election protest is fit, using a broad set of covariates. The fitted values (i.e. predicted probabilities of protest) for each observation are obtained as propensity scores for protest, to be used in the second set of models. The goal of this step is to estimate the underlying likelihood of post-election protest in each case, independent of the level of election fraud observed on election day. This propensity score can then be used as an explanatory variable in subsequent models of *electoral fraud*, to see if a greater risk of protest is associated with reduced manipulation (as the regime-weakness model would predict) or greater manipulation (as the regime-strength model might hold). 

Since this approach takes the overall level of fraud as a dependent variable (rather than deviation from baseline), it is also appropriate to include electoral democracies in this analysis. As countries democratize, we might expect that protest risk increases, through greater freedom to organize, fewer media restrictions, and greater respect for individual political liberties (this is the logic behind the idea of 'self-enforcing' democracy). Alternatively, we might expect that as countries democratize, incumbents on average become more constrained and less capable of wielding state power to influence election outcomes (through reduced control over patronage resources and greater accountability for law-breaking). Both of these factors should limit incumbents' freedom to 'dial manipulation up' for particular elections in electoral democracies. 

It is important to note that I am not implementing a propensity-score *matching* or weighting approach here. In this framework, actual protest cannot be considered a cause of electoral manipulation which occurs *before* the protest, and thus cannot be operationalized as a treatment variable. Instead, I aim to estimate the latent protest risk in each case, and analyze that risk as a predictor of observed electoral manipulation. 

The effectiveness of this propensity score approach hinges on the success of the first-stage model, which should include variables that influence the likelihood of treatment. The first-stage model here includes lagged *judicial independence* [@Linzer2015], which can influence parties' decision to challenge elections in the streets vs. the courtroom. It also includes lagged measures of civil society organization and legislative constraints on the executive, as estimates of the degree of contestation in the political system. The lagged openness of the media environment is included, since this could influence citizens' ability to gather information about the quality of elections and to communicate their grievances. De facto protection of individual physical integrity is included (lagged), to account for the linear and curvilinear effects of repression on protest. The urban population share is included, since population density is a common correlate of protest. Lagged GDP growth is included as a measure of economic grievances. Three binary variables capture relevant features of the election. *Presidential* elections may have higher stakes, and thus be more prone to sparking protest. *Off-schedule elections* may be strategically timed by incumbents to account for more elevated protest risk. In addition, whether the executive is facing *term limits* is also included, since this can lead to fragmentation in the elite that can in turn yield protest [@hale_patronal_2014,@robertson_protest_2010]. These variables are all taken from V-Dem. The occurrence of a successful regional electoral protest in the previous year is included, given its potential to inspire protests in neighboring countries. Lastly, two final---and very important---covariates are included. These are: the number of pre-election protests (a very direct measure of protest risk), and lagged overall election integrity (i.e. overall manipulation in the previous national election). This model appears to capture the latent risk of protest reasonably well; the average propensity score for country-years where no protest ultimately occurred is 0.12. By contrast, the average propensity score for country-years where protests *did* occur is 0.66.

The results of the first stage model are presented in Table \@ref(tab:stage1-model). 


```{r, include = F, echo=FALSE}
### Calculate propensity scores for protest, but use lagged electoral fairness as variable

complete_cases <- complete.cases(nelda2.sub[, c("any_protest", "lji.lag" , "legcon.1lag" , "gdpgrowth.1lag" , "altinf.1lag"  , "frefair.1lag"  , "osorg.1lag"   ,   "presidential" , "inc.termlimit.num"  ,  "physinteg.1lag" , "physintlag.sq" , "urban.pct" , "n.events.pre" , "unscheduled_election" , "regional_success_lag")])

model.protest.null <- glm(any_protest ~ 1, family = binomial(link = "logit"), data = nelda2.sub[complete_cases, ])
summary(model.protest.null)

model.protest.nofraud <- glm(any_protest ~ lji.lag + legcon.1lag + gdpgrowth.1lag + altinf.1lag  + frefair.1lag  + osorg.1lag   +       + presidential + inc.termlimit.num  +  physinteg.1lag + physintlag.sq + urban.pct + n.events.pre  +  unscheduled_election + regional_success_lag, family = binomial(link = "logit"), data = nelda2.sub[complete_cases, ])
#summary(model.protest.nofraud)

nelda2.sub$propensity_score <- NA

# Assign scores only to complete cases
nelda2.sub$propensity_score[complete_cases] <- predict(model.protest.nofraud, type = "response")

propensity_scores <- predict(model.protest.nofraud, type = "response")

```



```{r stage1-model, echo=FALSE, message=FALSE, warning=FALSE}
models4 <- list()
models4[['Model 9']] <- model.protest.nofraud






modelsummary::modelsummary(models4, gof_omit = "BIC|Log", longtable=F, stars = T, title = 'Logit model of protest occurrence in electoral authoritarian and electoral democratic regimes', coef_map = 
               c("(Intercept)" = "(Intercept)", "presidential1" = "Presidential election", "lji.lag" = "Judical independence (lag)", "legcon.1lag" = "Leg. constraints (lag)", 
                 "gdpgrowth.1lag" = "GDP growth rate (lag)", "altinf.1lag" = "Alternative info. (lag)", 
                 "osorg.1lag" = "Civil soc. openness (lag)", 
                 "inc.termlimit.num" = "Incumbent term-limited", 
                 "urban.pct" = "Urbanization", 
                  "unscheduled_election" = "Off-schedule election",
                 "frefair.1lag" = "Overall elec. integrity (lag)",
                 "n.events.pre" = "Number of pre-election protests", "physinteg.1lag" = "Phys. integrity index (lag)",
                 "physintlag.sq" = "Phys. int. lag squared", "regional_success_lag" = "Regional protest diffusion"
                 ))
```

```{r include=FALSE, echo=FALSE}
# Create propensity score quartiles or tertiles for stratification to account for bias in the propensity scores
#https://onlinelibrary.wiley.com/doi/full/10.1002/sim.5884?casa_token=f64GrFUJvEcAAAAA%3A8a3hmyjFqZMEmT7-sqebxcxa2EvN1QPzP59iThEx_jZEc4x5PgI8OM-vNPOaNU8i3livvOzAyTGpNg
nelda2.sub$ps_quartile[complete_cases] <- cut(propensity_scores, 
                            breaks = quantile(propensity_scores, 
                                            probs = seq(0, 1, 0.25)),
                            include.lowest = TRUE)

summary(lm(v2elirreg.inv ~ ps_quartile , data = nelda2.sub[complete_cases, ]))
```



```{r stage2-models, echo=FALSE}
###lji.lag + legcon.1lag + gdpgrowth.1lag + altinf.1lag  + frefair.1lag  + osorg.1lag   +       + presidential + inc.termlimit.num  +  physinteg.1lag + physintlag.sq + urban.pct + n.events.pre  +  unscheduled_election + regional_success_lag

model.fraud_ps <- lm(v2elirreg.inv ~ propensity_score , data = nelda2.sub)

model.fraud.covars <- lm(v2elirreg.inv ~ lji.lag + legcon.1lag + gdpgrowth.1lag + altinf.1lag  +  + osorg.1lag     + presidential + inc.termlimit.num  +  physinteg.1lag + physintlag.sq + urban.pct + n.events.pre + unscheduled_election + regional_success_lag , data = nelda2.sub)
#summary(model.fraud.covars)

model.fraud.all <- lm(v2elirreg.inv ~ lji.lag + legcon.1lag + gdpgrowth.1lag + altinf.1lag  +  + osorg.1lag     + presidential + inc.termlimit.num  +  physinteg.1lag + physintlag.sq + urban.pct + n.events.pre + unscheduled_election + regional_success_lag + propensity_score, data = nelda2.sub)
#summary(model.fraud.all)



model.fraud.psquart <- lm(v2elirreg.inv ~ lji.lag + legcon.1lag + gdpgrowth.1lag + altinf.1lag  +  + osorg.1lag     + presidential + inc.termlimit.num  +  physinteg.1lag + physintlag.sq + urban.pct + n.events.pre + unscheduled_election + regional_success_lag + ps_quartile, data = nelda2.sub)
#summary(model.fraud.psquart)

models5 <- list()
models5[['Model 7']] <- model.fraud_ps
models5[['Model 8']] <- model.fraud.covars
models5[['Model 9']] <- model.fraud.all
models5[['Model 10']] <- model.fraud.psquart





modelsummary::modelsummary(models5, gof_omit = "BIC|Log", longtable=F, stars = T, title = 'OLS models of election fraud in electoral authoritarian and electoral democratic regimes', coef_map = 
               c("(Intercept)" = "(Intercept)", "presidential1" = "Presidential election", "lji.lag" = "Judical independence (lag)", "legcon.1lag" = "Leg. constraints (lag)", 
                 "gdpgrowth.1lag" = "GDP growth rate (lag)", "altinf.1lag" = "Alternative info. (lag)", 
                 "osorg.1lag" = "Civil soc. openness (lag)", 
                 "inc.termlimit.num" = "Incumbent term-limited", 
                 "urban.pct" = "Urbanization", 
                  "unscheduled_election" = "Off-schedule election",
                 "frefair.1lag" = "Overall elec. integrity (lag)",
                 "n.events.pre" = "Number of pre-election protests", "physinteg.1lag" = "Phys. integrity index (lag)",
                 "physintlag.sq" = "Phys. int. lag squared", "regional_success_lag" = "Regional protest diffusion",
                 "propensity_score" = "Protest risk prop. score",
                 "ps_quartile" = "Prop. score quartile"
                 ))

```


The second-stage models are reported in Table \@ref(tab:stage2-models). Model 7 includes the protest propensity score as the only covariate, as a baseline. For comparsion, Model 8 includes all other covariates but excludes the protest propensity score. Model 9 includes the full set of variables. Finally, Model 10 replaces *protest propensity score* with ordinal indicators for each quartile of latent protest risk. That is, observations with propensity scores in the lowest quartile are assigned a value of 1, those in the second quartile, 2, and so on. This is done as a robustness check against the possible introduction of bias in the model from using a propensity score as a regression covariate [@hadeBiasAssociatedUsing2014].

The results are strongly disconfirmatory for the regime-weakness model. Models 7, 9, and 10 show that latent protest risk (as operationalized by the protest propensity score) is associated with *increased* election-day manipulation. This is the opposite of the prediction that greater protest risk should be associated with restraint on the part of incumbents. We also observe the same relationship with successful regional protest; rather than making incumbents more cautious, the success of a neighboring protest movement appears to increase incumbents' incentives to manipulate.

Moreover, the models in Table \@ref(tab:stage2-models) suggest that factors like judicial independence and the state's repressive capacity retain their sign and statistical significance even after controlling for the underlying risk of protest. This suggests that, even if such features of the political environment do contribute to protest risk, they have additional, separate effects on electoral manipulation. In particular, these results are consistent with the theory that more independent courts increase the risk of punishment for low-level election manipulators, making it more difficult for incumbents to secure their services. Likewise, a more active opposition in the legislature appears to reduce the severity of electoral fraud *after controlling for* the latent likelihood of protest in a case. In turn, this is consistent with a worldview that suggests that greater political fragmentation and contestation deprives the executive of centralized control over resources and the ability to promise impunity to its lower level agents. 


```{r include=FALSE, message=FALSE, warning=FALSE}
## Making plots for Russia chapter

russia <- nelda2.sub %>% filter(country == "Russia (Soviet Union)")

russia_plot_1 <- ggplot(russia, aes(x = propensity_score, y = v2elirreg.inv)) +
  geom_point() +
  labs(x = "Protest risk", y = "Election fraud", title = "Protest propensity score and election fraud in Russia") +
  theme_bw()

russia_plot_2 <- ggplot(russia, aes(x = n.events.pre, y = v2elirreg.inv)) +
  geom_point() +
  labs(x = "N. pre-election protests", y = "Election fraud", title = "Pre-election protests and election fraud in Russia") +
  theme_bw()

png("russia_risk_fraud.png", height = 3, width = 5, units = "in", res = 200)
russia_plot_1
dev.off()


png("russia_preprotest_fraud.png", height = 3, width = 5, units = "in", res = 200)
russia_plot_2
dev.off()
```

## Conclusion


Scholars disagree about how election fraud relates to post-election protest. One family of models — which I have called the regime-weakness model — predicts that incumbents may opt for cleaner elections when the risk of post-election protest is higher. A competing family of models holds that election fraud can itself be a signal of incumbent strength. Building on the principal-agent literature on election fraud, I argue that large-scale fraud draws on the same resources that make incumbents resilient to protest, even when low popularity is reflected in weak vote-shares. Fraud is easier to carry out when incumbents control patronage resources and can shield front-line agents from legal consequences — and those same capacities help incumbents defuse mass protest. Resolving this debate has significant implications for our understanding of both election fraud and authoritarian stability.

This chapter tests the core assumptions of these two models by examining the relationship between election fraud and two dimensions of protest: its initiation and its duration. The key insight is that the regime-weakness and regime-strength models may generate identical predictions about whether protest occurs, but diverge in what they predict about how long protest lasts. Drawing on cross-national data from the V-Dem, NELDA, and ECAV datasets — covering 659 elections in electoral authoritarian regimes — I show that protest is most likely when incumbent vote-shares are very low and fraud is extensive, a finding consistent with both models. More importantly, however, incumbents who engage in larger-scale fraud are able to put down protest movements more quickly.

Additional tests, including on a broader dataset that includes electoral democracies, confirm the regime-strength conception of fraud. When confronted with a higher risk of electoral protest, incumbents tend to produce more election fraud rather than less. This correlation holds true across multiple modeling approaches and measures of manipulation.

Taken together, these findings suggest that the threat of protest is unlikely to constrain election manipulation in the way the regime-weakness model predicts. The incumbents who manipulate most aggressively are also the best equipped to defeat the protests that may follow. This points to electoral authoritarianism as a remarkably durable equilibrium — and paints a bleak picture for those who hope that the threat of protest can discipline incumbents or open pathways to democratization. It also suggests that researchers should look beyond mass protest when modeling the forces that shape electoral manipulation. As subsequent chapters show, a promising approach for upholding election integrity involves breaking the links that tie political principals to their front-line agents.
