# Election fraud and agent risk in Ukraine


The previous chapter shows how the machinery of manipulation has become increasingly effective in Russian elections. The story in that case follows the logic of the regime-strength model of manipulation. In brief: in an effort to forestall challenges by opposition parties and regional elites, President Putin and his allies steadily increased their control over patronage resources, the repressive apparatus, the media, and the judiciary. In turn, even when facing protest risk, the Kremlin was able to recruit agents to falsify election results in places where risk of exposure was low, and to engage in extra-legal vote mobilization in places where that risk was high. It is a case of the vicious cycle of electoral manipulation.

The Ukrainian case provides an opportunity to test the theory in a very different set of circumstances. No ruling party has ever managed to dominate Ukrainian politics in the way United Russia does. Instead, Ukraine's political history from 2002 to 2014 (the period covered here) is one of crisis, alternation in power, polarization, democratization and democratic backsliding. In other words, the case provides an opportunity to test the regime-strength theory of election manipulation at the opposite end of the spectrum from Russia: where the regime is highly competitive and unconsolidated. 

Ukrainian politics at the time was roughly divided into two broad camps: a largely pro-Western camp which came to be known as the Orange coalition, and a largely pro-Russian camp centered around the Party of Regions.^[This is, of course, a shorthand description that overlooks considerable nuance in the two camps.] Each of these camps had durable bases of political support, substantial legislative representation at the national level, and support from wealthy oligarchs. Alternative models of election manipulation make distinct predictions for the Ukrainian case, as a result. 

Models that assume election manipulation signals weakness to elites and opposition actors would predict restraint on incumbents' part. In a closely divided polity, where the opposition is well-resourced and well-organized, incumbents would anticipate that substantial, costly protest would be the outcome of any manipulation effort large enough to make a difference in the election. Likewise, a grievance-based model of electoral protest would predict election manipulation would galvanize a large portion of the population---supporters of the out-of-power group---to protest. Finally, Ukraine during this period likely fails to meet Simpser's expectations for excessive or blatant manipulation, given the absence of an unconstrained government and a dominant ruling party.

As this chapter will show, none of these patterns are quite borne out in the election-forensic data, or in the crucial example of the Orange Revolution. Instead, overall patterns of manipulation fit the expectations of the principal-agent model of electoral manipulation. More importantly, this chapter shows how variation in falsification and other forms of manipulation fit the expectations of the principal-agent model. In Ukraine, where the risk of exposure is much higher than in Russia, highly attributable falsification is scarce. Less attributable (but more public) forms of manipulation, like voter pressure and vote-buying, are most common when patronage resources are highly consolidated and agents' risk of punishment is low.

## Ukraine's political history, 2000 to 2014

[*This section will need edits for focus; is largely drawn from dissertation*]

Though analysts should be wary of over-emphasizing the point, Ukraine’s political divisions are largely regional.  Voters in the country’s east have generally supported parties and candidates that adopt a more interventionist economic policy and take a more favorable view toward Russia.  By contrast, voters in the west have usually supported parties and candidates with economic and foreign policies that are oriented towards the market and towards Europe (Birch, 2000).  The two regions are characterized by ethnic and linguistic divisions, with the west home to more ethnic Ukrainians and Ukrainian-speakers, while Russian ethnicity and language is more common in the east. 

These differences can be attributed, in part, to historical legacies: much of western Ukraine was ruled by the Austro-Hungarian Empire prior to World War I, and by Poland, Romania, or Czechoslovakia during the interwar years.  Government policy in these regions, before they were incorporated into the Soviet Union in 1939, was relatively democratic and favorable to the development of Ukrainian national identity. By contrast, eastern Ukraine was part of the Russian Empire, which emphasized a Ukrainian identity that was a regional subset of the broader Russian ethnicity and language [@Katchanovski2006]. Despite the well-documented importance of language [@fournier_mapping_2002;@kulyk_language_2011] and ethnicity [@barrington_examining_2002] in Ukraine, however, the regions’ voters are not homogeneous blocs. For example, using a survey experiment, Frye (2015) finds that a hypothetical candidate’s position on closer ties with Europe or Russia is more likely to drive a voter’s preference than either the candidate’s reported ethnicity or language. Furthermore, many of Ukraine’s eastern regions are more diverse linguistically and ethnically than is generally reported [@Sasse2010]. 

Finally, as in other post-communist countries, Ukraine’s economic transition created a class of oligarchs who were able to gain control of the state’s major economic assets.  In Ukraine, these oligarchs are often linked in regional clans and can use their resources to back local politicians.  These patterns can be seen even in independent Ukraine’s earliest elections, but intensified following the Orange Revolution [@DAnieri2011]. Spoken language in Ukraine is an inexact proxy for ethnicity, which in turn is an inexact proxy for identity and political orientation [@Sasse2001].  

While regional attitudes follow the general contours described above, it is important to note that major Ukrainian political parties have not been characterized by stark ideological conflicts so much as by regional loyalties, affiliation with particular leaders, and the distribution of patronage. Parties in Ukraine are highly personalized, and frequently depend on the resources, reputation, and charisma of particular leaders for their electoral success [@Rybiy2013]. This is especially true of the western-oriented parties associated with Viktor Yushchenko and Yulia Tymoshenko [@Kuzio2012;@Kudelia2015]. This personalization contributes to parties’ ideological fluidity^[Some parties are exceptions to this rule: the ideologically cohesive Communist Party of Ukraine and Svoboda (Freedom) did well in the 2012 parliamentary election, winning 13\% and 10\% of the vote, respectively [@Rybiy2013].] and voters’ difficulty in placing parties on an ideological spectrum [@Rybiy2013].  

This weak ideological commitment on the part of voters, combined with social inequality and overlapping political and economic responsibilities of local leaders [@Allina-Pisano2010], allows parties to rely on `administrative resources' like access to state employment, favorable (or unfavorable) business treatment, and selective law enforcement to build and maintain networks of supporters [@DAnieri2011;@protsyk_centre_2003]. Tools like these give political leaders the ability to raise funds, mobilize voters, and manipulate elections [@Hale2003]. These tools are usually held by regional or local actors in Ukraine, rather than being institutionalized at the national level [@Way2005]. In particular, such methods are characteristic of the pro-eastern parties exemplified by the Party of Regions [@Kudelia2015;@Kuzio2012].

A large public sector and a relatively compliant judiciary [@kudelia_sources_2012] provide ample opportunity for the executive branch to apply pressure to voters and opposition figures in Ukraine. Post-Soviet reforms subordinating regional governors to the president created ``a strong vertical executive power running from the president down to the village level"; in order to retain office, regional and local officials were required to mobilize voters on behalf of their bosses [@Konitzer-Smirnov2005]. As in Russia [@frye2014], state employees and workers in industries that are vulnerable to state regulation are susceptible to pressure from above.  In one particularly stark example, school children at a school in eastern Ukraine informed their parents that their teachers would be fired if the parents did not support the government in an upcoming election [@danieri_understanding_2007]. During the 2004 presidential election, which sparked the Orange Revolution, the government leaned on administrative resources to encourage “the police, the army, prisons, public universities, schools, and hospitals, to campaign and vote for Prime Minister Viktor Yanukovych,” who enjoyed the backing of the sitting president and many Ukrainian oligarchs [@Katchanovski2008]. Moreover, government officials engaged in tampering with the electronic results of the second round of the election; the combined efforts of mobilization and falsification appeared sufficient to tip the election in favor of Yanukovych and against his pro-Western rival Viktor Yushchenko—a rare example of a stolen election [@simpser2013].The Party of Regions, which was led by Yanukovych and came to be the dominant party in eastern Ukraine and Crimea, relied on a mix of state administrative resources, political patronage, and intra-elite bargaining to maintain its support in the region [@Kudelia2015;@myagkov_forensics_2009]. Patronage, clientelism, and voter pressure are all common features of the Ukrainian election system, and—like Russia—it is characterized by a political hierarchy that links the president to local officials through intermediaries like regional officials and oligarchs. As a result, we might expect to see widespread electoral manipulation under either the signaling or local-information frameworks.

One peculiar feature of election management in Ukraine is an apparent emphasis on deflationary tactics; that is, efforts to reduce one’s opponents vote-share rather than (or in addition to) inflating one’s own. Ukrainian parties have attempted to influence election results by using manufactured parties and candidates to drain votes away from rivals and to gain additional influence on election commissions. Ukrainian elections have often been characterized by an abundance of so-called ‘technical parties,’ small parties artificially created by main parties in order to divert votes away from their opponents. One such technical party, which voiced strong opposition to Yanukovych and the Party of Regions and whose leader was reminiscent of former Orange-Revolution hero Yulia Tymoshenko, diverted as much as 300,000 votes from the nationalist party UDAR, depriving it of six to eight seats in parliament [@kovalov_electoral_2014]. Perhaps more important than confusing voters and diverting votes, however, is the role that technical parties play on Ukraine’s election commissions.

Ukraine’s election law divides the country into 225 district election commissions. Rules determining membership on the committees have varied from year to year, but have generally granted automatic representation to representatives of the parliamentary parties, with remaining seats filled by non-parliamentary parties via lotteries. This setup allows for small, hastily registered parties—operating informally on behalf of major parties or politicians—to gain representation on the district commissions. This problem became especially egregious in 2012, when the Central Election Commission implemented a change in the lottery process five days before the drawing: instead of individual lotteries for each district commission, a single lottery would be held that would determine the makeup of all 225 commissions. As a result, small parties that had nominated only one or two candidates nationwide gained representation on all district commissions, while larger opposition parties were excluded [@kovalov_electoral_2014;@OSCEODIHR2013]. District election commissions then implemented a similar nominating procedure for all their subordinate precinct election commissions.

This system can be especially problematic when combined with the deep patronage roots enjoyed by the Party of Regions in eastern Ukraine, which it used to gain outsize influence on the commissions. In one notable example, OSCE observers in a single election district of Donetsk oblast' determined that 1,667 of the 2,551 precinct election commissioners in the district worked for an enterprise headed by the Party of Regions candidate—meaning that more than two-thirds of precinct commissioners presided over balloting for their employer [@OSCEODIHR2013]. While the presence of technical representatives on election commissions was perhaps most extensive in the 2012 campaign, it figured prominently in previous elections as well. In the first round of the 2004 election, technical candidates had the right to nominate up to 66,000 members of precinct election commissions; while both major candidates took advantage of technical candidates, the OSCE concluded that Viktor Yanukovych benefited the most, finding that supportive technical candidates gave him ``a de facto majority on election commissions'' [@OSCEODIHR2005]. The 2010 election, in which Yanukovych was first elected, was relatively free of electoral fraud, with candidates emphasizing voter mobilization [@Herron2011], though the analysis presented here suggests that some of these mobilization efforts may have been extra-legal and/or characterized by the misuse of state resources [@OSCEODIHR2010]. 

Partisan control of election commissions at various levels facilitates electoral manipulation in various forms, which observers frequently witnessed in Ukraine. In 2004, the OSCE observed—among other issues—ballot box stuffing, irregularities with absentee voting, large numbers of voters being turned away from the polls, implausible turnout figures, and suspicious tabulation process. All of these forms of electoral manipulation are either implemented or influenced by election commissioners at the precinct, territorial, or central level.  Similar problems were observed by the OSCE in 2012, when observers negatively assessed the vote count in eleven percent of monitored precincts, and tabulation was negatively assessed in almost half of the 161 district election commissions monitored [@OSCEODIHR2013].


### Frustrated efforts to consolidate power

Soviet Ukraine was characterized by heavy reliance on patronage and clientelism, transmitted by informal networks operating within the formal Communist Party and command economy structures in a system Kitschelt calls [-@Kitschelt1995] terms patrimonial communism. This framework persisted in post-Soviet Ukraine, as many political and bureaucratic officeholders remained in place, while communist-era managers and officials became private-sector owners of capital [@Zimmer2008]. Especially in the heavily industrialized eastern parts of the country, where small-scale enterprises are less common [@Aslund2005], large portions of the population are dependent on state- or employer-based patronage [@Zimmer2005]. A power vertical has been solidly established in the eastern part of the country, based on an alliance of regional and local administrative officials and business actors [@Zimmer2005]. Nationally, the pro-presidential constitution of 1996 subordinated regional governors and local executives to the president [@Konitzer-Smirnov2005], a structure which remained largely unchanged under the post-Orange constitution. 

However, the Cassette Scandal crisis\textemdash in which illicit recordings of the president allegedly ordering a kidnapping, among other unsavory details, were made public\textemdash limited the ability of President Kuchma to further consolidate control over the patronage resources and hand them to his preferred candidate in the 2004 presidential election, Viktor Yanukovych. As a result of the ensuing political crisis, the party system fractured, oligarchic alliances were ruptured, momentum toward a super-presidential constitution was scuppered, President Kuchma’s succession plans were thrown into turmoil, and an opposition movement based in civil society began to form [@Kuzio2007]. The scandal helped lay the foundations for the 2004 Orange Revolution, in which hundreds of thousands of citizens took to the streets to protest electoral manipulation in the presidential elections. The subsequent opposition victory in a repeat election upended the party structure, and helped usher in a new constitution that divided power more evenly between the presidency and the parliament.

As part of a compromise following the Orange Revolution, a new constitution was adopted which greatly increased the patronage capabilities of the parliament and prime minister. As a result of the new constitution, the parliamentary majority gained more discretion over the nomination of prime ministers, more influence over the defense and security ministries and the prosecutor general, and shifted much executive staffing power away from the president and to the cabinet [@Christensen2005]. This shift reinforced the de facto competing-pyramid structure of patronage in the post-revolutionary period, by expanding and legitimizing the influence of the parliamentary majority while still retaining some patronage powers in the office of the presidency [@hale_patronal_2014]. 

The competing pyramid structure would not survive long into the Yanukovych presidency, when the Party of Regions was eager to press its advantage in patronage resources. After defeating Tymoshenko’s network in the 2010 presidential election, Yanukovych moved swiftly to encourage elite defections within the Rada, building a pro-presidential coalition that secured control of the premiership [@hale_patronal_2014;@kudelia_sources_2012]. Subsequently, Yanukovych was able to win restoration of the pre-revolution constitution favoring the president; he rapidly reconsolidated patronage networks that had become fragmented during the Orange period. 

This centralizing period would be short-lived however, as massive crowds during the 2014 Euromaidan crisis helped depose Yanukovych, leading to the seizure of the Crimean peninsula by Russia, the de facto secession of two of Ukraine’s most populous and industrialized regions, a prolonged civil war, and the effective collapse of the pro-eastern Party of Regions. The patronage networks that had helped propel the success of the Party of Regions locally and nationally began to unravel [@hale_patronal_2014]. 

### The role of the oligarchs

In addition to the shifting consolidation of state-based patronage, analysts of Ukrainian politics must also contend with a complex arrangement oligarchic ‘clans,’ which deploy their resources to support favored parties. Unlike in Russia, where the major oligarchs generally do not risk challenging the Kremlin Ukraine’s oligarchic clans have been more fractious. Focusing on pro-eastern parties helps navigate this dynamic, since the pro-eastern oligarchic clan is relatively more unified than their pro-western counterparts. During Kuchma’s presidency, three regional oligarchic clans became influential in politics: the Dnipropetrovsk network, the Donetsk network, and the Kyiv network. All three were generally aligned with Kuchma until the Orange Revolution [@Pleine2016], and used their resources to influence election outcomes [@Katchanovski2008]. The revolution split the political-oligarchic patronage networks into three competing camps\textemdash one each aligned with Viktor Yushchenko, Yulia Tymoshenko, and Viktor Yanukovych\textemdash a framework that persisted until the 2010 presidential election [@Hale2010]. Even after being cast into the political wilderness nationally by the Orange crowds, Yanukovych and the Party of Regions could rely on regional patronage networks and elected local mayors to retain influence in the east [@Hale2010]. 

When Yanukovych assumed the presidency in 2010, the fractiousness of the opposition coalition of oligarchs gave way once again to a more united patronage pyramid around the presidency and the Donetsk clan. In turn, this informal patronage network collapsed as a result of the Euromaidan and subsequent civil war, generally reconstituting itself around oligarch Petro Poroshenko after his 2014 presidential victory [@Pleine2016]. Poroshenko, though his own election was largely clean, used patronage resources to good effect (especially in the East, where networks still remained largely intact) in the subsequent legislative elections [@Fedorenko2016].

^ *Dissertation text here*; Edit to expand on oligarchs' role in producing manipulation (specifically, they can pay for stuff but can't promise legal impunity)

### Risks of manipulation exposure


Given the divisions of the country, local constraints in Ukraine must be conceptualized for each broad political group, referred to here by the shorthand terms ‘pro-western’ and ‘pro-eastern.’ Risks will vary not only by local political conditions, but by party group; in one particular locality the risk of exposure might be severe for pro-western agents, but non-existent for pro-eastern agents. For example, the pro-eastern Party of Regions has been able to exert influence over local courts in regions it controls [@Katchanovski2008], reducing the risk that local agents will face punishment for supporting the party in illegal ways but perhaps increasing the risk for supporters of pro-western parties. The Party of Regions inherited much of the political culture and resources of the Communist Party of Ukraine, and used these advantages to build a locally dominant coalition of government administrators, red directors (communist-era managers of large firms), and local business / oligarchic interests [@Kudelia2015;@Zimmer2008]. As a result, the heartland of the PoR is politically sealed off [@Osipian2006].

The situation is more complicated in regions where pro-western parties tend to dominate. Machine politics, essential to delivering votes in places like Donetsk and Luhansk in the east, were difficult to export to territories in Western Ukraine [@Sushko2002;@Zimmer2005], due in large part to a more politically active civil society there [@DAnieri2005].  The presence of a more active civil society and the relative absence of locally dominant economic interests led to the development of more highly personalized parties, primarily oriented around the figures of Yushchenko and Tymoshenko during this period [@Kudelia2015;@Kuzio2012]. As a result, these two blocs competed with each other for votes within their territories [@Kuzio2013], limiting the ability of either bloc to develop the kind of local political monopoly that reduces risks toward election manipulating agents.

The development of active civil society in the west accelerated as a result of the Kuchma-gate crisis, and was nurtured by opposition parties during the Orange Revolution [@Kudelia2007]. However, the historical roots of civil society and multiparty democracy run deeper in Western Ukraine than in the East. During perestroika, dissidents and former political prisoners in Western Ukraine formed Rukh (the Ukrainian Popular Movement for Restructuring) to challenge the Communist Party; while Rukh gained popularity in Kyiv and central Ukraine, it never gained much of a foothold in the east [@Kuzio2010]. In surveys from the early post-Soviet period, respondents from western Lviv were twice as likely as residents of Donetsk to attempt to ``resolve social problems'' by contacting mass media, contacting officials, or joining an organization, and three times as likely to have signed a petition or participated in a demonstration [@Aberg2000]. These connections persisted through the Orange Revolution: protesters in that movement were motivated less by democratic ideology and more by the fact that they were highly networked, with higher incomes, and less dependent on the state [@Beissinger2013].

In sum, local risks vary considerably across Ukraine by both party and region. The risk of exposure for pro-western agents is especially high in eastern regions where the political machine organized around the Party of Regions predominates, due to that party’s success in creating a subnational enclave prior to the Euromaidan crisis, Russian intervention, and civil war. However, the existence of competing pro-western parties and a robust civil society in many western regions makes electoral manipulation difficult to achieve undetected in those regions as well. These differences are largely structural, as described above; they rest on broad historical and social factors rather than on parties’ see-sawing electoral fortunes. As a result, the level of risk in a particular locality does not shift much over time in Ukraine. Instead, shifts in manipulation patterns over time are due largely to major changes occur in the consolidation of patronage during the period studied.

^ *The above is dissertation text*

Will also need to add details on election monitoring by the orange coalition (OPORA?) and evidence of legal cases, and how the judiciary works in Ukraine (i.e. vulnerable to outside pressure)


## Hypotheses and other empirical tests

The regime-strength model of electoral manipulation claims that incumbents reduce agency loss among front-line agents to the extent that they are able to develop tools of social control that reduce agents' level of risk. Where incumbents accomplish this goal, as in Russia, the model predicts high levels of manipulation, with falsification occurring in places where the regime is most consolidated and less efficient forms of manipulation occurring where risks are locally higher.

In Ukraine, agency loss is expected to be high. Uncertainty over which camp will control national executive over the medium term---and thus uncertainty over who will control the prosecutorial service and have influence over the judiciary should increase the expected cost of engaging in falsification. This risk is exacerbated by the widespread presence of partisan and civil-society observers in precinct and district election commissions (**add OSCE cites here**). It is thus expected that falsification will be unsystematic, and forensic measures of fraud will therefore be uncorrelated with parties' vote-shares. In other words, I expect that the null hypothesis will *not* be rejected in statistical tests of that correlation; since my expectation is that the null will not be rejected, I list this as an empirical implication of the theory, rather than a hypothesis. 

Instead, I predict that other forms of manipulation will predominate, as principals recruit agents for tasks further down the attribution ladder (see Chapter 2). When outright falsification is unavailable, incumbents with resources will seek to conver those resources into votes through ballot-stuffing, vote-buying, and voter-pressure. All of these tools are less cost-effective than falsification however, and open up the possibility that agents will accept patronage benefits without fully delivering their share of votes. This suggests two hypotheses.

> H1: Overall manipulation in favor of a party will increase with that party's control over patronage resources.

As a corollary of this hypothesis, it is expected that oligarchs can make use of their own regional patronage networks to boost election outcomes for a favored party.

>H2a: Overall manipulation in favor of a party will increase in regions where  supportive oligarchs control significant assets.

>H2b: The effect of supportive oligarchs on manipulation should increase with urbanization.

^ Explain H2 is an indicator of Soviet centralization.


## Data and methods

To test these hypotheses, I draw on precinct-level results from nine elections in Ukraine from 2002 to 2014. This includes three presidential elections---the third round of the 2004 elections, along with both rounds of the 2010 and 2014 elections---and four legislative elections (2002, 2006, 2007, and 2012). Unfortunately, the dataset excludes the first two rounds of the 2004 presidential elections, for which precinct-level results are not publicly available. Each election-year includes data from approximately 30,000 to 35,000 precincts nested within Ukraine’s 27 regions. As a result of the annexation of Crimea and the conflict in Eastern Ukraine, the legislative and presidential elections in 2014 draw on approximately 28,000 precincts within 23 regions: Crimea, Sevastopol, Donetsk, and Luhansk regions drop out of the data. This precinct-level data is used to conduct election-forensic estimates of falsification and overall evidence of electoral manipulation. Unlike in the Russian case, I am unable to estimate levels of vote-buying and related tactics, since Ukraine did not release precinct-level numbers of absentee or mobile-ballot-box voting until the 2014 elections. Explanatory and control variables are taken at the regional level.



### Dependent variables

Two dependent variables are estimated for each region in each election year, using precinct-level data. First, an estimate of overall manipulation is calculated by regressing the absolute-vote share for the largest pro-Western and pro-Eastern parties (per election) on precinct-level turnout [@myagkov_forensics_2009], using multilevel models with random slopes and intercepts.^[This allows for the turnout coefficient to vary freely by region, an accounts for varying 'baseline' support for each party in the region.] Such a test assumes that, in a clean election, a party’s absolute vote-share (the party’s number of votes divided by the number of registered voters) should increase consistently across precincts within a country, as precinct turnout increases. This approach assumes homogeneous precincts across the country, an assumption which is bolstered by testing precincts within regions or through the use of relevant socioeconomic control variables to weed out potentially confounding relationships [@deckert2013patterns]. I employ both approaches here. Figures \@ref(fig:hist-east) and \@ref(fig:hist-west) show the distribution of coefficients for the major parties.  

Second, election falsification is estimated by measuring the degree to which the ones-digit across a party's vote totals deviate from the expected uniform distribution, by region and year [@Skovoroda2017]. Whereas the analysis of the Russian case presented in Chapter 4 used a stricter measure---deviance in last-digit zero only---the analysis of Ukraine uses deviance across all digits. This is a more sensitive measure, intended to pose a tougher test of the implication that falsification in Ukraine will be rare and non-systematic. As an additional robustness check, I also use Beber and Scacco's [-@beber2012] binary indicator of severe manipulation, which takes a value of 1 if last-digit deviance is statistically significant at the p < .05 in a chi-square test.



```{r}
ggplot(merged.data, aes(g_statistic_east)) + geom_histogram()
```













<!--In sum, Ukraine is a case where many of the conditions are in place for widespread electoral manipulation to occur. Patronage systems are embedded in society, especially in eastern regions of the country. Presidential influence over governors potentially unites the state apparatus in a network that can mobilize patronage resources to deliver votes for the incumbent. There are deep links between the state and business interests [@kudelia_sources_2012]. Nevertheless, the actual level of manipulation has varied considerably between elections and across regions, according to international monitors and experts [@Hale2010;@Herron2010;@Herron2011;@Herron2014;@Katchanovski2008]. This variation can be explained largely by shifts in patronage consolidation. Patronage is substantially less consolidated than in Russia; there is no ruling party akin to United Russia and oligarchic clans compete to influence election results. The constitution has at times fostered competition between patronage networks affiliated with the president and those affiliated with the prime minister. Local constraints are generally higher where pro-Western parties prevail than in pro-Eastern regions, but were elevated generally by the electoral reforms that were in force from the third round of the 2004 presidential election through the 2010 presidential election.

Of the three cases considered here, Ukraine is in several ways the most difficult in which to test the consolidation and constraint model of electoral manipulation. Parties in Ukraine, especially those with roots in its western regions, have tended to be personalized and short-lived, making it more difficult to track their electoral performance over time. Additionally, over the time period for which electoral data are available (2002 to 2014), Ukraine were beset by multiple political crises. The two most significant of these include the 2004 Orange Revolution and the 2014 Euromaidan revolution and subsequent military conflict in the east. These are by no means the only crises to befall the country, however, and it is telling that the so-called Cassette Tape or Kuchma-gate scandal—which implicated President Kuchma in a variety of high-level misdeeds including the abduction and murder of a critical journalist—is considered a lesser incident in recent Ukrainian political history. While these patterns make it more difficult to analyze patterns of electoral manipulation than in Russia, for example, where a single ruling party predominates, it also offers opportunities to test the theory in the context of large shifts in patronage consolidation.-->

```{r}
library(here)
library(tidyverse)

ukrdata <-read.csv("C:/Users/colej/Documents/Grad school/Dissertation/Ukraine main dataset.csv")  #This dataset is ready for turnout analysis
urban.pct <- ukrdata$urban.population / ukrdata$population
logpop <- log(ukrdata$population)
loggov <- log(ukrdata$gov.employ)
logold <- log(ukrdata$over.sixty)
logincome <- log(ukrdata$income.percap)
west.pres.approval <- ukrdata$west.president * ukrdata$pres.approval
east.pres.approval <- ukrdata$east.president * ukrdata$pres.approval
ukrdata <- cbind(ukrdata, urban.pct, logpop, loggov, logold, logincome, west.pres.approval, east.pres.approval)
ukrdata$prewar.west <- as.factor(ifelse(ukrdata$region == "Ivano-Frankivska" | 
                                ukrdata$region == "Lvivska" |
                                ukrdata$region == "Ternopilska" |
                                ukrdata$region == "Chernivitska" |
                                ukrdata$region == "Zakarpatska" | 
                                ukrdata$region == "Volinska" |
                                ukrdata$region == "Rivnenska", 1, 0)) 

ukrdata$galicia <- ifelse(ukrdata$region == "Ivano-Frankivska" | 
                            ukrdata$region == "Lvivska" |
                            ukrdata$region == "Ternopilska", 1, 0)

ukrdata <- ukrdata %>% mutate(prewar.east = ifelse(region == "Donetska" | region == "Luganska" | region == "Kharkivska", 1, 0))

ukrdata <- ukrdata %>% mutate(PoR_bastion = ifelse(region == "Donetska" | region == "Luganska", 1, 0))

ukrdata <- ukrdata %>% mutate(crimea = ifelse(region == "Crimea" , 1, 0))

ukrdata <- ukrdata %>% mutate(prewar.region = as.factor(ifelse(region == "Crimea", "Crimea", ifelse(prewar.west == "1", "West", 
                                                     ifelse(prewar.east == 1, "East", "Central")))))

ukrdata$party.margin <- NA
ukrdata$party.margin <- ifelse(ukrdata$east.party.margin > 0, ukrdata$east.party.margin, 
                               ((ukrdata$west.party.margin) * -1)) #party.margin: negative for w. parties, positive for e. parties
ukrdata$pres.approval <- ukrdata$pres.approval / 100

ukrdata$post.orange <- NA
ukrdata$post.orange <- ifelse(ukrdata$year >= 2004,
                                    1, 0) #Marks after 2004 protests

ukrdata$orangerev <- NA
ukrdata$orangerev <- ifelse(ukrdata$year == 2004,
                              1, 0) #Marks after 2004 protests

ukrdata$wartime.election <- NA
ukrdata$wartime.election <- ifelse(ukrdata$year == 2014,
                            1, 0) 


ukrdata$pres.constitution <- NA
ukrdata$pres.constitution <- ifelse(ukrdata$year <= 2004 | ukrdata$year > 2010 & ukrdata$year < 2014,
                                    1, 0) #Marks when constitution was more presidential, based on 2004 reform agreement
#ukrdata$crisis.name <- relevel(ukrdata$crisis.name, ref = "None")

ukrdata$east.roz.percent <- ukrdata$e.roz.count.total / ukrdata$n.precincts
ukrdata$west.roz.percent <- ukrdata$w.roz.count.total / ukrdata$n.precincts


#ukrdata$pres.constitution2 <- NA
#ukrdata$pres.constitution2 <- ifelse(ukrdata$year <= 2006 | ukrdata$year > 2010 & ukrdata$year < 2014,
#                                    1, 0) #Marks when constitution was more presidential, based on implementation of reform

fraudscores <- read.csv("C:/Users/colej/Documents/Grad school/Dissertation/Ukraine independent variables/Ukraine fraud scores.csv")

merged.data <- ukrdata %>% left_join(fraudscores, by = c("regionid", "electionid"))

merged.data$president.name <- relevel(as.factor(merged.data$president.name), ref = "Poroshenko")

#Deviance scores
deviancescores <- read.csv(here("Ukraine_forensics", "Data", "region_deviance_ukraine.csv"))

merged.data <- merged.data %>% left_join(deviancescores, by = c("region", "electionid"))

merged.data <- merged.data %>% mutate(east.party.margin2 = east.party.voteshare - west.party.voteshare)

merged.data <- merged.data %>% mutate(electionid = as.factor(electionid))

## Relevel elections to set 2012 as baseline if wanted (otherwise 2002 is baseline)


#merged.data$electionid <- relevel(merged.data$electionid, ref = "2012_0_0")

```

```{r}
#Adding in total deviance (not just last digit zero deviance)


ukraine_ids <- read.csv("C:\\Users\\colej\\Documents\\Grad school\\Dissertation\\Region ID lists\\ukraine region ids.csv")
east_party_deviance <- read.csv(here::here("Ukraine_forensics", "Data", "ukraine_full_deviance_east.csv"))
west_party_deviance <- read.csv(here::here("Ukraine_forensics", "Data", "ukraine_full_deviance_west.csv"))

east_party_deviance <- east_party_deviance %>% rename(region = group)
west_party_deviance <- west_party_deviance %>% rename(region = group)
ukraine_ids <- ukraine_ids %>% rename(region = name)


east_party_deviance <- east_party_deviance %>% left_join(ukraine_ids, by = "region")
west_party_deviance <- west_party_deviance %>% left_join(ukraine_ids, by = "region")

## Remove unneeded variables for merging

east_party_deviance <- east_party_deviance %>% dplyr::select(!c(dataset, warning, year, presidential, round.2))
west_party_deviance <- west_party_deviance %>% dplyr::select(!c(dataset, warning, total_rows, year, presidential, round.2))

## Merging back in with merged.data

merged.data <- merged.data %>% left_join(east_party_deviance, by = c("region", "electionid"))
merged.data <- merged.data %>% left_join(west_party_deviance, by = c("region", "electionid"))

merged.data <- merged.data %>% mutate(dev_1_9_east = g_statistic_east - deviance_region_east)
merged.data <- merged.data %>% mutate(dev_1_9_west = g_statistic_west - deviance_region_west)

merged.data <- merged.data %>% mutate(year = as.factor(year))


```


```{r}
## Adding in oligarch count data

ukraine_oligarchs <- read.csv(here::here("Ukraine_forensics", "Data", "ukraine_oligarch_count.csv"))

ukraine_oligarchs <- ukraine_oligarchs %>%
  pivot_longer(
    cols = starts_with("east_oligcount"),
    names_to = "year",
    names_prefix = "east_oligcount_",       # removes "east_count" prefix
    values_to = "east_oligcount"
  )
ukraine_oligarchs <- ukraine_oligarchs %>% mutate(year = as.factor(year))

## Merging back in with merged.data

merged.data <- merged.data %>% left_join(ukraine_oligarchs, by = c("regionid", "year"))
```

<!--
In particular, the Ukrainian case provides a useful test against the hypotheses of the local-information model of electoral manipulation [@rundlett2016deliver].^[Ukraine is less effective as a test of the signaling model of manipulation, since it does not meet Simpser’s requirements for excessive manipulation to occur: an unconstrained government and a dominant ruling party.] The local-information model claims that when election-manipulating agents perceive their patron’s popularity to be low locally, they extrapolate from this information that their patron is likely to be unpopular nationally and begin to doubt that she will win the election. As a result, they become less likely to engage in electoral manipulation, fearing that patronage rewards will be revoked and the risk of punishment will increase. This theory produces two testable hypotheses that compete with those of the consolidation and constraint model. First, it expects that agents in local strongholds will continue to engage in manipulation on behalf of the locally dominant party, even if national political conditions worsen for that party. As will be shown below, this pattern does not emerge in Ukraine, where manipulation efforts against the pro-eastern Party of Regions decline dramatically—even in pro-western regions—during a period of high patronage consolidation. Second, the local-information model predicts that agents in areas where a party is unpopular will refrain from manipulation on behalf of that party, even if that leader is dominant at the national level; instead, evidence of manipulation in favor of pro-western parties can be seen even in highly pro-eastern regions in the aftermath of the Maidan protests.  

In both cases, the results of election-forensic analysis of data from Ukraine support the consolidation and constraint model, and do not uphold the local information approach. I test these two hypotheses using precinct-level election results for Ukraine’s major pro-eastern party in each election year from 2002 to 2014.^[In most cases this party is the Party of Regions; however in 2002 it the Communist Party of Ukraine, and in 2014 the Opposition Bloc (essentially the rump Party of Regions following the Maidan crisis).] Analysis of the east-party data is more likely to be useful for testing the theory than the pro-western parties. First, the eastern parties—based in the more industrial part of the country—are more strongly rooted in patronage politics than the major pro-western parties. And it is the pro-eastern Party of Regions that provides the only instance during the time period studied where a party gains unambiguously consolidated control over state and para-state patronage resources—the 2012 legislative election during which the Party of Regions holds the presidency and control over the legislature. This control over patronage networks then rapidly deteriorates in the wake of the 2014 Euromaidan protest movement, annexation of Crimea, and prolonged conflict in the Donbass region. This trajectory makes it possible to test the effect of patronage consolidation on east-party results. The pro-western parties do not achieve high levels of patronage consolidation, due to splits between the major pro-western leaders, constitutional reforms, and a societal base of support less dependent on patronage resources.

Lastly, the Ukrainian case also offers an opportunity to test the theory on a different class of manipulation techniques. While the ruling party in Russia appears to mostly engage in inflationary manipulation techniques (at least on election day itself), deflationary tactics appear to predominate in Ukraine. In addition to mobilizing tools like vote-buying or voter pressure, which can only be employed where a principal’s patronage network extends, party’s in Ukraine have also tried to use ‘technical parties’ and ‘clone candidates’ to divert votes away from their genuine opponents. More importantly for this study, such techniques have also allowed parties to stack regional and local electoral commissions with pliant supporters; enabling fraud and other forms of electoral malpractice. Testing the consolidation and constraint model in this context indicates that agents weigh the benefits of patronage against local risks even when employing these kinds of techniques, which may be subtler and harder to observe than tools like ballot-stuffing.

In sum, this chapter shows that deflationary manipulation is characteristic of elections in Ukraine, and that the local information model does not accurately predict those patterns of electoral manipulation. Using two election-forensic techniques, it demonstrates that deflationary techniques deployed against pro-eastern parties and candidates are less likely in regions with higher local constraints against such manipulation, and respond as predicted to shifts in patronage consolidation. That is, during the Yanukovych presidency when patronage resources were highly consolidated around the Party of Regions, evidence of deflationary manipulation against that party drops dramatically, even in highly pro-western regions. In turn, in elections after the Euromaidan crisis, evidence of deflationary manipulation against pro-eastern candidates increases, especially in pro-eastern regions. These results contradict the expectations of the local information model, and support the theory articulated here.

-->









## Consolidation and constraint in Ukraine

The above review makes clear that the conditions exist in Ukraine for parties and candidates to make use of illegal electoral manipulation. International election observer reports indicate that such tactics were substantial in the legislative elections of 2002, 2012, and 2014, as well as during the first two rounds of the 2004 presidential election, with considerable regional variation in the severity of manipulation efforts [@OSCEODIHR2013;@OSCEODIHR2005]. Other elections were considered to be markedly cleaner [@Hale2010;@Katchanovski2008;@OSCEODIHR2007;@OSCEODIHR2010;@OSCEODIHR2006]. While the local-information model attempts to explain this variation via the popularity of candidates in agents’ local territories, it is more accurately explained through a focus on local risks to agents and shifts in the consolidation of patronage networks around the Party of Regions. 




### Patronage consolidation

Soviet Ukraine was characterized by heavy reliance on patronage and clientelism, transmitted by informal networks operating within the formal Communist Party and command economy structures in a system Kitschelt calls [-@Kitschelt1995] terms patrimonial communism. This framework persisted in post-Soviet Ukraine, as many political and bureaucratic officeholders remained in place, while communist-era managers and officials became private-sector owners of capital [@Zimmer2008]. Especially in the heavily industrialized eastern parts of the country, where small-scale enterprises are less common [@Aslund2005], large portions of the population are dependent on state- or employer-based patronage [@Zimmer2005]. A power vertical has been solidly established in the eastern part of the country, based on an alliance of regional and local administrative officials and business actors [@Zimmer2005]. Nationally, the pro-presidential constitution of 1996 subordinated regional governors and local executives to the president [@Konitzer-Smirnov2005], a structure which remained largely unchanged under the post-Orange constitution. 

However, the Cassette Scandal crisis\textemdash in which illicit recordings of the president allegedly ordering a kidnapping, among other unsavory details, were made public\textemdash limited the ability of President Kuchma to further consolidate control over the patronage resources and hand them to his preferred candidate in the 2004 presidential election, Viktor Yanukovych. As a result of the ensuing political crisis, the party system fractured, oligarchic alliances were ruptured, momentum toward a super-presidential constitution was scuppered, President Kuchma’s succession plans were thrown into turmoil, and an opposition movement based in civil society began to form [@Kuzio2007]. The scandal helped lay the foundations for the 2004 Orange Revolution, in which hundreds of thousands of citizens took to the streets to protest electoral manipulation in the presidential elections. The subsequent opposition victory in a repeat election upended the party structure, and helped usher in a new constitution that divided power more evenly between the presidency and the parliament.

As part of a compromise following the Orange Revolution, a new constitution was adopted which greatly increased the patronage capabilities of the parliament and prime minister. As a result of the new constitution, the parliamentary majority gained more discretion over the nomination of prime ministers, more influence over the defense and security ministries and the prosecutor general, and shifted much executive staffing power away from the president and to the cabinet [@Christensen2005]. This shift reinforced the de facto competing-pyramid structure of patronage in the post-revolutionary period, by expanding and legitimizing the influence of the parliamentary majority while still retaining some patronage powers in the office of the presidency [@hale_patronal_2014]. 

The competing pyramid structure would not survive long into the Yanukovych presidency, when the Party of Regions was eager to press its advantage in patronage resources. After defeating Tymoshenko’s network in the 2010 presidential election, Yanukovych moved swiftly to encourage elite defections within the Rada, building a pro-presidential coalition that secured control of the premiership [@hale_patronal_2014;@kudelia_sources_2012]. Subsequently, Yanukovych was able to win restoration of the pre-revolution constitution favoring the president; he rapidly reconsolidated patronage networks that had become fragmented during the Orange period. 

This centralizing period would be short-lived however, as massive crowds during the 2014 Euromaidan crisis helped depose Yanukovych, leading to the seizure of the Crimean peninsula by Russia, the de facto secession of two of Ukraine’s most populous and industrialized regions, a prolonged civil war, and the effective collapse of the pro-eastern Party of Regions. The patronage networks that had helped propel the success of the Party of Regions locally and nationally began to unravel [@hale_patronal_2014]. 

In this context, to say that the consolidation of political patronage varies wildly over time is an understatement. Presidents Kuchma and Yanukovych, who drew their political support in large part from the eastern political machines, regularly sought to integrate other regions of Ukraine into the power vertical [@kudelia_sources_2012;@Pleine2016]. With the exception of applying pressure to state employees at election time, it proved challenging to export the ‘Donetsk’ model to wider Ukraine [@Zimmer2005]. At the same time, variation in consolidation tracks closely with changes in government: pro-eastern parties and presidents have generally presided over periods of growing consolidation, while the more western-oriented President Yushchenko oversaw the period of greater parliamentary influence. As a result, the influence of partisanship and patronage consolidation largely overlap empirically. 

In addition to the shifting consolidation of state-based patronage, analysts of Ukrainian politics must also contend with a complex arrangement oligarchic ‘clans,’ which deploy their resources to support favored parties. Unlike in Russia, where the major oligarchs generally do not risk challenging the Kremlin Ukraine’s oligarchic clans have been more fractious. Focusing on pro-eastern parties helps navigate this dynamic, since the pro-eastern oligarchic clan is relatively more unified than their pro-western counterparts. During Kuchma’s presidency, three regional oligarchic clans became influential in politics: the Dnipropetrovsk network, the Donetsk network, and the Kyiv network. All three were generally aligned with Kuchma until the Orange Revolution [@Pleine2016], and used their resources to influence election outcomes [@Katchanovski2008]. The revolution split the political-oligarchic patronage networks into three competing camps\textemdash one each aligned with Viktor Yushchenko, Yulia Tymoshenko, and Viktor Yanukovych\textemdash a framework that persisted until the 2010 presidential election [@Hale2010]. Even after being cast into the political wilderness nationally by the Orange crowds, Yanukovych and the Party of Regions could rely on regional patronage networks and elected local mayors to retain influence in the east [@Hale2010]. 

When Yanukovych assumed the presidency in 2010, the fractiousness of the opposition coalition of oligarchs gave way once again to a more united patronage pyramid around the presidency and the Donetsk clan. In turn, this informal patronage network collapsed as a result of the Euromaidan and subsequent civil war, generally reconstituting itself around oligarch Petro Poroshenko after his 2014 presidential victory [@Pleine2016]. Poroshenko, though his own election was largely clean, used patronage resources to good effect (especially in the East, where networks still remained largely intact) in the subsequent legislative elections [@Fedorenko2016].

In sum, Ukraine is a case where many of the conditions are in place for widespread electoral manipulation to occur. Patronage systems are embedded in society, especially in eastern regions of the country. Presidential influence over governors potentially unites the state apparatus in a network that can mobilize patronage resources to deliver votes for the incumbent. There are deep links between the state and business interests [@kudelia_sources_2012]. Nevertheless, the actual level of manipulation has varied considerably between elections and across regions, according to international monitors and experts [@Hale2010;@Herron2010;@Herron2011;@Herron2014;@Katchanovski2008]. This variation can be explained largely by shifts in patronage consolidation. Patronage is substantially less consolidated than in Russia; there is no ruling party akin to United Russia and oligarchic clans compete to influence election results. The constitution has at times fostered competition between patronage networks affiliated with the president and those affiliated with the prime minister. Local constraints are generally higher where pro-Western parties prevail than in pro-Eastern regions, but were elevated generally by the electoral reforms that were in force from the third round of the 2004 presidential election through the 2010 presidential election.



## Data and methods


```{r hist-east, echo=FALSE, fig.cap="Histogram of turnout coefficients for major pro-East parties (2002-2014)", out.width = '80%'}
ggplot(merged.data, aes(coefs.east)) + geom_histogram(binwidth = .02) + theme_bw() + labs(x = "Turnout coefficients (pro-Eastern)", y = "Count")
```

```{r hist-west, echo=FALSE, fig.cap="Histogram of turnout coefficients for major pro-West parties (2002-2014)", out.width = '80%'}
ggplot(merged.data, aes(coefs.west)) + geom_histogram(binwidth = .02) + theme_bw() + labs(x = "Turnout coefficients (pro-Western)", y = "Count")
```



### Explanatory variables

In order to test the theory, explanatory variable must capture patronage constraints and local constraints against illegal electoral manipulation. Local constraints are operationalized, following Harvey [-@harvey_manipulation_2016], using the regional legislative margin of victory between the largest parties in regional legislative elections and their largest political rival. This measure is lagged, so that it precedes the national elections that produce estimates of manipulation that serve as dependent variables. Despite Ukraine’s unitary system, elected regional legislatures do provide an opportunity for out-parties to place constraints on the governing party [@Romanova2013]. Margins of victory in regional elections are also indicative of the underlying degree of contestation, which is undergirded by civil society activism, media coverage, oligarchic influence, and other elements of local constraint. 

The Ukraine case, with its dramatic political history, provides one unambiguous case of elevated patronage consolidation during the period covered here: the 2012 legislative election, which takes place during the Yanukovych presidency but prior to the Euromaidan crisis. This is the only instance in which the political institutions of the country align with unified government under a patronage-based party machine, the Party of Regions, and without a major political crisis. Following Yanukovych’s victory in 2010, the pro-presidential Ukrainian constitution was restored, replacing the post-Orange constitution that vested greater power in the parliament. Under the pro-parliament constitution, Ukraine was more fully semi-presidential, with the president and prime minister controlling distinct patronage pyramids. This division was especially acute after the 2006 legislative election, which ushered in divided government under President Yushchenko and Prime Minister Yanukovych. The 2004 election, of course, takes place during the Orange Revolution and the attendant collapse of the patronage networks organized around President Kuchma, a collapse which had its roots in the earlier Kuchma-gate crisis. The 2012 election represents the height of patronage consolidation in Ukraine, as well, since the in 2014 the Euromaidan revolution and subsequent conflict results in the complete dissolution of the Party of Regions and a more fragmented political scene. A more detailed description of these dynamics is provided by Hale (2014). This period is captured with a dummy variable **Yanukovych**, which takes on a value of 1 for the 2012 legislative election.

The history sketched above also suggests that the two 2014 elections, which took place after Yanukovych fled the country and in the midst of the collapse of the Party of Regions as a political force, represent the nadir of patronage consolidation from the perspective of pro-eastern candidates. Since this crisis was sparked by the mass pro-Europe protests and anti-Yanukovych demonstrates at Kyiv’s Independence Square (**Maidan Nezalezhnosti**), a dummy variable labelled **Post-maidan** takes on a value of 1 for these two elections. 	

### Control variables

A set of control variables are included to help account for other possible explanations for manipulation, or for innocent voting patterns that could appear suspicious using the election forensic tools described above. One of these is an election-year variable: an indicator for \textit{presidential elections}, which may be more intense drivers of electoral manipulation [@simpser2013}. The remaining controls are socioeconomic variables at the regional level, taken from Ukrainian government statistics. Since greater education levels and economic development are thought to help deter electoral manipulation, I include a measure of \textit{higher education} as well as \textit{per capita income}. The first of these measures the number of people with post-secondary education in a region, per capita.^[Unfortunately, this variable is only available from the 2001 Ukrainian census.} The second is each region’s gross regional product divided by its population. \textit{Population size} is also included, since manipulation may be more difficult to implement in larger polities. The \textit{urban} share of the population is included, because cities may be more vulnerable to forms of manipulation like vote-buying. As \textit{government employees} may be particular susceptible to political pressure and the use of administrative resources, the number of government workers logged is used. Similarly, older Ukrainians are more likely to be dependent on the state for regular pension dispersal and other services, and more liable to administrative pressure. A measure of the number of citizens \textit{over sixty} in each region, logged, is included. For overviews of the political and socioeconomic correlates of electoral manipulation, see [@lehoucq2003;@kitschelt2007;@stokes2013brokers;@schedler2002]. 

## Results

The results of this analysis are supportive of the consolidation-constraint model, and challenge the local-information model of electoral manipulation. Table \ref{tab:turnout} reports the results of the turnout coefficient models. The local-constraint model holds that regional party margin-east should have a negative effect on deflationary manipulation against pro-eastern parties. The results in Table \ref{tab:turnout} show that this is the case during the non-Yanukovych years (Models 2) and non-Maidan years (Model 3), while Figures \ref{fig:negcoef_yanukovych} and \ref{fig:negcoef_maidan} show the interaction effects. As predicted, the interaction of regional party margin-east and Yanukovych shows that by 2012, the Party of Regions and its allies were able to reduce the incidence of deflationary manipulation against their party nearly across the board. Only in the regions of very highest levels of pro-Western support were they unable to reduce deflationary manipulation, suggesting that agents in those areas still were willing to throw in their lots with pro-western patronage networks.

Figure \ref{fig:negcoef_yanukovych} shows that the reverse is also true. During the period of post-Maidan elections, when pro-eastern patronage networks were suddenly fragmented by cascading crises, deflationary manipulation against pro-eastern candidates becomes more likely in regions with pro-eastern regional legislatures, but not in pro-western regions. This result can be interpreted as a shift by agents away from pro-eastern patronage networks and toward pro-western ones as the relative appeal of the two options shifts. Where high risks for pro-western manipulation had previously made western patronage offers less appealing than their eastern counterparts, the sudden de-consolidation of eastern networks makes the competing offer more appealing. In pro-western regions, this effect is muted since anti-eastern deflationary manipulation was already in practice in these regions due to low local risks for such agents. 


These results suggest that patronage networks and local risks both influence levels of election manipulation, but how attributable are they to principal-agent dynamics? The next set of models, using the number of possibly fraudulent precincts per region as the dependent variable, offers additional support for the theoretical framework. Since the dependent variable is based on apparently non-random clustering of precincts around specific vote-shares, it offers an indicator of agent behavior—the fulfillment of vote-share targets set by principals. Given that the overall indicator of manipulation provided by the turnout test almost universally detects deflationary rather than inflationary manipulation, I assume that the fraudulent precincts detected by the vote-share test are generally indicators of deflationary manipulation as well.

As in the prior set of models, increasing local risks for pro-western agents results in reduced evidence of falsification against pro-eastern parties, both during the Yanukovych presidency (Model 5) and after the Maidan crisis (Model 6). The marginal effects of those shifts in patronage consolidation are presented in Figures \ref{fig:roz_yanukovych} and \ref{fig:roz_maidan}. Figure \ref{fig:roz_yanukovych} shows that the effect of Yanukovych is negative at all values of regional party margin, implying that the high consolidation of the pro-eastern patronage network during that period strongly reduced the incentive for pro-western agents to engage in deflationary manipulation. This effect is smallest in pro-eastern regions, where high local constraints on pro-western agents made such manipulation difficult to achieve even during the pre-Yanukovych period. The largest shifts occur in pro-western regions, where the apparent national dominance of the Party of Regions machine removes an incentive for pro-western behavior by agents. Finally, Figure \ref{fig:roz_maidan} shows the effect of the breakdown of that nationwide pro-eastern patronage network: an increase in evidence of fraud affecting pro-eastern parties, which is especially sharp in pro-eastern regions. In pro-western regions the effect of this shift in patronage consolidation is minimal, since such regions are already more closely aligned with pro-western parties, making anti-eastern manipulation easier in the pre-crisis period. The effect is largest in pro-eastern regions where the sudden absence of a consolidated pro-eastern network makes participation in pro-western efforts much more appealing. It is important to note here, however, that this analysis excludes election results from several of the most solidly pro-eastern regions—Donetsk, Luhansk, Crimea, and Sevastopol\textemdash where the Party of Regions network might have been expected to survive in the absence of war and annexation.

Lastly, Figures \ref{fig:pp_roz_yan} and \ref{fig:pp_roz_maidan} show the predicted number of suspect precincts per region, by regional party margin—east and the variables that capture patronage consolidation, to illustrate substantive significance. To produce the predicted counts, continuous control variables are held constant at their mean values. The categorical controls presidential election and Orange Revolution were held constant at zero. Figure \ref{fig:pp_roz_yan} shows the near complete absence of suspicious precincts in heavily pro-eastern regions, with little change when Yanukovych is president. By contrast, the predicted number of suspicious precincts in a pro-western region is between twenty and forty in the non-Yanukovych years, but falls to eastern levels during his presidency.  Figure \ref{fig:pp_roz_maidan} shows that the Maidan crisis led to a large jump in the predicted number of suspicious precincts across the board; while the absolute increase is largest in pro-western regions (from about 10 to about 50), the relative increase is larger in pro-eastern regions (from near zero to about 20).

A quick back-of-the-envelope example demonstrates the number of votes at stake is substantively significant, though not of overweening importance. In the 2007 Ukrainian legislative election, 450 seats in the Verkhovna Rada were allocated by proportional representation to parties winning more than three percent of the national vote-share. Collectively these parties won approximately twenty-one million votes, meaning that each seat in the Rada was worth about 46,000 votes. Given that each precinct includes about 1,000 registered voters on average, the presence of thirty suspicious precincts in a region might reflect deflationary manipulation that deprived a party of a few thousand votes. Magnified across Ukraine’s regions, this kind of manipulation could conceivably adjust the balance of power in the Rada by a handful of seats. And of course, falsification of this kind represents only one technique available to parties.

## Discussion and conclusion

Despite the difficulties of the case, Ukraine presents an important opportunity to test the implications of the model, especially in comparison with Russia. Russia shares many institutional and historical features with Ukraine (particularly in eastern parts of the country) as a result of a shared imperial and Soviet experience. However, patterns of patronage consolidation and local constraints vary considerably between the two states. Patronage consolidation has consistently been much lower in Ukraine than in Russia during the period of analysis. A single party of power largely dominates in Russia, while the party system in Ukraine is much more fragmented and fluid. Oligarchic interests were largely tamed during the early years of the first Putin administration; rival oligarchic clans continue to clash behind the scenes in Ukraine. The period in which Ukraine’s parliamentary-presidential constitution was in force helped establish a two-pyramid structure of patronage, in Hale’s (2014) terminology, while Russia continues to operate within a more consolidated vertical of power.

As a result, we should expect to see different patterns of manipulation occur in the two cases. First, while the direct comparison is not statistically tested here, widespread falsification does appear to be considerably less common in Ukraine than in Russia. In the Russian case, suspicious results were captured by the digits test in 24 percent of region-years, whereas in Ukraine the same figure is approximately five percent—roughly what would be expected due to random chance alone.^[Note that the digits test results differ from those of the vote-share test presented here; the mean number of suspicious precincts per region using the vote-share test is approximately 10 percent.] Furthermore, while inflationary manipulation is common in Russia, to the benefit of the ruling party, suspicious turnout coefficients tend to be negative in Ukraine. This apparent emphasis on deflationary techniques may be due to the lower levels of patronage consolidation in the country. Since widespread voter mobilization and voter pressure campaigns can only be deployed by parties that control substantial patronage resources, cheaper deflationary tactics may easier to pull off for parties operating in a more fragmented patronage environment. This is a question that may be usefully addressed in further work.

However, the main benefit of the case in the present study is derived from the rapid fluctuation in patronage consolidation over time. Unlike the Mexican case (where patronage networks are deconsolidating during the period of study) or the Russian case (where they remain highly consolidated), Ukraine shifts from a fragmented patronage system after the Orange Revolution, to one with considerable consolidation around the Party of Regions under President Yanukovych, and subsequently back to a fragmented system after the 2014 crisis. The results shown here indicate that the consolidation and constraint model is better suited to explaining electoral manipulation in this kind of setting than is the local-information model; that is, while local conditions certainly matter for explaining electoral manipulation, those conditions are also informed by the overall distribution of patronage resources backing one party or another.

In particular, the election-forensic analysis conducted here shows that overall deflationary manipulation targeting the Party of Regions became less severe during the Yanukovych presidency, but only in regions where constraints on pro-western agents were high. It also found a corresponding increase in deflationary manipulation in pro-eastern regions when the Party of Regions patronage network collapsed. Analysis of the models using the vote-share test results as a dependent variable, while only measuring one type of manipulation, provide further support for the principal-agent model based on patronage and risk. In those models, higher patronage consolidation around the Party of Regions led to a steep reduction in the number of suspicious precincts in pro-western regions; the subsequent breakdown resulted in increased fraud across the board (with especially steep increases in the east).

In sum, election-forensic analysis shows that overall electoral manipulation and specific signatures of fraud both respond to shifts in local risks and national patronage consolidation in ways predicted by the model. Where the Mexican case allowed for an investigation of the effects of shifting local competition while holding patronage consolidation constant, the Ukrainian case allows for analysis of shifting levels of patronage consolidation. Both cases demonstrate that the two factors can move independently, and interactive effect the level of election manipulation that parties can achieve. 


## Results

Trying again with a clean setup

Party margin is regional legislative vote-share in prior election

There may be something here: When there is a pro-east president, pro-east regions are free to run up the score using dispersed methods. Infusion of state resources too. But we see that dev_1_9_east is higher in east regions regardless of president. This seems to go against expectations. Total deviance (g_statistic_east) doesn't change across regions or with presidents. G-statistic appears strongly correlated with over-sixty percent, which could have something to do with falsification of older people's ballots?

```{r}
merged.data <- merged.data %>% mutate(coef_ratio = coefs.east / (coefs.east + coefs.west))
merged.data <- merged.data %>% mutate(over.sixty_pct = over.sixty / (population*1000))

model.vs_margin_coef <- lm(east.party.margin2 ~ coefs.east + coefs.west  + pres.approval + party.margin + gov.employ + over.sixty_pct  + urban.pct + population + ukr.lang.pct, data = merged.data)
summary(model.vs_margin_coef)

model.vs_margin_coef2 <- lm(east.party.margin2 ~ coefs.east + pres.approval + party.margin + gov.employ + over.sixty_pct  + urban.pct + population + ukr.lang.pct, data = merged.data)
summary(model.vs_margin_coef2)

model.coef_margin_pres <- lm(coefs.east ~ party.margin + pres.approval + party.margin*east.president + gov.employ + over.sixty_pct + average.wage + population, data = merged.data)
summary(model.coef_margin_pres)
sjPlot::plot_model(model.coef_margin_pres, type = "int")

model.coef_rcl <- lm(coefs.east ~ rcl.proeast + pres.approval + east.president  + gov.employ + over.sixty_pct + average.wage + population + ukr.lang.pct, data = merged.data)
summary(model.coef_rcl)

model.coef_region <- lm(coefs.east ~ prewar.region + pres.approval + east.president +  gov.employ + over.sixty_pct + average.wage + population + ukr.lang.pct, data = merged.data)
summary(model.coef_region)

model.coef_elex <- lm(coefs.east ~ electionid + gov.employ + over.sixty_pct + average.wage + population + ukr.lang.pct, data = merged.data)
summary(model.coef_elex)



model.coef_elex_west <- lm(coefs.west ~ electionid + gov.employ + over.sixty_pct + average.wage + population + ukr.lang.pct, data = merged.data)
summary(model.coef_elex_west)


model.coef_olig <- lm(coefs.east ~ east_oligcount*urban.pct + pres.approval + east.president +  over.sixty_pct + average.wage + population + ukr.lang.pct, data = merged.data)
summary(model.coef_olig)
sjPlot::plot_model(model.coef_olig, type = "pred", terms = c("urban.pct", "east_oligcount [0, 2]"))


```

```{r}

model.vs_gs <- lm(east.party.margin2 ~ g_statistic_east + pres.approval + party.margin + gov.employ + over.sixty_pct  + population + urban.pct + ukr.lang.pct +  sample_size_east, data = merged.data)
summary(model.vs_gs)

model.gs_margin_int <- lm(g_statistic_east ~ party.margin + pres.approval + party.margin*east.president + gov.employ + over.sixty_pct + population + average.wage + ukr.lang.pct + sample_size_east, data = merged.data)
summary(model.gs_margin_int)
sjPlot::plot_model(model.gs_margin_int, type = "int")

model.gs_margin_int2 <- lm(dev_1_9_east ~ party.margin + pres.approval + party.margin*east.president + gov.employ + over.sixty_pct + population + average.wage + ukr.lang.pct + sample_size_east, data = merged.data)
summary(model.gs_margin_int2)
sjPlot::plot_model(model.gs_margin_int2, type = "int")


model.gs_margin <- lm(g_statistic_east ~ party.margin + pres.approval + party.margin + gov.employ + over.sixty_pct + population + average.wage + ukr.lang.pct + sample_size_east, data = merged.data)
summary(model.gs_margin)


model.gs_rcl <- lm(g_statistic_east ~ rcl.proeast + pres.approval + east.president  + gov.employ + over.sixty_pct + population + average.wage + ukr.lang.pct + sample_size_east, data = merged.data)
summary(model.gs_rcl)

model.gs_region <- lm(g_statistic_east ~ prewar.west + pres.approval + east.president +  gov.employ + over.sixty_pct + population + average.wage + ukr.lang.pct + sample_size_east, data = merged.data)
summary(model.gs_region)

model.gs_elex <- lm(g_statistic_east ~ electionid + gov.employ + over.sixty_pct + population + average.wage + ukr.lang.pct + sample_size_east, data = merged.data)
summary(model.gs_elex)

model.gs_elex_east <- lm(g_statistic_east ~ electionid + gov.employ + over.sixty_pct + average.wage + ukr.lang.pct + sample_size_east, data = merged.data)
summary(model.gs_elex_east)

model.gs_elex_west <- lm(g_statistic_west ~ electionid + gov.employ + over.sixty_pct + average.wage + ukr.lang.pct + sample_size_west, data = merged.data)
summary(model.gs_elex_west)

model.gs_olig <- lm(g_statistic_east ~ east_oligcount*urban.pct + pres.approval + east.president + gov.employ + over.sixty_pct + average.wage + ukr.lang.pct + sample_size_east, data = merged.data)
summary(model.gs_olig)
```


Table \@ref(tab:vsmodels) shows the expected relationship between the pro-Eastern turnout coefficients, and pro-Eastern vote-share. Model 1, which does not control for pro-Western parties' turnout coefficients, shows a positive and statistically significant effect. The effect remains more or less constant once pro-Western turnout coefficient is included as a control in Model 2. In either case, the effect size is substantial: a one standard deviance increase in pro-Eastern turnout coefficient is associated with a 6-point increase in the pro-Eastern party's margin of victory. In an otherwise tied election, in other words, increasing the pro-Eastern party's turnout coefficient by one standard deviation would change the outcome of the election to 53 - 47 in their favor. 

Model 3 shows that there is no significant effect for deviance from the uniform distribution of last-digits for pro-East parties. This suggestions that outright falsification of the type captured by this measure is rare in Ukraine across the elections covered in the dataset, at least for major pro-Eastern parties. There is one caveat to this conclusion; following best practice [@beber2012], I excluded precincts from the deviance calculations where fewer than 100 votes were counted in that party's favor. A high proportion of precincts with double-digit vote totals will increase the deviance from the uniform distribution naturally, even if the votes are fair. This proportion is extremely high in some of Ukraine's more western regions. For instance, in the 2002 election in Ivano-Frankivska oblast', all but 24 out of 958 precincts are excluded for having fewer than 100 votes for the pro-Eastern party. In 2014, all 914 precincts in Ivano-Frankivska are dropped, as are all precincts in L'vivska, Ternopilska, Vinnitska, Ternopilska, and Cherkaska oblasts. This highlights the difficulty of conducting this sort of analysis conclusively in Ukraine's highly polarized and frequently crisis-driven politics under the period of study.   

```{r vsmodels}
models <- list()
models[['Model 1']] <- model.vs_margin_coef2
models[['Model 2']] <- model.vs_margin_coef
models[['Model 3']] <- model.vs_gs


library(kableExtra)

reg_table_vs <- modelsummary(models, output = "kableExtra",  stars = T, title = 'Table 1: OLS models of pro-Eastern party vote-share', coef_map = c("(Intercept)" = "(Intercept)",
             "coefs.east" = "Pro-Eastern turnout coefficient",
             "coefs.west" = "Pro-Western turnout coefficient",
             "g_statistic_east" = "Pro-East party last-digit deviance",
             "pres.approval" = "Presidential approval rating",
             "party.margin" = "Pro-Eastern margin in regional election",
             "gov.employ" = "Gov. employees",
             "over.sixty_pct" = "Over-sixty pop. share",
             "average.wage" = "Average wage",
             "population" = "Population",
             "sample_size_east" = "N. precincts > 99 pro-East votes"))

reg_table_vs





```

Table \@ref(tab:manip-models) shows the results of models that take pro-Eastern turnout coefficients or last-digit deviance for pro-Eastern parties as dependent variables.


```{r manip-models}
models <- list() 
models[['Model 1']] <- model.coef_margin_pres
models[['Model 2']] <- model.coef_rcl
models[['Model 3']] <- model.coef_region
models[['Model 4']] <- model.gs_margin_int
models[['Model 5']] <- model.gs_rcl
models[['Model 6']] <- model.gs_region


table1 <- modelsummary(
  models,
  output = "kableExtra",
  title = "Regression Results",
#  coef_map = c("x1" = "Variable 1",
#               "x2" = "Variable 2", 
#               "x3" = "Variable 3",
#               "(Intercept)" = "Constant"),
#  gof_map = c("nobs", "r.squared", "adj.r.squared"),
  stars = c('*' = .1, '**' = .05, '***' = .01)
) %>%
  # Add model numbers as column names
  #`colnames<-`(c("", "(1)", "(2)", "(3)", "(4)", "(5)", "(6)")) %>%
  # Add grouped headers for dependent variables
  add_header_above(c(" " = 1, "Pro-East turnout coef." = 3, "Last-digit deviance" = 3))

table1
```

Can looks at Rozenas scores as an appendix / RC type deal. Bottom line is that we see huge effects in places where Eastern parties get no votes (driven by zeroes)


Lastly, Table \@ref(tab:yearmodels) shows how turnoutcoefficients and last-digit deviance for the major pro-Eastern party in each election year varies over time. Since this model includes election-year fixed effects, other annual control variables (e.g. *presidential election* or *pro-Eastern Rada consolidation*) must be excluded. As expected, the results show that evidence of overall election manipulation varies over time, while evidence of falsification remains essentially constant.

```{r yearmodels}
models <- list()
models[['Model 1']] <- model.coef_elex
models[['Model 2']] <- model.gs_elex


library(kableExtra)

reg_table_year <- modelsummary(models, output = "kableExtra",  stars = T, title = 'Table 3: OLS models of election manipulation by year', coef_map = c("(Intercept)" = "(Intercept)",
             "electionid2004_1_0" = "2004 (R3)",
             "electionid2006_0_0" = "2006",
             "electionid2007_0_0" = "2007",
             "electionid2010_1_0" = "2010 (R1)",
            "electionid2010_1_1" =  "2010 (R2)",
            "electionid2012_0_0" = "2012",
            "electionid2014_0_0" = "2014 (Leg.)",
            "electionid2014_1_0" =  "2014 (Pres.)",
             "gov.employ" = "Gov. employees",
             "over.sixty_pct" = "Over-sixty pop. share",
             "average.wage" = "Average wage",
             "population" = "Population",
             "sample_size_east" = "N. precincts > 99 pro-East votes"
             ))

reg_table_year



```



### There are dragons below here:
Voteshare patterns
```{r error=FALSE, message=FALSE, warning=FALSE}
merged.data <- merged.data %>% mutate(dev_ratio = deviance_region_east / (deviance_region_east + deviance_region_west))
merged.data <- merged.data %>% mutate(dev_diff = deviance_region_east - deviance_region_west)
merged.data <- merged.data %>% mutate(dev_sum = deviance_region_east + deviance_region_west)

merged.data <- merged.data %>% mutate(rozpct_diff_east = east.roz.percent - west.roz.percent)
merged.data <- merged.data %>% mutate(log_dev_east = log(deviance_region_east))
merged.data <- merged.data %>% mutate(log_dev_west = log(deviance_region_west))
merged.data <- merged.data %>% mutate(log_dev_diff = log_dev_east - log_dev_west)



model.vs <- lm(east.party.voteshare ~ prewar.region*rcl.proeast, data = merged.data)
sjPlot::plot_model(model.vs, type = "int")

model.vs <- lm(east.party.voteshare ~ coef_ratio, data = merged.data)
summary(model.vs)

model.vs <- lm(east.party.voteshare ~ g_statistic_east + g_statistic_west + n.precincts, data = merged.data)
summary(model.vs)

model.vs <- lm(west.party.voteshare ~ prewar.region*log_dev_diff + n.precincts + presidential, data = merged.data %>% filter(log_dev_east != -Inf))
summary(model.vs)
sjPlot::plot_model(model.vs, type = "pred", terms = c("prewar.region", "log_dev_diff"))
sjPlot::plot_model(model.vs, type = "pred", terms = c("log_dev_diff", "prewar.region"))


model.vs <- lm(east.party.voteshare ~ prewar.region*rozpct_diff_east + presidential, data = merged.data)
summary(model.vs)
sjPlot::plot_model(model.vs, type = "pred", terms = c("prewar.region", "rozpct_diff_east"))


### Overall coef ratio

model.coef <- lm(coef_ratio ~ prewar.region*rcl.proeast +   logpop + income.percap  
                            + highered 
                            + presidential, data = merged.data)
summary(model.coef)
sjPlot::plot_model(model.coef, type = "int")



### Rozenas
model.rozdiff <- lm(east.roz.percent ~ prewar.region*rcl.proeast   +  logpop + income.percap +
                           n.precincts + highered 
                            + presidential, data = merged.data)
summary(model.rozdiff)
sjPlot::plot_model(model.rozdiff, type = "pred", terms = c("prewar.region [Central, West, East]", "rcl.proeast"))
sjPlot::plot_model(model.rozdiff, type = "pred", terms = c("rcl.proeast", "prewar.region [Central, West, East]"))

model.count.roz <- MASS::glm.nb(w.roz.count.total ~ prewar.region*rcl.proeast   +  logpop + income.percap +
                           n.precincts + highered 
                            + presidential, data = merged.data )
summary(model.count.roz)
sjPlot::plot_model(model.count.roz, type = "pred", terms = c("prewar.region [Central, West, East]", "rcl.proeast"))
sjPlot::plot_model(model.count.roz, type = "pred", terms = c("rcl.proeast", "prewar.region [Central, West, East]"))



model.vs.manip <- MASS::glm.nb(e.roz.count.total ~ prewar.region*rcl.proeast   +  logpop + income.percap  + n.precincts
                            + highered 
                            + presidential, data = merged.data )
summary(model.vs.manip)
sjPlot::plot_model(model.vs.manip, type = "pred", terms = c("prewar.region", "rcl.proeast"))
sjPlot::plot_model(model.vs.manip, type = "pred", terms = c("rcl.proeast", "prewar.region [Central, West, East]"))
```

Current thoughts:

1) There is not a lot of extreme falsification in Ukraine. Rozenas spike score at the 90th percentile in Ukraine is 1.2%. In Russia it's 2.5%. Deviance is pretty similar to Russia though...at least for east parties.  Non-extreme distributions are pretty similar.
2) What do we know?
  A) Increasing pro-east Rada consolidation -> increased east-party vote-share in the Central regions, and a little in the East, but not in the West.
  B) Increasing pro-east overall turnout coefficient ratio -> eastern vote-share. (Ratio is needed because both parties can manipulate)
  C) Rozenas percent (i.e. spikes) is associated with increased east party vote-share in the East and the Center, but not the West.
  D) Increasing pro-east Rada consolidation leads to increased overall pro-East manipulation in the West and Center, but not the East.
  E) There is high east Rozenas score in the pre-war West region when the Rada is more pro-Western, declining as PoR rises. No change across other regions.
3) Putting this all together:


Here there be dragons:

```{r error=FALSE, warning=FALSE, message=FALSE}
##East parties; deviance score

merged.data <- merged.data %>% mutate(post2010 = as.factor(ifelse(year > 2010, 1, 0)))

model.controls.east.r <- lm(g_statistic_east ~ #average.wage + 
                              over.sixty + gov.employ #+ edu.employ + health.employ
                            +urban.pct + logpop + income.percap #+ ukr.lang.pct 
                            + highered 
                            + presidential +  n.precincts#+ crisis.year
                            , data=merged.data)
summary(model.controls.east.r)  

model.local.east.r <- lm(g_statistic_east ~ #average.wage + 
                           over.sixty + gov.employ #+ edu.employ + health.employ
                         +urban.pct + logpop + income.percap #+ ukr.lang.pct 
                         + highered 
                         + presidential + east.party.margin2  + n.precincts #+ crisis.year
                         , data=merged.data)
summary(model.local.east.r)  
#sjPlot::plot_model(model.local.east.r, type = "pred", terms = c("post2010", "party.margin"))

model.regions.east.r <- lm(g_statistic_east ~ #average.wage + 
                             over.sixty + gov.employ #+ edu.employ + health.employ
                           +urban.pct + logpop + income.percap #+ ukr.lang.pct 
                           + highered 
                           + presidential + n.precincts + factor(region.group) #+ crisis.year
                           , data=merged.data)
summary(model.regions.east.r)  

model.national.east.r <- lm(g_statistic_east ~ #average.wage + 
                              over.sixty + gov.employ #+ edu.employ + health.employ
                            +urban.pct + logpop + income.percap #+ ukr.lang.pct 
                            + highered 
                            + presidential + n.precincts + pres.constitution #+ crisis.year
                            , data=merged.data)
summary(model.national.east.r) 

model.years.east.r <- lm(g_statistic_east ~ #average.wage + 
                           over.sixty + gov.employ #+ edu.employ + health.employ
                         +urban.pct + logpop + income.percap #+ ukr.lang.pct 
                         + highered 
                         #+ presidential 
                         + factor(electionid) + n.precincts #+ crisis.year
                         , data=merged.data)
summary(model.years.east.r) #2014 parl positive

model.interactionyears.east.r <- lm(g_statistic_east ~ #average.wage + 
                                      over.sixty + gov.employ #+ edu.employ + health.employ
                                    +urban.pct  + income.percap #+ ukr.lang.pct 
                                    + highered 
                                    #+ presidential 
                                    + east.party.margin2 + electionid + 
                                      east.party.margin2*electionid + n.precincts #+ crisis.year
                                    , data=merged.data)
summary(model.interactionyears.east.r)
sjPlot::plot_model(model.interactionyears.east.r, type = "pred", terms = c("electionid", "east.party.margin2 [-.51, .32]"))

model.interaction.east.r <- lm(g_statistic_east ~ #average.wage + 
                                 over.sixty + gov.employ #+ edu.employ + health.employ
                               +urban.pct  + income.percap #+ ukr.lang.pct 
                               +  
                               + presidential +  east.party.margin2 +  president.name + n.precincts +
                                 prewar.west*president.name #+ crisis.year
                               , data=merged.data)
summary(model.interaction.east.r) #

sjPlot::plot_model(model.interaction.east.r, type = "pred", terms = c("president.name", "prewar.west"))
```

```{r}
model.interaction.east.all <- lm(coefs.east ~ #average.wage + 
                                 over.sixty + gov.employ #+ edu.employ + health.employ
                               +urban.pct  + income.percap #+ ukr.lang.pct 
                               
                               +  east.party.margin2 * electionid + n.precincts +
                                 electionid #+ crisis.year
                               , data=merged.data)
summary(model.interaction.east.all) #
sjPlot::plot_model(model.interaction.east.all, type = "pred", terms = c("electionid", "east.party.margin2 [-.51, .32]"))

model.interaction.east.all <- lm(g_statistic_east ~ #average.wage + 
                                 over.sixty + gov.employ #+ edu.employ + health.employ
                               +urban.pct  + income.percap #+ ukr.lang.pct 
                               
                               +  east.party.margin2* electionid + n.precincts +
                                 electionid #+ crisis.year
                               , data=merged.data)
summary(model.interaction.east.all) #
sjPlot::plot_model(model.interaction.east.all, type = "pred", terms = c("electionid", "east.party.margin2 [-.51, .32]"))



model.interaction.east.all.rcl <- lm(coefs.east ~ #average.wage + 
                                 over.sixty + gov.employ #+ edu.employ + health.employ
                               +urban.pct  + income.percap #+ ukr.lang.pct 
                               +  pres.approve.margin +
                               + presidential +  east.party.margin2 *rcl.proeast + n.precincts +
                                 rcl.proeast #+ crisis.year
                               , data=merged.data)
summary(model.interaction.east.all.rcl) #




model.interaction.east.all.rcl <- lm(g_statistic_east ~ #average.wage + 
                                 over.sixty + gov.employ #+ edu.employ + health.employ
                               +urban.pct  + income.percap #+ ukr.lang.pct 
                               +  pres.approve.margin +
                               + presidential +  east.party.margin2 *  rcl.proeast + n.precincts +
                                 rcl.proeast #+ crisis.year
                               , data=merged.data)
summary(model.interaction.east.all.rcl) #


model.interaction.east.dev.rcl <- lm(dev_1_9_east ~ #average.wage + 
                                 over.sixty + gov.employ #+ edu.employ + health.employ
                               +urban.pct  + income.percap #+ ukr.lang.pct 
                               +  pres.approve.margin
                               + presidential +  east.party.margin2*rcl.proeast + n.precincts  #+ crisis.year
                               , data=merged.data)
summary(model.interaction.east.dev.rcl) #
sjPlot::plot_model(model.interaction.east.dev.rcl, type = "pred", terms = c("rcl.proeast", "east.party.margin2 [-.51, .32]"))


### post-2010

model.interaction.east.all.2010 <- lm(coefs.east ~ #average.wage + 
                                 over.sixty + gov.employ #+ edu.employ + health.employ
                               +urban.pct  + income.percap #+ ukr.lang.pct 
                               +  pres.approve.margin
                               + presidential +  east.party.margin2  + n.precincts +
                                 east.party.margin2*post2010 #+ crisis.year
                               , data=merged.data)
summary(model.interaction.east.all.2010) #
sjPlot::plot_model(model.interaction.east.all.2010, type = "pred", terms = c("east.party.margin2", "post2010"))

model.interaction.east.dev.2010 <- lm(dev_1_9_east ~ #average.wage + 
                                 over.sixty + gov.employ #+ edu.employ + health.employ
                               +urban.pct  + income.percap #+ ukr.lang.pct 
                               +  
                               + presidential +  east.party.margin2  + n.precincts +
                                 east.party.margin2*post2010 #+ crisis.year
                               , data=merged.data)
summary(model.interaction.east.dev.2010) #
sjPlot::plot_model(model.interaction.east.dev.2010, type = "pred", terms = c("east.party.margin2", "post2010"))
```



```{r error=FALSE, message=FALSE, warning=FALSE}
##West parties; deviance score

model.controls.west.r <- lm(g_statistic_west ~ #average.wage + 
                              over.sixty + gov.employ #+ edu.employ + health.employ
                            +urban.pct + logpop + income.percap #+ ukr.lang.pct 
                            + highered 
                            + presidential + n.precincts#+ crisis.year
                            , data=merged.data)
summary(model.controls.west.r)  

model.local.west.r <- lm(g_statistic_west ~ #average.wage + 
                           over.sixty + gov.employ #+ edu.employ + health.employ
                         +urban.pct + logpop + income.percap #+ ukr.lang.pct 
                         + highered 
                         + presidential + party.margin*post2010 + n.precincts #+ crisis.year
                         , data=merged.data)
summary(model.local.west.r)  



model.regions.west.r <- lm(g_statistic_west ~ #average.wage + 
                             over.sixty + gov.employ #+ edu.employ + health.employ
                           +urban.pct + logpop + income.percap #+ ukr.lang.pct 
                           + highered 
                           + presidential + n.precincts + factor(region.group) #+ crisis.year
                           , data=merged.data)
summary(model.regions.west.r)  

model.national.west.r <- lm(g_statistic_west ~ #average.wage + 
                              over.sixty + gov.employ #+ edu.employ + health.employ
                            +urban.pct + logpop + income.percap #+ ukr.lang.pct 
                            + highered 
                            + presidential + n.precincts + pres.constitution #+ crisis.year
                            , data=merged.data)
summary(model.national.west.r) 

model.years.west.r <- lm(g_statistic_west ~ #average.wage + 
                           over.sixty + gov.employ #+ edu.employ + health.employ
                         +urban.pct + logpop + income.percap #+ ukr.lang.pct 
                         + highered 
                         #+ presidential 
                         + factor(electionid) + n.precincts #+ crisis.year
                         , data=merged.data)
summary(model.years.west.r) #2014 parl positive

model.interactionyears.west.r <- lm(g_statistic_west ~ #average.wage + 
                                      over.sixty + gov.employ #+ edu.employ + health.employ
                                    +urban.pct + logpop + income.percap #+ ukr.lang.pct 
                                    + highered 
                                    #+ presidential 
                                    + party.margin + factor(electionid) + 
                                      party.margin*factor(electionid) + n.precincts #+ crisis.year
                                    , data=merged.data)
summary(model.interactionyears.west.r)

model.interaction.west.r <- lm(g_statistic_west ~ #average.wage + 
                                 over.sixty + gov.employ #+ edu.employ + health.employ
                               +urban.pct + logpop + income.percap #+ ukr.lang.pct 
                               +  
                               + presidential +  party.margin +  president.name + n.precincts +
                                 party.margin*president.name #+ crisis.year
                               , data=merged.data)
summary(model.interaction.west.r) #

sjPlot::plot_model(model.interaction.west.r, type = "pred", terms = c("party.margin", "president.name"))
```
### Deviance by p-value

```{r}
merged.data <- merged.data %>% mutate(deviance_binary_east = ifelse(p_value_east <= .05, 1, 0))

merged.data <- merged.data %>% mutate(deviance_binary_west = ifelse(p_value_west <= .05, 1, 0))

total_dev_east_logit <- glm(deviance_binary_east ~ #average.wage + 
                              over.sixty + gov.employ #+ edu.employ + health.employ
                            +urban.pct + logpop + income.percap #+ ukr.lang.pct 
                            + highered 
                            + presidential + east.president*post2010*prewar.west + n.precincts#+ crisis.year
                            , data=merged.data, family = binomial(link = "logit"))
sjPlot::plot_model(total_dev_east_logit, type = "pred", terms = c("prewar.west", "east.president", "post2010"))

total_dev_west_logit <- glm(deviance_binary_west ~ #average.wage + 
                              over.sixty + gov.employ #+ edu.employ + health.employ
                            +urban.pct + logpop + income.percap #+ ukr.lang.pct 
                            + highered 
                            + presidential + east.president*post2010*prewar.west + n.precincts#+ crisis.year
                            , data=merged.data, family = binomial(link = "logit"))
sjPlot::plot_model(total_dev_west_logit, type = "pred", terms = c("prewar.west", "east.president", "post2010"))
```


### Spikes

```{r error=FALSE, message=FALSE, warning=FALSE}
ukrdata$scale.west.rozenas <- scale(ukrdata$west.rozenas)
ukrdata$scale.east.rozenas <- scale(ukrdata$east.rozenas)


model.controls.west.r <- lm(west.roz.percent ~ #average.wage + 
                             over.sixty + gov.employ #+ edu.employ + health.employ
                           +urban.pct + logpop + income.percap #+ ukr.lang.pct 
                           + highered 
                           + presidential #+ crisis.year
                           , data=ukrdata)
summary(model.controls.west.r)  

model.local.west.r <- lm(east.roz.percent ~ #average.wage + 
                              over.sixty + gov.employ #+ edu.employ + health.employ
                            +urban.pct + logpop + income.percap #+ ukr.lang.pct 
                            + highered 
                            + presidential + party.margin*east.president*post2010 #+ crisis.year
                            , data=merged.data)
summary(model.local.west.r)  #party margin negative at .1
sjPlot::plot_model(model.local.west.r, type = "pred", terms = c("party.margin", "east.president", "post2010"))

model.regions.west.r <- lm(west.roz.percent ~ #average.wage + 
                           over.sixty + gov.employ #+ edu.employ + health.employ
                         +urban.pct + logpop + income.percap #+ ukr.lang.pct 
                         + highered 
                         + presidential + factor(region.group) #+ crisis.year
                         , data=ukrdata)
summary(model.regions.west.r)  

model.national.west.r <- lm(west.roz.percent ~ #average.wage + 
                           over.sixty + gov.employ #+ edu.employ + health.employ
                         +urban.pct + logpop + income.percap #+ ukr.lang.pct 
                         + highered 
                         + presidential + pres.constitution + yanukovych #+ crisis.year
                         , data=ukrdata)
summary(model.national.west.r) 

model.years.west.r <- lm(scale.west.rozenas ~ #average.wage + 
                              over.sixty + gov.employ #+ edu.employ + health.employ
                            +urban.pct + logpop + income.percap #+ ukr.lang.pct 
                            + highered 
                            + presidential + factor(electionid) #+ crisis.year
                            , data=ukrdata)
summary(model.years.west.r) 

model.interactionyears.west.r <- lm(scale.west.rozenas ~ #average.wage + 
                                 over.sixty + gov.employ #+ edu.employ + health.employ
                               +urban.pct + logpop + income.percap #+ ukr.lang.pct 
                               + highered 
                               + presidential + party.margin + factor(electionid) + 
                                 party.margin*factor(electionid) #+ crisis.year
                               , data=ukrdata)
summary(model.interactionyears.west.r)

model.interaction.west.r <- lm(west.roz.percent ~ #average.wage + 
                              over.sixty + gov.employ #+ edu.employ + health.employ
                            +urban.pct + logpop + income.percap #+ ukr.lang.pct 
                            + highered 
                            + presidential + party.margin  + yanukovych + 
                              party.margin*yanukovych #+ crisis.year
                            , data=ukrdata)
summary(model.interaction.west.r) #margin is negative sig, interaction is positive sig

interplot::interplot(model.interaction.west.r, var1="party.margin", var2="yanukovych") #No real effect
interplot::interplot(model.interaction.west.r, var2="party.margin", var1="yanukovych") #No significance either way, as expected


##East parties

model.controls.east.r <- lm(east.roz.percent ~ #average.wage + 
                              over.sixty + gov.employ #+ edu.employ + health.employ
                            +urban.pct + logpop + income.percap #+ ukr.lang.pct 
                            + highered 
                            + presidential #+ crisis.year
                            , data=ukrdata)
summary(model.controls.east.r)  

model.local.east.r <- lm(east.roz.percent ~ #average.wage + 
                           over.sixty + gov.employ #+ edu.employ + health.employ
                         +urban.pct + logpop + income.percap #+ ukr.lang.pct 
                         + highered 
                         + presidential + party.margin #+ crisis.year
                         , data=ukrdata)
summary(model.local.east.r)  

model.regions.east.r <- lm(east.roz.percent ~ #average.wage + 
                             over.sixty + gov.employ #+ edu.employ + health.employ
                           +urban.pct + logpop + income.percap #+ ukr.lang.pct 
                           + highered 
                           + presidential + factor(region.group) #+ crisis.year
                           , data=ukrdata)
summary(model.regions.east.r)  

model.national.east.r <- lm(east.roz.percent ~ #average.wage + 
                              over.sixty + gov.employ #+ edu.employ + health.employ
                            +urban.pct + logpop + income.percap #+ ukr.lang.pct 
                            + highered 
                            + presidential + pres.constitution #+ crisis.year
                            , data=ukrdata)
summary(model.national.east.r) 

model.years.east.r <- lm(east.roz.percent ~ #average.wage + 
                           over.sixty + gov.employ #+ edu.employ + health.employ
                         +urban.pct + logpop + income.percap #+ ukr.lang.pct 
                         + highered 
                         #+ presidential 
                         + factor(electionid) #+ crisis.year
                         , data=ukrdata)
summary(model.years.east.r) #2014 parl positive

model.interactionyears.east.r <- lm(scale.east.rozenas ~ #average.wage + 
                                      over.sixty + gov.employ #+ edu.employ + health.employ
                                    +urban.pct + logpop + income.percap #+ ukr.lang.pct 
                                    + highered 
                                    #+ presidential 
                                    + party.margin + electionid + 
                                      party.margin*electionid #+ crisis.year
                                    , data=ukrdata)
summary(model.interactionyears.east.r)
sjPlot::plot_model(model.interactionyears.east.r, type = "pred", terms = c("electionid", "party.margin [-.23, .46]"))

model.interaction.east.r <- lm(scale.east.rozenas ~ #average.wage + 
                                 over.sixty + gov.employ #+ edu.employ + health.employ
                               +urban.pct + logpop + income.percap #+ ukr.lang.pct 
                               + highered 
                               + presidential  + party.margin + pres.constitution + yanukovych + 
                                 party.margin*yanukovych #+ crisis.year
                               , data=ukrdata)
summary(model.interaction.east.r) #
sjPlot::plot_model(model.interaction.east.r, type = "pred", terms = c("yanukovych", "party.margin [-.23, .46]"))

interplot::interplot(model.interaction.east.r, var1="party.margin", var2="yanukovych") #Sig here: higher pro-east margin -> less fraud during pres const
interplot::interplot(model.interaction.east.r, var2="party.margin", var1="yanukovych") #Pres const has negative effect on east-party fraud in pro-east regions, pos effect in pro-west regions

model.interaction.east.r2 <- lm(scale.east.rozenas ~ #average.wage + 
                                 over.sixty + gov.employ #+ edu.employ + health.employ
                               +urban.pct + logpop + income.percap #+ ukr.lang.pct 
                               + highered 
                               + presidential + orangerev + party.margin + rcl.proeast + pres.constitution+
                                 party.margin*rcl.proeast #+ crisis.year
                               , data=ukrdata)
summary(model.interaction.east.r2) #

sjPlot::plot_model(model.interaction.east.r2, type = "pred", terms = c("party.margin [-.23, .46]"))

```
### Rozenas count models

```{r}
###Rozenas score models
###Poisson count models
###
library(MASS)


model.controls.west.p <- glm.nb(w.roz.count.total ~ #average.wage + 
                              over.sixty + loggov #+ edu.employ + health.employ
                            +urban.pct + logpop + logincome + ukr.lang.pct 
                            + highered 
                            + presidential + orangerev  + n.precincts #+ crisis.year
                            , data=ukrdata)
summary(model.controls.west.p)  

model.local.west.p <- glm.nb(w.roz.count.total ~ #average.wage + 
                           over.sixty + loggov #+ edu.employ + health.employ
                         +urban.pct + logpop + income.percap #+ ukr.lang.pct 
                         + highered 
                         + presidential  + wartime.election + n.precincts + party.margin #+ crisis.year
                        , data=ukrdata)
summary(model.local.west.p)  #party margin negative at

model.regions.west.p <- glm.nb(w.roz.count.total ~ #average.wage + 
                             over.sixty + loggov #+ edu.employ + health.employ
                           +urban.pct + logpop + income.percap #+ ukr.lang.pct 
                           + highered 
                           + presidential + orangerev + wartime.election + n.precincts + factor(region.group) #+ crisis.year
                           , data=ukrdata)
summary(model.regions.west.p)  

model.national.west.p <- glm.nb(w.roz.count.total ~ #average.wage + 
                              over.sixty + loggov #+ edu.employ + health.employ
                            +urban.pct + logpop + income.percap #+ ukr.lang.pct 
                            + highered 
                            + presidential + orangerev + wartime.election + n.precincts + pres.constitution #+ crisis.year
                           , data=ukrdata)
summary(model.national.west.p) 

model.years.west.p <- glm.nb(w.roz.count.total ~ #average.wage + 
                           over.sixty + loggov #+ edu.employ + health.employ
                         +urban.pct + logpop + income.percap #+ ukr.lang.pct 
                         + highered  + n.precincts
                         #+ presidential 
                         + factor(electionid) #+ crisis.year
                         , data=ukrdata)
summary(model.years.west.p) #Morein 2004, less in 2010 compared to 2002

model.interactionyears.west.p <- glm.nb(w.roz.count.total ~ #average.wage + 
                                      over.sixty + loggov #+ edu.employ + health.employ
                                    +urban.pct + logpop + income.percap #+ ukr.lang.pct 
                                    + highered  + n.precincts
                                    #+ presidential 
                                    + party.margin + electionid + 
                                      party.margin*electionid #+ crisis.year
                                    , data=merged.data)
summary(model.interactionyears.west.p) #margin positive, interaction negative in 04, 06, and 14x2

model.interaction.west.p <- glm.nb(w.roz.count.total ~ #average.wage + 
                                 over.sixty + loggov #+ edu.employ + health.employ
                               +urban.pct + logpop + logincome + ukr.lang.pct +
                               + highered 
                               + presidential   + n.precincts + party.margin + yanukovych + 
                                 party.margin*yanukovych #+ crisis.year
                              , data=ukrdata)
summary(model.interaction.west.p) #margin is negative sig, const is negative, interaction is positive sig






model.interaction.west.p2 <- glm.nb(w.roz.count.total ~ #average.wage + 
                                  over.sixty + loggov #+ edu.employ + health.employ
                                +urban.pct + logpop + logincome + ukr.lang.pct + prewar.west
                                + highered 
                                + presidential +  n.precincts + party.margin + rcl.proeast + 
                                  party.margin*rcl.proeast#+ crisis.year
                                , data=merged.data)
summary(model.interaction.west.p2) #margin is negative sig, const is negative, interaction is positive sig



model.interaction.west.p3 <- glm.nb(w.roz.count.total ~ #average.wage + 
                                   over.sixty + loggov #+ edu.employ + health.employ
                                 +urban.pct + logpop + income.percap + ukr.lang.pct + prewar.west
                                 + highered 
                                 + presidential + n.precincts   + 
                                   prewar.west*rcl.proeast #+ crisis.year
                                 , data=merged.data)
summary(model.interaction.west.p3) #margin is negative sig, const is negative, interaction is negative sig

sjPlot::plot_model(model.interaction.west.p3, type = "pred", terms = c("rcl.proeast", "prewar.west"))

model.interaction.east.p3 <- glm.nb(e.roz.count.total ~ #average.wage + 
                                   over.sixty + loggov #+ edu.employ + health.employ
                                 +urban.pct + logpop + income.percap + ukr.lang.pct + prewar.west
                                 + highered 
                                 + presidential + n.precincts   + 
                                   prewar.west*rcl.proeast #+ crisis.year
                                 , data=ukrdata)
summary(model.interaction.east.p3) #margin is negative sig, const is negative, interaction is negative sig

sjPlot::plot_model(model.interaction.east.p3, type = "pred", terms = c("prewar.west", "rcl.proeast [-.25, .3]"))


##East parties

model.controls.east.p <- glm(e.roz.count.total ~ #average.wage + 
                               logold + loggov #+ edu.employ + health.employ
                             +urban.pct + logpop + logincome + ukr.lang.pct 
                             + highered 
                             + presidential + orangerev  + n.precincts #+ crisis.year
                             ,family=poisson, data=merged.data)
summary(model.controls.east.p)  

model.local.east.p <- glm(e.roz.count.total ~ #average.wage + 
                            over.sixty + gov.employ #+ edu.employ + health.employ
                          +urban.pct + logpop + income.percap #+ ukr.lang.pct 
                          + highered 
                          + presidential + orangerev + wartime.election + n.precincts + party.margin #+ crisis.year
                          , family=poisson, data=merged.data)
summary(model.local.east.p)  #party margin negative 

model.regions.east.p <- glm(e.roz.count.total ~ #average.wage + 
                              over.sixty + gov.employ #+ edu.employ + health.employ
                            +urban.pct + logpop + income.percap #+ ukr.lang.pct 
                            + highered 
                            + presidential + orangerev + wartime.election + n.precincts + factor(region.group) #+ crisis.year
                            , family=poisson, data=merged.data)
summary(model.regions.east.p)  

model.national.east.p <- glm(e.roz.count.total ~ #average.wage + 
                               over.sixty + gov.employ #+ edu.employ + health.employ
                             +urban.pct + logpop + income.percap #+ ukr.lang.pct 
                             + highered 
                             + presidential + orangerev + wartime.election + n.precincts + pres.constitution #+ crisis.year
                             , family=poisson, data=merged.data)
summary(model.national.east.p) #Pres const is neg

model.years.east.p <- glm(e.roz.count.total ~ #average.wage + 
                            over.sixty + gov.employ #+ edu.employ + health.employ
                          +urban.pct + logpop + income.percap #+ ukr.lang.pct 
                          + highered + wartime.election + n.precincts
                          #+ presidential 
                          + factor(electionid) #+ crisis.year
                          , family=poisson, data=merged.data)
summary(model.years.east.p) #Less in 2004 through 2012 (compared to 02), more in 2014 parl

model.interactionyears.east.p <- glm(e.roz.count.total ~ #average.wage + 
                                       logold + gov.employ #+ edu.employ + health.employ
                                     +urban.pct + logpop + income.percap #+ ukr.lang.pct 
                                     + highered + wartime.election + n.precincts
                                     #+ presidential 
                                     + party.margin + factor(electionid) + 
                                       party.margin*factor(electionid) #+ crisis.year
                                     ,family=poisson, data=merged.data)
summary(model.interactionyears.east.p) #margin negative, interaction varies

model.interaction.east.p <- glm(e.roz.count.total ~ #average.wage + 
                                  logold + loggov #+ edu.employ + health.employ
                                +urban.pct + logpop + logincome + ukr.lang.pct + 
                                + highered 
                                + presidential + orangerev  + n.precincts + party.margin + yanukovych + 
                                  party.margin*yanukovych #+ crisis.year
                                , family=poisson, data=merged.data)
summary(model.interaction.east.p) #margin is negative sig, const is negative, interaction is negative sig

interplot(model.interaction.east.p, var1="party.margin", var2="yanukovych") #Yan -> increase in east fraud in pro-east regions
p.roz.int.e <- interplot(model.interaction.east.p, var2="party.margin", var1="yanukovych") + #Yan increases east fraud in pro-east regions, reduces it in pro-west
  xlab("Regional margin of victory") + 
  ylab("Estimated coefficient for Yanukovych")+
  theme_bw()   #In pro-west regions, Yan -> reduced west fraud; in pro-east regions -> increase




model.interaction.east.p2 <- glm(e.roz.count.total ~ #average.wage + 
                                  logold + loggov #+ edu.employ + health.employ
                                +urban.pct + logpop + logincome + ukr.lang.pct + 
                                  + highered 
                                + presidential + orangerev  + n.precincts + party.margin + wartime.election + 
                                  party.margin*wartime.election #+ crisis.year
                                , family=poisson, data=merged.data)
summary(model.interaction.east.p2) #margin is negative sig, const is negative, interaction is negative sig

interplot::interplot(model.interaction.east.p2, var1="party.margin", var2="wartime.election") #Yan -> increase in east fraud in pro-east regions
p.roz.int.e2 <- interplot::interplot(model.interaction.east.p2, var2="party.margin", var1="wartime.election") + #Yan increases east fraud in pro-east regions, reduces it in pro-west
  xlab("Regional margin of victory") + 
  ylab("Estimated coefficient for Euromaidan")+
  theme_bw() #+ geom_hline(yintercept=0, linetype=2)  #In pro-west regions, Yan -> reduced west fraud; in pro-east regions -> increase

```

### Interesting models

```{r}
model.interactionyears.east.r <- lm(scale.east.rozenas ~ #average.wage + 
                                      over.sixty + gov.employ #+ edu.employ + health.employ
                                    +urban.pct + logpop + income.percap #+ ukr.lang.pct 
                                    + highered 
                                    #+ presidential 
                                    + party.margin + electionid + 
                                      party.margin*electionid #+ crisis.year
                                    , data=ukrdata)
summary(model.interactionyears.east.r)
sjPlot::plot_model(model.interactionyears.east.r, type = "pred", terms = c("electionid", "party.margin [-.23, .46]"))


model.interactionyears.west.r <- lm(scale.west.rozenas ~ #average.wage + 
                                      over.sixty + gov.employ #+ edu.employ + health.employ
                                    +urban.pct + logpop + income.percap #+ ukr.lang.pct 
                                    + highered 
                                    #+ presidential 
                                    + party.margin + electionid + 
                                      party.margin*electionid #+ crisis.year
                                    , data=ukrdata)
summary(model.interactionyears.west.r)
sjPlot::plot_model(model.interactionyears.west.r, type = "pred", terms = c("electionid", "party.margin [-.23, .46]"))


###########



model.interaction.west.p3 <- glm.nb(w.roz.count.total ~ #average.wage + 
                                   over.sixty + loggov #+ edu.employ + health.employ
                                 +urban.pct + logpop + income.percap + ukr.lang.pct + prewar.west
                                 + highered 
                                 + presidential + n.precincts   + 
                                   prewar.west*rcl.proeast #+ crisis.year
                                 , data=merged.data)
summary(model.interaction.west.p3) #margin is negative sig, const is negative, interaction is negative sig

sjPlot::plot_model(model.interaction.west.p3, type = "pred", terms = c("rcl.proeast", "prewar.west"))

model.interaction.east.p3 <- glm.nb(e.roz.count.total ~ #average.wage + 
                                   over.sixty + loggov #+ edu.employ + health.employ
                                 +urban.pct + logpop + income.percap + ukr.lang.pct + prewar.west
                                 + highered 
                                 + presidential + n.precincts   + 
                                   prewar.west*rcl.proeast #+ crisis.year
                                 , data=ukrdata)
summary(model.interaction.east.p3) #margin is negative sig, const is negative, interaction is negative sig

sjPlot::plot_model(model.interaction.east.p3, type = "pred", terms = c("prewar.west", "rcl.proeast [-.25, .3]"))

```

```{r}
model.interactionyears.east.dev <- lm(deviance_region_east ~ #average.wage + 
                                      over.sixty + gov.employ #+ edu.employ + health.employ
                                    +urban.pct + logpop + income.percap #+ ukr.lang.pct 
                                    + highered 
                                    #+ presidential 
                                    + party.margin + electionid + 
                                      party.margin*electionid #+ crisis.year
                                    , data=merged.data)
summary(model.interactionyears.east.dev)
sjPlot::plot_model(model.interactionyears.east.dev, type = "pred", terms = c("electionid", "party.margin [-.23, .46]"))



model.interactionyears.west.dev <- lm(deviance_region_west ~ #average.wage + 
                                      over.sixty + gov.employ #+ edu.employ + health.employ
                                    +urban.pct + logpop + income.percap #+ ukr.lang.pct 
                                    + highered 
                                    #+ presidential 
                                    + party.margin + electionid + 
                                      party.margin*electionid #+ crisis.year
                                    , data=merged.data)
summary(model.interactionyears.west.dev)
sjPlot::plot_model(model.interactionyears.west.dev, type = "pred", terms = c("electionid", "party.margin [-.23, .46]"))
```

```{r}

model.interactionprewar.east.dev <- lm(deviance_region_east ~ #average.wage + 
                                      over.sixty + gov.employ #+ edu.employ + health.employ
                                    +urban.pct + logpop + income.percap #+ ukr.lang.pct 
                                    + highered 
                                    #+ presidential 
                                    + prewar.west + electionid + 
                                      prewar.west*electionid #+ crisis.year
                                    , data=merged.data)
summary(model.interactionprewar.east.dev)
sjPlot::plot_model(model.interactionprewar.east.dev, type = "pred", terms = c("electionid", "prewar.west"))



model.interactionprewar.west.dev <- lm(deviance_region_west ~ #average.wage + 
                                      over.sixty + gov.employ #+ edu.employ + health.employ
                                    +urban.pct + logpop + income.percap #+ ukr.lang.pct 
                                    + highered 
                                    #+ presidential 
                                    + prewar.west + electionid + 
                                      prewar.west*electionid #+ crisis.year
                                    , data=merged.data)
summary(model.interactionprewar.west.dev)
sjPlot::plot_model(model.interactionprewar.west.dev, type = "pred", terms = c("electionid", "prewar.west"))

```

```{r}
model.interactionprewar.east.dev <- lm(deviance_region_east ~ #average.wage + 
                                      over.sixty + gov.employ #+ edu.employ + health.employ
                                    +urban.pct + logpop + income.percap #+ ukr.lang.pct 
                                    + highered 
                                    #+ presidential 
                                    + prewar.west + rcl.proeast + 
                                      prewar.west*rcl.proeast #+ crisis.year
                                    , data=merged.data)
summary(model.interactionprewar.east.dev)
sjPlot::plot_model(model.interactionprewar.east.dev, type = "pred", terms = c("rcl.proeast", "prewar.west"))



model.interactionprewar.west.dev <- lm(deviance_region_west ~ #average.wage + 
                                      over.sixty + gov.employ #+ edu.employ + health.employ
                                    +urban.pct + logpop + income.percap #+ ukr.lang.pct 
                                    + highered 
                                    #+ presidential 
                                    +  rcl.proeast + 
                                      prewar.west*rcl.proeast #+ crisis.year
                                    , data=merged.data)
summary(model.interactionprewar.west.dev)
sjPlot::plot_model(model.interactionprewar.west.dev, type = "pred", terms = c("rcl.proeast", "prewar.west"))


ggplot(data = merged.data %>% filter(prewar.west == "1"), aes(x = deviance_region_east, y = east.party.voteshare)) + geom_point()


```





OPORA trained police officers in 2019: https://www.oporaua.org/en/vybory/17992-pid-chas-prezidentskikh-viboriv-treneri-opori-navchili-bilshe-1-5-tisiachi-politseiskikh-17992 

Court cases can be found here: https://reyestr.court.gov.ua/
Look for Criminal cases, under the criminal code, then violations of electoral rights. Then there are subcategories like bribery of voters (pidkup vybortsya) and falsification. Legal code is here: https://sherloc.unodc.org/cld/uploads/res/document/ukr/2001/criminal-code-of-the-republic-of-ukraine-en_html/Ukraine_Criminal_Code_as_of_2010_EN.pdf

Ухвала is decision, Постанова  is decree, Вирок is sentence.

This could be used to provide some qualitative support for the argument about risk.