# Election manipulation, patronage, and agent risk in Ukraine

\chaptermark{Election manipulation in Ukraine}

Ukraine's 2004 presidential election was a moment of crisis for the incumbent regime. The stakes, as with many post-communist elections, were high. Turnover in the presidency would put the post-communist distribution of resources and power at risk, threatening the fortunes (and perhaps the liberty) of allies of the incumbent [@kuzio_kuchma_2005]. A deeply unpopular and scandal-plagued President Kuchma was unable to run again. Instead, the prime minister, Viktor Yanukovych, became the standard-bearer for the incumbent coalition. Despite coming in second in the first round of the election, the official results declared Yanukovych to be the winner in the second round. Amid allegations of widespread fraud, hundreds of thousands of citizens took to the streets, braving the Ukrainian winter for weeks on end. Amid cascading defections from within the incumbent elite, the parliament and the supreme court endorsed a re-run of the second round [@kuzio_kuchma_2005]. Yanukovych lost the re-run to Viktor Yushchenko, ushering in several years of economic and political reforms.

On one level, the outcome of the Orange Revolution vividly demonstrates the risk to incumbents of rigging an election. As such, the Orange Revolution looms large in the literature on election manipulation and protest; it is perhaps the paradigmatic example of the costs that governments can pay after rigging an election. On a closer read of the case, however, presents a more complicated picture. Preparations by civil society and opposition groups to monitor the election and protest the outcome were extensive and highly visible. The two main opposition camps coordinated in advance of the election. Opposition activists mobilized as many as 50,000 protesters in the streets *before* the election. In a highly polarized society, the risk and cost of protest were both high. Understood in this way, the Orange Revolution is a deviant case for the 'regime-weakness' model of manipulation. If electoral manipulation signals weakness, and protest risk can therefore deter manipulation, why didn't the Kuchma / Yanukovych group back down and hold clean(er) elections?

"Seldom has the coming of a major political event been so evident," Anders Åslund wrote of the Orange Revolution [@aslundHowUkraineBecame2009]. Both the incumbent's intention to manipulate the results and the opposition's intention to stage mass protest were apparent long in advance of election day. By early 2004, a coalition of Ukrainian civil society groups had developed "Wave of Freedom," a program to combat election manipulation through analysis, election monitoring, recording of violations, and social mobilization [@demesPora2006]. The youth-led movement Pora ("It's Time"), modeled on earlier pro-democratic movements in Serbia and Georgia, announced its intention to organize mass protest in the event of fraud already in March of 2004 [@demesPora2006]. In advance of the presidential vote, Pora activists used local elections to develop their techniques and build networks. By the time of the fall campaign, Pora had amassed a network of over 30,000 volunteers across the country, and staged seventeen pre-election rallies [@demesPora2006]. 

Its activities touched a nerve, and the authorities began a campaign to repress the organization. Several hundred activists were arrested in the fall of 2004; in October, Pora's Kyiv headquarters were raided by the police who "discovered" planted explosives [@demesPora2006]. By mid-October, Pora was taking a lead role in coordinating protest mobilization with the "Freedom of Choice" coalition of NGOs and the Yushchenko campaign [@demesPora2006]. Yushchenko himself drew as many as 100,000 rally-goers to the streets on more than one occasion, including one where the demonstrators marched on the Central Election Commission to demand an honest election  [@aslundHowUkraineBecame2009, p. 187]. Anticipating fraud in the second round, Pora activists established the first protest camps on city squares in Kyiv and other major cities after the first round of the vote. Activists' presence and logistical work provided the conditions for hundreds of thousands of Ukrainians to join them, after the heavily manipulated second round [@demesPora2006].

On the other side, the incumbent coalition was divided. Only one oligarchic group fully supported Yanukovych, with the others remaining ambivalent or supporting Yushchenko. The security services were likewise divided [@aslundHowUkraineBecame2009]. Bessinger [-@beissingerMechanismsMaidanStructure2011, p. 28] argues that the Orange Revolution was part of a long-term development, and the high degree of popular mobilization was "less a spontaneous reaction by individuals to the stolen election than the product of the heightened organization that emerged during the electoral campaign." In this context, the decision to go all-in on manipulation---rather than a signal of incumbent weakness that prompted strategic protest---can be interpret as an attempt to unify the elites around Yanukovych, knowing full well that protest was likely no matter the level of manipulation.

The Ukrainian case provides an opportunity to test ther regime-strength model of electoral manipulation in a less politically consolidated case (compared to Russia). No ruling party has ever managed to dominate Ukrainian politics in the way United Russia does. Instead, Ukraine's political history from 2002 to 2014 (the period covered here) is one of crisis, alternation in power, polarization, democratization and democratic backsliding. In other words, the case provides an opportunity to test the regime-strength theory of election manipulation at the opposite end of the spectrum from Russia: where the regime is highly competitive and unconsolidated, though still in the context of extensive patronage networks and weak rule of law.

Ukrainian politics at the time was roughly divided into two broad camps: a largely pro-Western camp which came to be known as the Orange coalition, and a largely pro-Russian camp centered around the Party of Regions.^[This is, of course, a shorthand description that overlooks considerable nuance in the two camps. These nuances are discussed in more detail below.] Each of these camps had durable bases of political support, substantial legislative representation at the national level, and support from wealthy oligarchs. Alternative models of election manipulation make distinct predictions for the Ukrainian case, as a result. 

Models that assume election manipulation signals weakness would predict restraint on incumbents' part. In a closely divided polity, where the opposition is well-resourced and well-organized, incumbents should anticipate that substantial, costly protest would be the outcome of any manipulation effort large enough to make a difference in the election. Likewise, a grievance-based model of electoral protest would predict election manipulation would galvanize a large portion of the population---supporters of the out-of-power group---to protest. In either case, a regime-weakness framework should expect low levels of manipulation generally in Ukraine, especially after the Orange Revolution.

By contrast, the regime-strength framework---which holds that illicit electoral manipulation acts as a costly signal of strength to elites---implies that manipulation should persist in Ukraine. Moreover, it predicts that heightened agent risk (due to political fragmentation) should make attributable manipulation like fraud uncommon. Instead, when patronage resources are available, agents should be more willing to participate in dispersed forms of illicit manipulation (e.g. vote-buying, voter pressure).

As this chapter will show, the regime-strength model is supported. Election forensic analysis of election data from 2002 to 2014 is used to estimate the level of overall manipulation in favor of pro-Eastern and pro-Western parties in each region. An election-forensic measure of falsification is likewise generated for each region. The results show no evidence of systematic falsification across the period. Nevertheless, other more dispersed forms of electoral manipulation are detected. This turnout-based manipulation is amplified when patronage resources are more consolidated in government or through coalitions of oligarchs. Manipulation of all kinds is least likely where the risk of agent exposure is highest

The results indicate that the implications of the regime-strength model travel beyond the Russian case. The comparison to Ukraine is especially useful, since it allows for variation of the central factors in the model---patronage resources and local agent risk---while holding many other institutional and historical factors constant. 

## Ukraine's political history, 2000 to 2014

Though analysts should be wary of over-emphasizing the point, Ukraine’s political divisions during this period were broadly regional. Voters in the country’s east generally supported parties and candidates that adopted a more interventionist economic policy and took a more favorable view toward Russia. By contrast, voters in the west were more likely to support parties and candidates with economic and foreign policies oriented towards the market and towards Europe (Birch, 2000).  The two regions are characterized by ethnic and linguistic divisions, with the west home to more ethnic Ukrainians and Ukrainian-speakers, while Russian ethnicity and language is more common in the east. While this chapter uses 'pro-Western' and 'pro-Eastern' as an analytical shorthand, it is important to note that these labels average over considerable variation within each group.

First, "Pro-eastern" and "Pro-western" attitudes are not immutable. In the 1991 referendum, majorities in every region of the country---including the predominantly ethnic Russian Crimea---voted in favor of independence from the Soviet Union.^[While bitterly fought in the media, the referendum appears to have been conducted in a free and fair manner[@potichnyjReferendumPresidentialElections1991].] Politicians, voters, and regions have shifted in and out of each camp over time. After Russia's incursion in Ukraine in 2014 and full-scale invasion in 2022---outside the time period covered here---Ukrainian national identity and alienation from Russia have become more common across the country [@kulykNationalIdentityTime2024;@kulykSheddingRussiannessRecasting2018].

Moreover, there is considerable heterogeneity within each regional group. These differences can be attributed, in part, to historical legacies: much of western Ukraine was ruled by the Austro-Hungarian Empire prior to World War I, and by Poland, Romania, or Czechoslovakia during the interwar years. Government policy in these regions---Lvivska, Ternopilska, and Ivano-Frankivska oblasts---before they were incorporated into the Soviet Union in 1939, was relatively democratic and favorable to the development of Ukrainian national identity. At a time when national identities were beginning to form in Europe, a civil society imbued with early Ukrainian national identity developed in these regions: parties, newspapers, cultural and religious institutions, and more [@katchanovskiRegionalPoliticalCleavages]. These regions, along with Volhynia, did not become part of the Soviet Union until World War II. Moving east, a large central portion of modern Ukraine remained part of the Polish-Lithuanian Commonwealth for roughly another twenty years before being ceded to the Russian Empire. Lastly, the easternmost regions of modern Ukraine never experienced Polish or Habsburg rule; as part of the Russian Empire, civil society was limited and the Ukrainian people were considered a junior branch of a broader Russian ethnicity [@liebichUkrainianPresentLegacies2019]. These legacies, along with regional differences in language use, religion, industrialization, and urbanization continued to influence Ukrainian politics during the period studied here [@Barrington2009;@Osipian2012], differences that are particularly stark in election outcomes if not in pre-election public opinion [@Frye2015].


By contrast, eastern Ukraine was part of the Russian Empire, which emphasized a Ukrainian identity that was a regional subset of the broader Russian ethnicity and language [@Katchanovski2006]. Despite the well-documented importance of language [@fournier_mapping_2002;@kulyk_language_2011] and ethnicity [@barrington_examining_2002] in Ukraine, however, the regions’ voters are not homogeneous blocs [@Sasse2010]. For example, using a survey experiment, Frye (2015) finds that a hypothetical candidate’s position on closer ties with Europe or Russia is more likely to drive a voter’s preference than either the candidate’s reported ethnicity or language. Furthermore, many of Ukraine’s eastern regions are more diverse linguistically and ethnically than is generally reported .


While regional attitudes follow the general contours described above, it is important to note that major Ukrainian political parties have not been characterized by stark ideological conflicts so much as by regional loyalties, affiliation with particular leaders, and the distribution of patronage. Parties in Ukraine tend to be highly personalized, and frequently depend on the resources, reputation, and charisma of particular leaders for their electoral success [@Rybiy2013]. This is especially true of the western-oriented parties associated with Viktor Yushchenko and Yulia Tymoshenko [@Kuzio2012;@Kudelia2015]. This personalization contributed to parties’ ideological fluidity^[Some parties are exceptions to this rule: the ideologically cohesive Communist Party of Ukraine and Svoboda (Freedom) did well in the 2012 parliamentary election, winning 13\% and 10\% of the vote, respectively [@Rybiy2013].] and voters’ difficulty in placing parties on an ideological spectrum [@Rybiy2013].  

Additionally, as in other post-communist countries, Ukraine’s economic transition created a class of oligarchs who were able to gain control of the state’s major economic assets.  In Ukraine, these oligarchs are often linked in regional clans and can use their resources to back local politicians.  These patterns can be seen even in independent Ukraine’s earliest elections, but intensified following the Orange Revolution [@DAnieri2011]. 

This weak ideological commitment on the part of voters, combined with social inequality and overlapping political and economic responsibilities of local leaders [@Allina-Pisano2010], allows parties to rely on `administrative resources' like access to state employment, favorable (or unfavorable) business treatment, and selective law enforcement to build and maintain networks of supporters [@DAnieri2011;@protsyk_centre_2003]. Tools like these give political leaders the ability to raise funds, mobilize voters, and manipulate elections [@Hale2003]. These tools are usually held by regional or local actors in Ukraine, rather than being institutionalized at the national level [@Way2005]. In particular, such methods are characteristic of the pro-eastern parties exemplified by the Party of Regions [@Kudelia2015;@Kuzio2012].

A large public sector and a relatively compliant judiciary [@kudelia_sources_2012] provide ample opportunity for the executive branch to apply pressure to voters and opposition figures in Ukraine. Post-Soviet reforms subordinating regional governors to the president created ''a strong vertical executive power running from the president down to the village level"---in order to retain office, regional and local officials were required to mobilize voters on behalf of their bosses [@Konitzer-Smirnov2005]. As in Russia [@frye2014], state employees and workers in industries that are vulnerable to state regulation are susceptible to pressure from above.  In one particularly stark example, school children at a school in eastern Ukraine informed their parents that their teachers would be fired if the parents did not support the government in an upcoming election [@danieri_understanding_2007]. 

During the 2004 presidential election, which sparked the Orange Revolution, the government leaned on administrative resources to encourage “the police, the army, prisons, public universities, schools, and hospitals, to campaign and vote for Prime Minister Viktor Yanukovych” [@Katchanovski2008]. The Party of Regions, which was led by Yanukovych and came to be the dominant party in eastern Ukraine and Crimea, relied on a mix of state administrative resources, political patronage, and intra-elite bargaining to maintain its support in the region [@Kudelia2015;@myagkov_forensics_2009]. Patronage, clientelism, and voter pressure are all common features of the Ukrainian election system in this period, and---as in Russia---it is characterized by a political hierarchy that links the president to local officials through intermediaries like regional officials and oligarchs. 


Ukraine’s election law divides the country into 225 district election commissions. Rules determining membership on the committees have varied from year to year, but have generally granted automatic representation to representatives of the parliamentary parties, with remaining seats filled by non-parliamentary parties via lotteries. This setup allowed for small, hastily registered parties—operating informally on behalf of major parties or politicians—to gain representation on the district commissions. This problem became especially egregious in 2012, when the Central Election Commission implemented a change in the lottery process five days before the drawing: instead of individual lotteries for each district commission, a single lottery would be held that would determine the makeup of all 225 commissions. As a result, small parties that had nominated only one or two candidates nationwide gained representation on all district commissions, while larger opposition parties were excluded [@kovalov_electoral_2014;@OSCEODIHR2013]. District election commissions then implemented a similar nominating procedure for all their subordinate precinct election commissions.

This system can be especially problematic when combined with the deep patronage roots enjoyed by the Party of Regions in eastern Ukraine, which it used to gain outsize influence on the commissions. In one notable example, OSCE observers in a single election district of Donetsk oblast' determined that 1,667 of the 2,551 precinct election commissioners in the district worked for an enterprise headed by the Party of Regions candidate—meaning that more than two-thirds of precinct commissioners presided over balloting for their employer [@OSCEODIHR2013]. 

Partisan control of election commissions at various levels facilitates electoral manipulation in various forms, which observers frequently witnessed in Ukraine. In 2004, the OSCE observed—among other issues—ballot box stuffing, irregularities with absentee voting, large numbers of voters being turned away from the polls, implausible turnout figures, and suspicious tabulation process. All of these forms of electoral manipulation are either implemented or influenced by election commissioners at the precinct, territorial, or central level.  Similar problems were observed by the OSCE in 2012, when observers negatively assessed the vote count in eleven percent of monitored precincts, and tabulation was negatively assessed in almost half of the 161 district election commissions monitored [@OSCEODIHR2013].

While the presence of technical representatives on election commissions was perhaps most extensive in the 2012 campaign, it figured prominently in previous elections as well. In the first round of the 2004 election, technical candidates had the right to nominate up to 66,000 members of precinct election commissions; while both major candidates took advantage of technical candidates, the OSCE concluded that Yanukovych benefited the most, finding that supportive technical candidates gave him ``a de facto majority on election commissions'' [@OSCEODIHR2005]. The 2010 election, in which Yanukovych was first elected president, was relatively free of electoral fraud, with candidates emphasizing voter mobilization [@Herron2011]. However, the analysis presented here suggests that some of these mobilization efforts may have been extra-legal and/or characterized by the misuse of state resources [@OSCEODIHR2010]. 


In sum, the context of the case is one in which formal democratic structures coexisted with deeply embedded patronage networks, which were consistently employed to undermine electoral integrity. A relatively polarized electorate, however---partially reflecting historical socioeconomic and cultural differences across regions---helped prevent any one party or faction from consolidating control.   particularly benefiting pro-eastern political forces. 


### Frustrated efforts to consolidate power

Soviet Ukraine was characterized by heavy reliance on patronage and clientelism, transmitted by informal networks operating within the formal Communist Party and command economy structures in a system Kitschelt calls [-@Kitschelt1995] terms patrimonial communism. This framework persisted in post-Soviet Ukraine, as many political and bureaucratic officeholders remained in place, while communist-era managers and officials became private-sector owners of capital [@Zimmer2008]. Especially in the heavily industrialized eastern parts of the country, where small-scale enterprises are less common [@Aslund2005], large portions of the population are dependent on state- or employer-based patronage [@Zimmer2005]. A power vertical has been solidly established in the eastern part of the country, based on an alliance of regional and local administrative officials and business actors [@Zimmer2005]. Nationally, the pro-presidential constitution of 1996 subordinated regional governors and local executives to the president [@Konitzer-Smirnov2005], a structure which remained largely unchanged under the post-Orange constitution. 

However, the Cassette Scandal crisis, also known as 'Kuchmagate'\textemdash in which illicit recordings of the president allegedly ordering a kidnapping, among other unsavory details, were made public\textemdash limited the ability of President Kuchma to further consolidate control over the patronage resources and hand them to his preferred candidate in the 2004 presidential election, Viktor Yanukovych. As a result of the ensuing political crisis, the party system fractured, oligarchic alliances were ruptured, momentum toward a super-presidential constitution was scuppered, President Kuchma’s succession plans were thrown into turmoil, and an opposition movement based in civil society began to form [@Kuzio2007]. The scandal helped lay the foundations for the 2004 Orange Revolution, in which hundreds of thousands of citizens took to the streets to protest electoral manipulation in the presidential elections. The subsequent opposition victory in a repeat election upended the party structure, and helped usher in a new constitution that divided power more evenly between the presidency and the parliament.

As part of a compromise following the Orange Revolution, a new constitution was adopted which greatly increased the patronage capabilities of the parliament and prime minister. As a result of the new constitution, the parliamentary majority gained more discretion over the nomination of prime ministers, more influence over the defense and security ministries and the prosecutor general, and shifted much executive staffing power away from the president and to the cabinet [@Christensen2005]. This shift reinforced the de facto competing-pyramid structure of patronage in the post-revolutionary period, by expanding and legitimizing the influence of the parliamentary majority while still retaining some patronage powers in the office of the presidency [@hale_patronal_2014]. 

The competing pyramid structure would not survive long into the Yanukovych presidency, when the Party of Regions was eager to press its advantage in patronage resources. After defeating Tymoshenko’s network in the 2010 presidential election, Yanukovych moved swiftly to encourage elite defections within the Rada, building a pro-presidential coalition that secured control of the premiership [@hale_patronal_2014;@kudelia_sources_2012]. Subsequently, Yanukovych was able to win restoration of the pre-revolution constitution favoring the president; he rapidly reconsolidated patronage networks that had become fragmented during the Orange period. 

This centralizing period would be short-lived however, as massive crowds during the 2014 Euromaidan crisis helped depose Yanukovych, leading to the seizure of the Crimean peninsula by Russia, the de facto secession of two of Ukraine’s most populous and industrialized regions, a prolonged civil war, and the effective collapse of the pro-eastern Party of Regions. The patronage networks that had helped propel the success of the Party of Regions locally and nationally began to unravel [@hale_patronal_2014]. 

### The role of the oligarchs

In addition to the shifting consolidation of state-based patronage, analysts of Ukrainian politics must also contend with a complex arrangement oligarchic ‘clans,’ which deploy their resources to support favored parties. Unlike in Russia, where the major oligarchs generally do not risk challenging the Kremlin Ukraine’s oligarchic clans have been more fractious. During Kuchma’s presidency, three regional oligarchic clans became influential in politics: the Dnipropetrovsk network, the Donetsk network, and the Kyiv network. All three were generally aligned with Kuchma until the Orange Revolution [@Pleine2016], and used their resources to influence election outcomes [@Katchanovski2008]. The revolution split the political-oligarchic patronage networks into three competing camps\textemdash one each aligned with Viktor Yushchenko, Yulia Tymoshenko, and Viktor Yanukovych\textemdash a framework that persisted until the 2010 presidential election [@Hale2010]. Even after being cast into the political wilderness nationally by the Orange crowds, Yanukovych and the Party of Regions could rely on regional patronage networks and elected local mayors to retain influence in the east [@Hale2010]. 

When Yanukovych assumed the presidency in 2010, the fractiousness of the opposition coalition of oligarchs gave way once again to a more united patronage pyramid around the presidency and the Donetsk clan. In turn, this informal patronage network collapsed as a result of the Euromaidan and subsequent civil war, generally reconstituting itself around oligarch Petro Poroshenko after his 2014 presidential victory [@Pleine2016]. Poroshenko, though his own election was largely clean, used patronage resources to good effect (especially in the East, where networks still remained largely intact) in the subsequent legislative elections [@Fedorenko2016].

Some oligarchs control industrial conglomerates and associated patronage networks [@puglisiRiseUkrainianOligarchs2003;@lankinaSovietLegaciesEconomic2019]; others control mass media outlets [@wilsonSurvivalRichestHow2016]; many use their wealth to prop up 'spoiler parties' and 'clone candidates' to draw votes away from opposition parties [@wilsonSurvivalRichestHow2016]. Oligarchs in Ukraine are a potential source of patronage resources, but cannot on their own promise impunity for electoral crimes. Their ability to protect themselves and their agents from legal punishment hinges on their relationship with the state authorities, which can shift over time and with election outcomes [@zubytskaRiseEstablishmentOligarchs2018].  


### Risks of manipulation exposure



Given the divisions of the country, local constraints in Ukraine must be conceptualized for each broad political group, referred to here by the shorthand terms ‘pro-western’ and ‘pro-eastern.’ Risks vary for each group according to local conditions. For example, the pro-eastern Party of Regions has been able to exert influence over local courts in regions it controls [@Katchanovski2008], reducing the risk faced by its own agents while exacerbating them for pro-western agents. The Party of Regions inherited much of the political culture and resources of the Communist Party of Ukraine, and used these advantages to build a locally dominant coalition of government administrators, red directors (communist-era managers of large firms), and local business / oligarchic interests [@Kudelia2015;@Zimmer2008]. As a result, the heartland of the PoR is politically sealed off [@Osipian2006].

The situation is more complicated in regions where pro-western parties tend to dominate. Machine politics, essential to delivering votes in places like Donetsk and Luhansk in the east, were difficult to export to territories in Western Ukraine [@Sushko2002;@Zimmer2005], due to a more politically active civil society there [@DAnieri2005] and the relative absence of locally dominant economic interests [@Kudelia2015;@Kuzio2012]. This lack of consolidation produced two major pro-Western party groups, primarily oriented around the figures of Yushchenko and Tymoshenko. As a result, these two blocs competed with each other for votes within their territories [@Kuzio2013], limiting the ability of either bloc to develop the kind of local political monopoly that reduces risks for election manipulating agents. Prior to the Orange Revolution, in fact, election administration in western and central Ukraine was largely controlled by the incumbent (largely pro-Eastern) authorities [@wilsonUkrainesOrangeRevolution2006, p. 117].

This elevated and unpredictable level of risk is illustrated in an anecdote from the second round of the 2004 presidential election, which sparked the Orange Revolution. As part of its wide-ranging efforts to manipulate the election, the Yanukovych camp had imported a large quantity of fake ballots from Russia with which to stuff the ballot boxes. Unfortunately for them, the fakes looked very different than the official ballots. Three days before the vote, an unknown subordinate anxiously calls Yurii Levenets, a leading political 'technologist' for the Yanukovych campaign, euphemistically referring to the fake ballots as "cookies." 

 >UNKNOWN: Have you seen at all what we have? Have you seen 'cookies' in boxes?
 
 >LEVENETS: Sure, I've seen 'cookies.'
 
 >UNKNOWN: Have you seen that it differs significantly from the 'originals'?
 
 >LEVENETS: I know. YES, YES YES. I just cannot understand, why it differs.
 
 >UNKNOWN: It differs drastically. I mean ... It's different everywhere. What should we do?
 
 >LEVENETS: I don't know.
 
Four minutes later, Levenets calls another leading member of the Donetsk clan, the minor oligarch Serhii Kliuiev to discuss the problem.

>LEVENETS: Vasilii Mykhailovych just called; he also noticed that the 'cookies' are very different. Did you talk about this with Andrii [Kliuiev]?

>SERHII KLIUIEV: No.

>LEVENETS: Why not?

>KLIUIEV: I don't know. This is bullshit. It can be easily discovered, especially here; it will work over there [in Russia], but not here. Damn, I don't know what to do.


Ultimately, the Yanukovych campaign utilized the poorly-baked 'cookies', but only in areas where it could control the electoral commissions. Underscoring the byzantine nature of risk in this case, we have access to these recordings because one element of the Security Service of Ukraine (the SBU) had tapped the Yanukovych team's phones, and passed the recordings to the Yushchenko campaign [@wilsonUkrainesOrangeRevolution2006, pp. 2-5]. 

The development of active civil society in the west accelerated as a result of the Kuchma-gate crisis, and was nurtured by opposition parties during the Orange Revolution [@Kudelia2007]. However, the historical roots of civil society and multiparty democracy run deeper in Western Ukraine than in the East. During perestroika, dissidents and former political prisoners in Western Ukraine formed Rukh (the Ukrainian Popular Movement for Restructuring) to challenge the Communist Party; while Rukh gained popularity in Kyiv and central Ukraine, it never gained much of a foothold in the east [@Kuzio2010]. In surveys from the early post-Soviet period, respondents from western Lviv were twice as likely as residents of Donetsk to attempt to "resolve social problems" by contacting mass media, contacting officials, or joining an organization, and three times as likely to have signed a petition or participated in a demonstration [@Aberg2000]. These connections persisted through the Orange Revolution: protesters in that movement were motivated less by democratic ideology and more by the fact that they were highly networked, with higher incomes, and less dependent on the state [@Beissinger2013].

As in the Russian case, demonstrated in more detail in Chapter 4, the Ukrainian judicial system does address allegations of election crimes. As in Russia (and as predicted in Chapter 2), it is difficult for the legal system to penalize the organizers of vote-buying or multiple-voting rings. Rather, a precinct election official is a more typical defendant. For example, three precinct election officials in one highly competitive single-member district were convicted of attempting to falsify the election returns [@No28335602015]. Effectively caught in the act---partisan and non-partisan observers found the pre-filled returns---the officials refused a plea bargain but attempted to convince the court that their actions were innocent mistakes. Upon weighing the testimony of eight witnesses, the court found the officials' protestations unconvincing, and sentenced them to five years imprisonment (converted to one year of probation). The race for the seat in District 66, where these events happened, was highly competitive---the winner claimed the seat with only 19 percent of the vote [@CentralElectionCommission].


Though there are, of course, differences across the two countries, Russia and Ukraine share a similar institutional heritage with regard to courts and criminal procedure. Both countries formally insulated the judiciary from political control in the 1990s (e.g. by putting corporate bodies of judges in charge of promotions to higher courts), while still retaining the influential role of the court chairperson; the chair as a result becomes an important lever for imposing outside influence on the court [@trochevPatronalPoliticsJudicial2018;@popova2012politicized]. Both before and after the Orange Revolution, judicial networks were embedded in broader patronage networks aligned around Ukraine's political divisions [@trochevPatronalPoliticsJudicial2018]. Writing on the post-Orange period, Trochev notes that:

>"Judges also no longer hide the fact that they receive orders from the important figures on how to decide the outcomes of cases. Judges from local courts in Eastern Ukraine complain that officials from the Presidential Secretariat phone them; judges from courts in Kyiv complain that MPs and Kyiv City Council members phone them; judges of economic courts complain that officials from the Cabinet of Ministers phone them [and so on]..." [@Trochev2010]. 

A 2008 survey of eight Ukrainian regions (from across the country) found that 34 percent of judges reported 'improper influence' by political parties; 40 percent reported improper influence from national legislators. Threats to their careers were the most common type of pressure reported [@Mon2008]. 

Prosecutors were also surveyed. Thirty-one percent of prosecutors reported improper influence from the chair of the court responsible for a case, and 37 percent reported pressure from higher courts [@Mon2008]. Just as in Russia, newly independent Ukraine inherited a neo-inquisitorial criminal justice system from the Soviet Union, "in which the preliminary investigation, not the trial, was the decisive stage of proceedings" [@foglesongCrimeCriminalJustice. p. 58]. Both countries inherited the institution of the Procuracy, a body with both investigatory and prosecutory powers. As in Russia, the Ukrainian police and procuracy had considerable discretion in the pre-trial phase; in the 1990s, for example, the number of crimes reported to police surged while the *percentage* of those crimes officially registered by police declined [@foglesongCrimeCriminalJustice]. 

Moreover, the procuracy during the 1990s and 2000s was a highly centralized and hierarchical institution. All prosecutors were ultimately under the direction of the Prosecutor General, who was himself appointed and dismissed by the president with the consent of parliament. An opinion of the Venice Commission found that prosecutors' short terms of office combined with possibility of reappointment undermined the independence of the procuracy. Prosecutors General are therefore "under pressure to act in a manner which conforms to the wishes of both the President and Parliament" [@veniceUkraineProcuracy2001, p. 3].  Politicization of the procuracy was a throughline across the Kuchma, Yushchenko,^[During the Yushchenko presidency, however, reforms shifted the the Prosecutor-General's appointment to rest with the majority in the Rada.] and Yanukovych presidencies: during Yanukovych's term, as many as 15,000 corruption cases were quashed by the procuracy [@kudelia_sources_2012]. From the perspective of low-level agents, however, the direction of the politicization may be uncertain when the presidency and the parliament are controlled by different factions.


Risks to election-manipulating agents are heightened by the presence of partisan and non-partisan election observers. In Ukraine's more competitive environment, parties can dispatch election monitors to try to catch their opponents' agents in the act. Observers' role in documenting the abuses that preceded the Orange Revolution is well documented [@Bunce2006;@mcfaulUkraineImportsDemocracy2007]; their importance in increasing risk to agents is less so. Nevertheless, as Ukraine has become more democratic, civil society has invested more resources in the criminal-legal aspect of election monitoring. Beginning in 2014, the civil society group OPORA increasingly worked to bring electoral crimes to the attention of the courts. This effort grew to include nationwide police trainings (ultimately reaching over 70,000 officers), and a full training course for officers at Lviv University of Internal Affairs. Highlighting this shift at the inauguration of the university training course, the OPORA chair Olga Aivazovska said:

> I would like to remind you that between 1991 and 2015 there was not a single conviction against persons who participated in or organized voter bribery. And I can say that those indictments that were handed over after 2015, those court decisions that imposed sanctions either in the form of restriction of liberty or imprisonment, or in the form of a fine...are a progressive consequence of this work [@ОПОРАСпільноЛьвівським].


In sum, local risks vary considerably across Ukraine by both party and region. The risk of exposure for pro-western agents is especially high in eastern regions where the political machine organized around the Party of Regions predominates, due to that party’s success in creating a subnational enclave prior to the Euromaidan crisis, Russian intervention, and civil war. However, the existence of competing pro-western parties and a robust civil society in many western regions makes electoral manipulation difficult to achieve undetected in those regions as well. 


## Hypotheses and other empirical tests

The regime-strength model of electoral manipulation claims that incumbents reduce agency loss among front-line agents to the extent that they are able to develop tools of social control that reduce agents' level of risk. Where incumbents accomplish this goal, as in Russia, the model predicts high levels of manipulation, with falsification occurring in places where the regime is most consolidated and less efficient forms of manipulation occurring where risks to agents are locally higher.

In Ukraine, agency loss is expected to be high. Uncertainty over which camp will control national executive over the medium term---and thus uncertainty over who will control the prosecutorial service and have influence over the judiciary should increase the expected cost of engaging in falsification. This risk of exposure is heightened by the widespread presence of partisan and civil-society observers in precinct and district election commissions. Greater political competitiveness increases the risk that agents of the out-of-power faction will face punishment in the courts [@popova2012politicized]; that same competitiveness makes it difficult to know *a priori* which side will win the election.

It is thus expected that falsification will be rare and unsystematic in Ukraine, as it will be difficult for parties to recruit agents willing to take the risk of exposure and punishment. Forensic measures of fraud should therefore be uncorrelated with parties' vote-shares. In other words, I expect that the null hypothesis will *not* be rejected in statistical tests of that correlation; consequently, I list this as an empirical implication of the theory, rather than a hypothesis. 

### Patronage consolidation

Instead, I expect that other forms of manipulation will predominate, as principals recruit agents for tasks further down the attribution ladder (see Chapter 2). When outright falsification is unavailable, incumbents with resources will seek to convert those resources into votes through ballot-stuffing, vote-buying, and voter-pressure.<!-- All of these tools are less cost-effective than falsification however, and open up the possibility that agents will accept patronage benefits without fully delivering their share of votes. --> Nationwide, I expect pro-eastern manipulation to increase when the president has a pro-eastern orientation (these presidencies also coincide with the pro-presidential constitution), and when pro-Eastern parties dominate in the legislature. 

> H1a: Overall manipulation pro-Eastern manipulation will increase when a pro-Eastern president is in office.

> H1b: Overall manipulation pro-Eastern manipulation will increase when pro-Eastern parties have greater representation in the Verkhovna Rada.

This expectation also yields hypotheses at the regional level, where it is possible to take into account the existence and orientation of oligarchic political machines. Where oligarchic political machines are well-established, pro-Eastern manipulation should increase when those oligarchs are supportive of the pro-Eastern party or candidate. In the Ukrainian context, these political machines are most well-developed in the areas of the country where Soviet industrial policy favored heavy industry and industrialization [@lankinaSovietLegaciesEconomic2019]. This produces the following hypotheses:

>H2a: Overall pro-Eastern manipulation will increase in regions where supportive oligarchs control significant assets.

>H2b: The effect of supportive oligarchs on pro-Eastern manipulation will increase with urbanization.

### Local risk

The theory expects that less attributable forms of electoral manipulation will increase in low-risk areas when patronage consolidation is high. In the Ukrainian case, I expect that pro-Eastern manipulation will increase in regions where pro-Eastern parties dominate (i.e. where risk of exposure is low), when a pro-Eastern president is in office. When a pro-Western president is in office, the reorientation of national-level patronage resources away from the pro-Eastern network should lead to a reduction in manipulation in those areas. Finally, overall pro-Eastern manipulation should remain low in regions where the risk of exposure for pro-Eastern agents is high, regardless of presidential control.

>H3: Pro-Eastern manipulation should increase with pro-Eastern parties' regional control, but only when a pro-Eastern president is in office.

>H4: Pro-Eastern manipulation should be lowest in regions that were not part of Soviet Ukraine before World War II.


## Data and methods

To test these hypotheses, I draw on precinct-level results from nine elections in Ukraine from 2002 to 2014. This includes three presidential elections---the third round of the 2004 elections, along with both rounds of the 2010 and 2014 elections---and four legislative elections (2002, 2006, 2007, and 2012). Unfortunately, the dataset excludes the first two rounds of the 2004 presidential elections, for which precinct-level results are not publicly available. Each election-year includes data from approximately 30,000 to 35,000 precincts nested within Ukraine’s 27 regions. As a result of the annexation of Crimea and the conflict in eastern Ukraine, the legislative and presidential elections in 2014 draw on approximately 28,000 precincts within 23 regions: Crimea, Sevastopol, Donetsk, and Luhansk regions drop out of the data. This precinct-level data is used to conduct election-forensic estimates of falsification and overall evidence of electoral manipulation. Unlike in the Russian case, I am unable to estimate levels of vote-buying and related tactics, since Ukraine did not release precinct-level numbers of absentee or mobile-ballot-box voting until the 2014 elections. Explanatory and control variables are taken at the regional level.



### Dependent variables

Two dependent variables are estimated for each region in each election year, using precinct-level data. First, an estimate of overall manipulation is calculated by regressing the absolute-vote share for the largest pro-Western and pro-Eastern parties (per election) on precinct-level turnout [@myagkov_forensics_2009], using multilevel models with random slopes and intercepts.^[This allows for the turnout coefficient to vary freely by region, an accounts for varying 'baseline' support for each party in the region.] Such a test assumes that, in a clean election, a party’s absolute vote-share (the party’s number of votes divided by the number of registered voters) should increase consistently across precincts within a country, as precinct turnout increases. This approach assumes homogeneous precincts across the country, an assumption which is bolstered by testing precincts within regions or through the use of relevant socioeconomic control variables to weed out potentially confounding relationships [@deckert2013patterns]. I employ both approaches here. Figures \@ref(fig:hist-east) and \@ref(fig:hist-west) show the distribution of coefficients for the major parties. Higher coefficients are more suspicious. While both party groups show some evidence of manipulation, there appear to be more large coefficients overall on the pro-Eastern side.

Second, election falsification is estimated by measuring the degree to which the ones-digit across a party's vote totals deviate from the expected uniform distribution, by region and year. For each region in each election, I calculate a G-statistic, a derivation of a log-likelihood ratio test [@Skovoroda2017]. Whereas the analysis of the Russian case presented in Chapter 4 used a stricter measure---deviance in last-digit zero only---the analysis of Ukraine uses deviance across all digits. This is a more sensitive measure, intended to pose a tougher test of the implication that falsification in Ukraine will be rare and non-systematic. Because a large number of single- or double-digit results for a party can mechanically produce larger deviations from the expected uniform distribution of last digits, for each party I exclude such precincts from the estimation. <!--As an additional robustness check, I also use Beber and Scacco's [-@beber2012] binary indicator of severe manipulation, which takes a value of 1 if last-digit deviance is statistically significant at the p < .05 in a chi-square test. -->
















<!--In sum, Ukraine is a case where many of the conditions are in place for widespread electoral manipulation to occur. Patronage systems are embedded in society, especially in eastern regions of the country. Presidential influence over governors potentially unites the state apparatus in a network that can mobilize patronage resources to deliver votes for the incumbent. There are deep links between the state and business interests [@kudelia_sources_2012]. Nevertheless, the actual level of manipulation has varied considerably between elections and across regions, according to international monitors and experts [@Hale2010;@Herron2010;@Herron2011;@Herron2014;@Katchanovski2008]. This variation can be explained largely by shifts in patronage consolidation. Patronage is substantially less consolidated than in Russia; there is no ruling party akin to United Russia and oligarchic clans compete to influence election results. The constitution has at times fostered competition between patronage networks affiliated with the president and those affiliated with the prime minister. Local constraints are generally higher where pro-Western parties prevail than in pro-Eastern regions, but were elevated generally by the electoral reforms that were in force from the third round of the 2004 presidential election through the 2010 presidential election.

Of the three cases considered here, Ukraine is in several ways the most difficult in which to test the consolidation and constraint model of electoral manipulation. Parties in Ukraine, especially those with roots in its western regions, have tended to be personalized and short-lived, making it more difficult to track their electoral performance over time. Additionally, over the time period for which electoral data are available (2002 to 2014), Ukraine were beset by multiple political crises. The two most significant of these include the 2004 Orange Revolution and the 2014 Euromaidan revolution and subsequent military conflict in the east. These are by no means the only crises to befall the country, however, and it is telling that the so-called Cassette Tape or Kuchma-gate scandal—which implicated President Kuchma in a variety of high-level misdeeds including the abduction and murder of a critical journalist—is considered a lesser incident in recent Ukrainian political history. While these patterns make it more difficult to analyze patterns of electoral manipulation than in Russia, for example, where a single ruling party predominates, it also offers opportunities to test the theory in the context of large shifts in patronage consolidation.-->

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(here)
library(tidyverse)

ukrdata <-read.csv("C:/Users/colej/Documents/Grad school/Dissertation/Ukraine main dataset.csv")  #This dataset is ready for turnout analysis
urban.pct <- ukrdata$urban.population / ukrdata$population
logpop <- log(ukrdata$population)
loggov <- log(ukrdata$gov.employ)
logold <- log(ukrdata$over.sixty)
logincome <- log(ukrdata$income.percap)
west.pres.approval <- ukrdata$west.president * ukrdata$pres.approval
east.pres.approval <- ukrdata$east.president * ukrdata$pres.approval
ukrdata <- cbind(ukrdata, urban.pct, logpop, loggov, logold, logincome, west.pres.approval, east.pres.approval)
ukrdata$prewar.west <- as.factor(ifelse(ukrdata$region == "Ivano-Frankivska" | 
                                ukrdata$region == "Lvivska" |
                                ukrdata$region == "Ternopilska" |
                                ukrdata$region == "Chernivitska" |
                                ukrdata$region == "Zakarpatska" | 
                                ukrdata$region == "Volinska" |
                                ukrdata$region == "Rivnenska", 1, 0)) 

ukrdata$galicia <- ifelse(ukrdata$region == "Ivano-Frankivska" | 
                            ukrdata$region == "Lvivska" |
                            ukrdata$region == "Ternopilska", 1, 0)

ukrdata <- ukrdata %>% mutate(prewar.east = ifelse(region == "Donetska" | region == "Luganska" | region == "Kharkivska", 1, 0))

ukrdata <- ukrdata %>% mutate(PoR_bastion = ifelse(region == "Donetska" | region == "Luganska", 1, 0))

ukrdata <- ukrdata %>% mutate(crimea = ifelse(region == "Crimea" , 1, 0))

ukrdata <- ukrdata %>% mutate(prewar.region = as.factor(ifelse(region == "Crimea", "Crimea", ifelse(prewar.west == "1", "West", 
                                                     ifelse(prewar.east == 1, "East", "Central")))))

ukrdata$party.margin <- NA
ukrdata$party.margin <- ifelse(ukrdata$east.party.margin > 0, ukrdata$east.party.margin, 
                               ((ukrdata$west.party.margin) * -1)) #party.margin: negative for w. parties, positive for e. parties
ukrdata$pres.approval <- ukrdata$pres.approval / 100

ukrdata$post.orange <- NA
ukrdata$post.orange <- ifelse(ukrdata$year >= 2004,
                                    1, 0) #Marks after 2004 protests

ukrdata$orangerev <- NA
ukrdata$orangerev <- ifelse(ukrdata$year == 2004,
                              1, 0) #Marks after 2004 protests

ukrdata$wartime.election <- NA
ukrdata$wartime.election <- ifelse(ukrdata$year == 2014,
                            1, 0) 


ukrdata$pres.constitution <- NA
ukrdata$pres.constitution <- ifelse(ukrdata$year <= 2004 | ukrdata$year > 2010 & ukrdata$year < 2014,
                                    1, 0) #Marks when constitution was more presidential, based on 2004 reform agreement
#ukrdata$crisis.name <- relevel(ukrdata$crisis.name, ref = "None")

ukrdata$east.roz.percent <- ukrdata$e.roz.count.total / ukrdata$n.precincts
ukrdata$west.roz.percent <- ukrdata$w.roz.count.total / ukrdata$n.precincts


#ukrdata$pres.constitution2 <- NA
#ukrdata$pres.constitution2 <- ifelse(ukrdata$year <= 2006 | ukrdata$year > 2010 & ukrdata$year < 2014,
#                                    1, 0) #Marks when constitution was more presidential, based on implementation of reform

fraudscores <- read.csv("C:/Users/colej/Documents/Grad school/Dissertation/Ukraine independent variables/Ukraine fraud scores.csv")

merged.data <- ukrdata %>% left_join(fraudscores, by = c("regionid", "electionid"))

merged.data$president.name <- relevel(as.factor(merged.data$president.name), ref = "Poroshenko")

#Deviance scores
deviancescores <- read.csv(here("Ukraine_forensics", "Data", "region_deviance_ukraine.csv"))

merged.data <- merged.data %>% left_join(deviancescores, by = c("region", "electionid"))

merged.data <- merged.data %>% mutate(east.party.margin2 = east.party.voteshare - west.party.voteshare)

merged.data <- merged.data %>% mutate(electionid = as.factor(electionid))

## Relevel elections to set 2012 as baseline if wanted (otherwise 2002 is baseline)


#merged.data$electionid <- relevel(merged.data$electionid, ref = "2012_0_0")

```

```{r echo=FALSE, message=FALSE, warning=FALSE}
#Adding in total deviance (not just last digit zero deviance)


ukraine_ids <- read.csv("C:\\Users\\colej\\Documents\\Grad school\\Dissertation\\Region ID lists\\ukraine region ids.csv")
east_party_deviance <- read.csv(here::here("Ukraine_forensics", "Data", "ukraine_full_deviance_east.csv"))
west_party_deviance <- read.csv(here::here("Ukraine_forensics", "Data", "ukraine_full_deviance_west.csv"))

east_party_deviance <- east_party_deviance %>% rename(region = group)
west_party_deviance <- west_party_deviance %>% rename(region = group)
ukraine_ids <- ukraine_ids %>% rename(region = name)


east_party_deviance <- east_party_deviance %>% left_join(ukraine_ids, by = "region")
west_party_deviance <- west_party_deviance %>% left_join(ukraine_ids, by = "region")

## Remove unneeded variables for merging

east_party_deviance <- east_party_deviance %>% dplyr::select(!c(dataset, warning, year, presidential, round.2))
west_party_deviance <- west_party_deviance %>% dplyr::select(!c(dataset, warning, total_rows, year, presidential, round.2))

## Merging back in with merged.data

merged.data <- merged.data %>% left_join(east_party_deviance, by = c("region", "electionid"))
merged.data <- merged.data %>% left_join(west_party_deviance, by = c("region", "electionid"))

merged.data <- merged.data %>% mutate(dev_1_9_east = g_statistic_east - deviance_region_east)
merged.data <- merged.data %>% mutate(dev_1_9_west = g_statistic_west - deviance_region_west)

merged.data <- merged.data %>% mutate(year = as.factor(year))


```


```{r echo=FALSE, message=FALSE, warning=FALSE}
## Adding in oligarch count data

ukraine_oligarchs <- read.csv(here::here("Ukraine_forensics", "Data", "ukraine_oligarch_count.csv"))

ukraine_oligarchs <- ukraine_oligarchs %>%
  pivot_longer(
    cols = starts_with("east_oligcount"),
    names_to = "year",
    names_prefix = "east_oligcount_",       # removes "east_count" prefix
    values_to = "east_oligcount"
  )
ukraine_oligarchs <- ukraine_oligarchs %>% mutate(year = as.factor(year))

## Merging back in with merged.data

merged.data <- merged.data %>% left_join(ukraine_oligarchs, by = c("regionid", "year"))
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
## Rada Herfindahl index

merged.data <- merged.data %>% mutate(rada_hh_proeast = ifelse(east.president == 1 & propres.pm.lagged == 1, rada_hh, ifelse(east.president == 0 & propres.pm.lagged == 0, rada_hh, (-1 * rada_hh))))
```

<!--
In particular, the Ukrainian case provides a useful test against the hypotheses of the local-information model of electoral manipulation [@rundlett2016deliver].^[Ukraine is less effective as a test of the signaling model of manipulation, since it does not meet Simpser’s requirements for excessive manipulation to occur: an unconstrained government and a dominant ruling party.] The local-information model claims that when election-manipulating agents perceive their patron’s popularity to be low locally, they extrapolate from this information that their patron is likely to be unpopular nationally and begin to doubt that she will win the election. As a result, they become less likely to engage in electoral manipulation, fearing that patronage rewards will be revoked and the risk of punishment will increase. This theory produces two testable hypotheses that compete with those of the consolidation and constraint model. First, it expects that agents in local strongholds will continue to engage in manipulation on behalf of the locally dominant party, even if national political conditions worsen for that party. As will be shown below, this pattern does not emerge in Ukraine, where manipulation efforts against the pro-eastern Party of Regions decline dramatically—even in pro-western regions—during a period of high patronage consolidation. Second, the local-information model predicts that agents in areas where a party is unpopular will refrain from manipulation on behalf of that party, even if that leader is dominant at the national level; instead, evidence of manipulation in favor of pro-western parties can be seen even in highly pro-eastern regions in the aftermath of the Maidan protests.  

In both cases, the results of election-forensic analysis of data from Ukraine support the consolidation and constraint model, and do not uphold the local information approach. I test these two hypotheses using precinct-level election results for Ukraine’s major pro-eastern party in each election year from 2002 to 2014.^[In most cases this party is the Party of Regions; however in 2002 it the Communist Party of Ukraine, and in 2014 the Opposition Bloc (essentially the rump Party of Regions following the Maidan crisis).] Analysis of the east-party data is more likely to be useful for testing the theory than the pro-western parties. First, the eastern parties—based in the more industrial part of the country—are more strongly rooted in patronage politics than the major pro-western parties. And it is the pro-eastern Party of Regions that provides the only instance during the time period studied where a party gains unambiguously consolidated control over state and para-state patronage resources—the 2012 legislative election during which the Party of Regions holds the presidency and control over the legislature. This control over patronage networks then rapidly deteriorates in the wake of the 2014 Euromaidan protest movement, annexation of Crimea, and prolonged conflict in the Donbass region. This trajectory makes it possible to test the effect of patronage consolidation on east-party results. The pro-western parties do not achieve high levels of patronage consolidation, due to splits between the major pro-western leaders, constitutional reforms, and a societal base of support less dependent on patronage resources.

Lastly, the Ukrainian case also offers an opportunity to test the theory on a different class of manipulation techniques. While the ruling party in Russia appears to mostly engage in inflationary manipulation techniques (at least on election day itself), deflationary tactics appear to predominate in Ukraine. In addition to mobilizing tools like vote-buying or voter pressure, which can only be employed where a principal’s patronage network extends, party’s in Ukraine have also tried to use ‘technical parties’ and ‘clone candidates’ to divert votes away from their genuine opponents. More importantly for this study, such techniques have also allowed parties to stack regional and local electoral commissions with pliant supporters; enabling fraud and other forms of electoral malpractice. Testing the consolidation and constraint model in this context indicates that agents weigh the benefits of patronage against local risks even when employing these kinds of techniques, which may be subtler and harder to observe than tools like ballot-stuffing.

In sum, this chapter shows that deflationary manipulation is characteristic of elections in Ukraine, and that the local information model does not accurately predict those patterns of electoral manipulation. Using two election-forensic techniques, it demonstrates that deflationary techniques deployed against pro-eastern parties and candidates are less likely in regions with higher local constraints against such manipulation, and respond as predicted to shifts in patronage consolidation. That is, during the Yanukovych presidency when patronage resources were highly consolidated around the Party of Regions, evidence of deflationary manipulation against that party drops dramatically, even in highly pro-western regions. In turn, in elections after the Euromaidan crisis, evidence of deflationary manipulation against pro-eastern candidates increases, especially in pro-eastern regions. These results contradict the expectations of the local information model, and support the theory articulated here.





The Ukraine case, with its dramatic political history, provides one unambiguous case of elevated patronage consolidation during the period covered here: the 2012 legislative election, which takes place during the Yanukovych presidency but prior to the Euromaidan crisis. This is the only instance in which the political institutions of the country align with unified government under a patronage-based party machine, the Party of Regions, and without a major political crisis. Following Yanukovych’s victory in 2010, the pro-presidential Ukrainian constitution was restored, replacing the post-Orange constitution that vested greater power in the parliament. Under the pro-parliament constitution, Ukraine was more fully semi-presidential, with the president and prime minister controlling distinct patronage pyramids. This division was especially acute after the 2006 legislative election, which ushered in divided government under President Yushchenko and Prime Minister Yanukovych. The 2004 election, of course, takes place during the Orange Revolution and the attendant collapse of the patronage networks organized around President Kuchma, a collapse which had its roots in the earlier Kuchma-gate crisis. The 2012 election represents the height of patronage consolidation in Ukraine, as well, since the in 2014 the Euromaidan revolution and subsequent conflict results in the complete dissolution of the Party of Regions and a more fragmented political scene. A more detailed description of these dynamics is provided by Hale (2014). This period is captured with a dummy variable **Yanukovych**, which takes on a value of 1 for the 2012 legislative election.

The history sketched above also suggests that the two 2014 elections, which took place after Yanukovych fled the country and in the midst of the collapse of the Party of Regions as a political force, represent the nadir of patronage consolidation from the perspective of pro-eastern candidates. Since this crisis was sparked by the mass pro-Europe protests and anti-Yanukovych demonstrates at Kyiv’s Independence Square (**Maidan Nezalezhnosti**), a dummy variable labelled **Post-maidan** takes on a value of 1 for these two elections. 

-->





```{r hist-east, echo=FALSE, fig.cap="Histogram of turnout coefficients for major pro-East parties (2002-2014)", out.width = '80%'}
ggplot(merged.data, aes(coefs.east)) + geom_histogram(binwidth = .02) + theme_bw() + labs(x = "Turnout coefficients (pro-Eastern)", y = "Count")
```

```{r hist-west, echo=FALSE, fig.cap="Histogram of turnout coefficients for major pro-West parties (2002-2014)", out.width = '80%'}
ggplot(merged.data, aes(coefs.west)) + geom_histogram(binwidth = .02) + theme_bw() + labs(x = "Turnout coefficients (pro-Western)", y = "Count")
```


```{r echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
ggplot(merged.data, aes(g_statistic_east)) + geom_histogram()
ggplot(merged.data, aes(g_statistic_west)) + geom_histogram()

```

### Explanatory variables

In places where pro-Western parties have thin networks and few supporters, it will be easier for pro-Eastern agents to operate without threat of exposure, and vice versa. Echoing the approach from Chapter 4, and of prior work [@harvey_manipulation_2016], I operationalize this risk in part by using the regional legislative margin of victory between the largest parties in regional legislative elections and their largest political rival. This measure is lagged, so that it precedes the national elections that produce estimates of manipulation that serve as dependent variables. Despite Ukraine’s unitary system, elected regional legislatures do provide an opportunity for out-parties to place constraints on the governing party [@Romanova2013]. Margins of victory in regional elections are also indicative of the underlying degree of contestation, which is undergirded by civil society activism, media coverage, oligarchic influence, and other elements of local risk to agents. The variable *Regional pro-East margin* is measured by subtracting the largest pro-Western party's vote-share from the largest pro-Eastern party's vote-share. Larger, positive numbers indicate a more dominant pro-Eastern party, while larger negative values indicate more dominant Western parties. This is a region-level measure.

I also measure local risk to agents using dummy variables that capture the historical differences in Ukraine's regional political cultures. Several oblasts in western Ukraine experienced decades or centuries of Austro-Hungarian rule, which deliberately fostered a different political culture and Ukrainian identity than that developed in areas of Ukraine which remained under Russian imperial rule prior to the Revolution of 1917. These include the historical regions of Galicia, Bukovina, Volhynia. In addition to their Austro-Hungarian legacies, these regions were also not included in Soviet Ukraine until 1939 or later, sparing them from Stalin-era repressions in the 1930s. These *pre-war Western* regions include: Chernivitska, Ivano-Frankivska, Lvivska, Rivnenska, Ternopilska, Volinska, and Zakarpatska oblasts. It is expected that it will be especially difficult for falsification or pro-Eastern manipulation to occur in these regions.

Each Ukrainian president during the time period covered is coded as pro-Western (Yuschenko, Poroshenko) or *pro-Eastern* (Kuchma, Yanukovych).^[The period after Yanukovych's ouster in 2014 is coded as zero, given the pro-Western orientation of the acting presidential administration.] This is a nationwide, year-level measure.

Finally, since vote-buying and voter-pressure (and to a lesser extent, ballot-stuffing) are costly and prone to agency loss, requiring the distribution of resources to agents and voters who are difficult to monitor, such efforts are likely to intensify when one party has more consolidated control over patronage resources. These consolidation is measured in two ways, one regional and one nationwide. At the national level, patronage consolidation is measured using an indicator of the degree to which pro-Western or pro-Eastern coalitions predominate in the national legislature, the Verkhovna Rada. I develop this indicator by first building a a Herfindahl–Hirschman index of the parties in the current legislature for each election year.^[The formula for this index is $HH = \sum_{1}^{x}s_i^2$, where $s_i$ is the size of the *i*-th party as a percentage of the total number of seats in the legislature (expressed as a decimal).] This is common measure of market concentration in economics and policy, and is a component of the better-known (to political scientists) measure of effective number of parties [@golosovEffectiveNumberParties2010]. I utilize the Herfindahl index itself, rather than ENP, because I am interested not in how many major parties are present in the legislature, but how concentrated voting blocs are. I then modify the H-H index to capture the political orientation of the governing coalition. If the prime minister comes from a pro-Eastern party, the H-H index is multiplied by 1; if the prime minister hails from a pro-Western party, the H-H index is multiplied by -1. The resulting variable, *Pro-eastern Rada concentration*, takes on larger positive values when a pro-Eastern majority controls a larger number of seats. It takes on larger, negative values when there is a larger pro-Western majority.

Lastly, I construct a region-level measure of oligarchs' influence, by combining two prior datasets. First, Lankina and Libman [-@lankinaSovietLegaciesEconomic2019] provide information on a set of Ukrainian oligarchs and the regions where their assets are concentrated. This dataset is used to calculate the total number of oligarchs with significant assets in each region. I combine this with Pleines' dataset on oligarchs [-@pleines_oligarchs_2020] to identify each oligarch's political orientation (pro-Eastern or pro-Western) per election year. The variable *pro-East oligarchs* is a count variable indicating the total number of oligarchs in each region-year who are supportive of the major pro-Eastern party. This is, admittedly, a somewhat coarse measure; it cannot capture the intensity of an oligarch's political efforts, for instance. However, it does provide an indicator of the degree to which each region is susceptible to patronage-driven efforts to mobilize voters.
	

### Control variables

A set of control variables are included to help account for other possible explanations for manipulation, or for innocent voting patterns that could appear suspicious using the election forensic tools described above. Several of these are region-level controls. They include the percentage of the population in *government employment*, the percentage of the population *over sixty* years of age, the *average wage* in the region, the *population size (log)*, and the percentage of the population claiming *Ukrainian language* as their primary language. All of these variables are taken from the Ukrainian State Statistical Service. In addition, the models also control for the election-year variable *presidential approval rating*, taken as the three-month average approval rating prior to the election. This variable is collected from data published by the Ukrainian survey firm Rating Group. For models that take the g-statistic as a dependent variable, it is essential to control for the number of precincts used in estimating that statistic; larger sample sizes will mechanically result in larger G-statistics, and will be more likely to produce statistically significant deviations from the uniform.^[See Skovoroda and Lankina [-@Skovoroda2017] for a discussion of this dynamic.] Consequently, in those models, I control for the number of precincts used to calculate the statistic (i.e. the number of precincts where votes for the party exceeded 99).  



## Results



```{r coefmodels, echo=FALSE, include=FALSE, warning=FALSE, message=FALSE}
merged.data <- merged.data %>% mutate(coef_ratio = coefs.east / (coefs.east + coefs.west))
merged.data <- merged.data %>% mutate(over.sixty_pct = over.sixty / (population*1000))
merged.data <- merged.data %>% mutate(govemploy_pct = gov.employ / population)


model.vs_margin_coef <- lm(east.party.margin2 ~ coefs.east + coefs.west  + pres.approval + party.margin + govemploy_pct + over.sixty_pct  + urban.pct + logpop + ukr.lang.pct, data = merged.data)
summary(model.vs_margin_coef)

model.vs_margin_coef2 <- lm(east.party.margin2 ~ coefs.east + pres.approval + party.margin + govemploy_pct + over.sixty_pct  + urban.pct + logpop + ukr.lang.pct, data = merged.data)
summary(model.vs_margin_coef2)

model.coef_margin_pres <- lm(coefs.east ~ party.margin + pres.approval + party.margin*east.president + govemploy_pct + over.sixty_pct + average.wage + logpop + ukr.lang.pct, data = merged.data)
summary(model.coef_margin_pres)
sjPlot::plot_model(model.coef_margin_pres, type = "int")

model.coef_rhh <- lm(coefs.east ~ rada_hh_proeast + pres.approval + east.president  + govemploy_pct + over.sixty_pct + average.wage + logpop + ukr.lang.pct, data = merged.data)
summary(model.coef_rhh)

model.coef_region <- lm(coefs.east ~ prewar.west + pres.approval + east.president +  govemploy_pct + over.sixty_pct + average.wage + logpop + ukr.lang.pct, data = merged.data)
summary(model.coef_region)

model.coef_elex <- lm(coefs.east ~ electionid + govemploy_pct + over.sixty_pct + average.wage + logpop + ukr.lang.pct, data = merged.data)
summary(model.coef_elex)



model.coef_elex_west <- lm(coefs.west ~ electionid + govemploy_pct + over.sixty_pct + average.wage + logpop + ukr.lang.pct, data = merged.data)
summary(model.coef_elex_west)


model.coef_olig <- lm(coefs.east ~ east_oligcount*urban.pct + pres.approval + east.president + govemploy_pct+  over.sixty_pct + average.wage + logpop + ukr.lang.pct, data = merged.data)
summary(model.coef_olig)
sjPlot::plot_model(model.coef_olig, type = "pred", terms = c("urban.pct", "east_oligcount [0, 3]"))


```




```{r gmodels, echo=FALSE, include=FALSE, warning=FALSE, message=FALSE}

model.vs_gs <- lm(east.party.margin2 ~ g_statistic_east + pres.approval + party.margin + govemploy_pct + over.sixty_pct  + logpop + urban.pct + ukr.lang.pct +  sample_size_east, data = merged.data)
summary(model.vs_gs)

model.gs_margin_int <- lm(g_statistic_east ~ party.margin + pres.approval + party.margin*east.president + govemploy_pct + over.sixty_pct + logpop + average.wage + ukr.lang.pct + sample_size_east, data = merged.data)
summary(model.gs_margin_int)
sjPlot::plot_model(model.gs_margin_int, type = "int")

model.gs_margin_int2 <- lm(dev_1_9_east ~ party.margin + pres.approval + party.margin*east.president + govemploy_pct + over.sixty_pct + logpop + average.wage + ukr.lang.pct + sample_size_east, data = merged.data)
summary(model.gs_margin_int2)
sjPlot::plot_model(model.gs_margin_int2, type = "int")


model.gs_margin <- lm(g_statistic_east ~ party.margin + pres.approval + party.margin + govemploy_pct + over.sixty_pct + logpop + average.wage + ukr.lang.pct + sample_size_east, data = merged.data)
summary(model.gs_margin)


model.gs_rhh <- lm(g_statistic_east ~ rada_hh_proeast + pres.approval + east.president  + govemploy_pct + over.sixty_pct + logpop + average.wage + ukr.lang.pct + sample_size_east, data = merged.data)
summary(model.gs_rhh)

model.gs_region <- lm(g_statistic_east ~ prewar.west + pres.approval + east.president +  govemploy_pct + over.sixty_pct + logpop + average.wage + ukr.lang.pct + sample_size_east, data = merged.data)
summary(model.gs_region)

model.gs_elex <- lm(g_statistic_east ~ electionid + govemploy_pct + over.sixty_pct + logpop + average.wage + ukr.lang.pct + sample_size_east, data = merged.data)
summary(model.gs_elex)

model.gs_elex_east <- lm(g_statistic_east ~ electionid + govemploy_pct + over.sixty_pct + average.wage + ukr.lang.pct + sample_size_east + logpop, data = merged.data)
summary(model.gs_elex_east)

model.gs_elex_west <- lm(g_statistic_west ~ electionid + govemploy_pct + over.sixty_pct + average.wage + ukr.lang.pct + sample_size_west + logpop, data = merged.data)
summary(model.gs_elex_west)

model.gs_olig <- lm(g_statistic_east ~ east_oligcount*urban.pct + pres.approval + east.president + govemploy_pct + over.sixty_pct + average.wage + ukr.lang.pct + sample_size_east + logpop, data = merged.data)
summary(model.gs_olig)


merged.data <- merged.data %>% mutate(g_statistic_east_binary = ifelse(p_value_east <= .05, 1, 0))
model_gs_binary <- lm(east.party.margin2 ~ g_statistic_east_binary + pres.approval + party.margin + govemploy_pct + over.sixty_pct  + logpop + urban.pct + ukr.lang.pct +  sample_size_east, data = merged.data)
```




Table \@ref(tab:vsmodels) reports the results of the election-forensic models. It shows that the indicator of overall pro-Eastern manipulation is positively correlated with pro-Eastern vote-share (as expected). Model 1, which does not control for pro-Western parties' turnout coefficients, shows a positive and statistically significant effect. The effect remains more or less constant once pro-Western turnout coefficient is included as a control in Model 2. In either case, the effect size is substantial: a one standard deviance increase in pro-Eastern turnout coefficient is associated with a 6-point increase in the pro-Eastern party's margin of victory. In other words, in an otherwise tied election, increasing the pro-Eastern party's turnout coefficient by one standard deviation would change the outcome of the election to 53 - 47 in their favor. 

Model 3 shows that there is no significant effect for deviance from the uniform distribution of last-digits for pro-East parties. This suggestions that outright falsification of the type captured by this measure is rare in Ukraine across the elections covered in the dataset, at least for major pro-Eastern parties. There is one caveat to this conclusion; following best practice [@beber2012], I excluded precincts from the deviance calculations where fewer than 100 votes were counted in that party's favor. A high proportion of precincts with double-digit vote totals will increase the deviance from the uniform distribution naturally, even if the votes are fair. This proportion is extremely high in some of Ukraine's more western regions. For instance, in the 2002 election in Ivano-Frankivska oblast', all but 24 out of 958 precincts are excluded for having fewer than 100 votes for the pro-Eastern party. In 2014, all 914 precincts in Ivano-Frankivska are dropped, as are all precincts in L'vivska, Ternopilska, Vinnitska, Ternopilska, and Cherkaska oblasts. When the total number of available precincts falls below 50 for the reason, I exclude the region from analsys. This results in a smaller number of observations for the G-statistic models, and highlights the difficulty of conducting this sort of analysis in Ukraine's highly polarized and frequently crisis-driven politics under the period of study. 

As an additional check, I also utilize a dummy variable to indicate regions where the g-statistic indicator of fraud is statistically significant, with results presented in Model 4. The results show that a statistically significant pro-Eastern G-statistic is unrelated to pro-Eastern vote-share. This suggests that deviation from the uniform distribution is largely random, a possibility which is bolstered by the fact that deviations are statistically significant at the five percent level in only `r sum(merged.data$g_statistic_east_binary, na.rm = T)` of `r nrow(merged.data)` cases---in other words, in roughly five percent of cases. All told, this supports the expectation that falsificaiton is uncommon, and allows us to interpret the 'overall manipulation' turnout coefficient as primarily reflective of dispersed forms of election manipulation.




```{r vsmodels, echo=FALSE, message=FALSE, warning=FALSE}
library(modelsummary)
models <- list()
models[['Model 1']] <- model.vs_margin_coef2
models[['Model 2']] <- model.vs_margin_coef
models[['Model 3']] <- model.vs_gs
models[['Model 4']] <- model_gs_binary



library(kableExtra)

reg_table_vs <- modelsummary(models, output = "kableExtra",  stars = T, title = 'OLS models of pro-Eastern party vote-share', coef_map = c("(Intercept)" = "(Intercept)",
             "coefs.east" = "Pro-Eastern turnout coefficient",
             "coefs.west" = "Pro-Western turnout coefficient",
             "g_statistic_east" = "Pro-East party last-digit deviance",
             "g_statistic_east_binary" = "Pro-East last-digit deviance (binary)",

             "pres.approval" = "Presidential approval rating",
             "party.margin" = "Pro-Eastern margin in regional election",
             "govemploy_pct" = "Gov. employees",
             "over.sixty_pct" = "Over-sixty pop. share",
             "average.wage" = "Average wage",
             "logpop" = "Population",
             "ukr.lang.pct" = "Pct. Ukrainian language",
             "sample_size_east" = "N. precincts > 99 pro-East votes"))

reg_table_vs





```

### Predictors of manipulation

Table \@ref(tab:manip-models) shows the results of models that take pro-Eastern turnout coefficients (Models 5 through 7) as the dependent variable. Model 5 tests the interaction of region-level partisan control and the presence of a pro-Eastern president. As the table and Figure \@ref(fig:coef-fig) show, there is a statistically significant interaction for overall manipulation. This is in line with expectations---it is in the most heavily pro-Eastern regions where patronage networks are most fully developed, owing to the region's Soviet heritage and post-Soviet accumulation of politically active oligarchs. When a pro-Eastern candidate occupies the presidency, the resources of the central state can be used to kick regional networks into higher gear. Where the risk of exposure is high, however, having a supportive president in office does not yield additional manipulation. This supports H3.


```{r manip-models, echo=FALSE, message=FALSE, warning=FALSE}
models <- list() 
models[['Model 5']] <- model.coef_margin_pres
models[['Model 6']] <- model.coef_rhh
models[['Model 7']] <- model.coef_region
#models[['Model 8']] <- model.gs_margin_int
#models[['Model 9']] <- model.gs_rhh
#models[['Model 10']] <- model.gs_region


table1 <- modelsummary(
  models,
  output = "kableExtra",
  title = "Models of pro-Eastern election manipulation (turnout coefficient)",
coef_map = c("(Intercept)" = "(Intercept)",
             
             "pres.approval" = "Presidential approval rating",
             
             "govemploy_pct" = "Gov. employees",
             "over.sixty_pct" = "Over-sixty pop. share",
             "average.wage" = "Average wage",
             "logpop" = "Population",
             "ukr.lang.pct" = "Pct. Ukrainian language",
             "sample_size_east" = "N. precincts > 99 pro-East votes",
             "party.margin" = "Pro-Eastern margin in regional election",
             "rada_hh_proeast" = "Pro-East Rada consolidation",
             "prewar.west1" = "Pre-war West",
             "east.president" = "Pro-East president",
             "party.margin:east.president" = "Regional margin : Pro-East pres."
             ),
  gof_map = c("nobs", "r.squared", "adj.r.squared"),
  stars = c('*' = .1, '**' = .05, '***' = .01)
) 

table1
```


```{r coef-fig, echo=FALSE, message=FALSE, warning=FALSE, fig.cap= "Overall pro-eastern manipulation and patronage consolidation"}
intplot_coefs <- sjPlot::plot_model(model.coef_margin_pres, type = "pred", terms = c("party.margin", "east.president")) + labs(x = "Region-level pro-East party margin", y = "Pro-East turnout coefficient", title = "Predicted values", color = "Pro-East president")

intplot_coefs + theme_bw()
```

```{r gs-fig, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Pro-eastern fraud and patronage consolidation", include=FALSE}
intplot_gs <- sjPlot::plot_model(model.gs_margin_int, type = "pred", terms = c("party.margin", "east.president")) + labs(x = "Region-level pro-East party margin", y = "Pro-East g-statistic", title = "Predicted values", color = "Pro-East president")

intplot_gs + theme_bw()
```

The same pattern is observed for pro-Eastern consolidation of the legislature (Model 6) and for the pre-war West dummy variable (Models 7). Pro-Eastern consolidation in the Rada is associated with increased turnout coefficients in subsequent elections. Likewise, regions in the pre-ar West show reduced turnout coefficients for Eastern parties. This supports H1b and H4. In all models, manipulation appears to increase when a pro-Eastern president is in office, supporting H1a.


```{r olig-models, echo=FALSE, message=FALSE, warning=FALSE}
models <- list() 
models[['Model 8']] <- model.coef_olig
#models[['Model 9']] <- model.gs_olig



table3 <- modelsummary(
  models,
  output = "kableExtra",
  title = "Regression Results",
coef_map = c("(Intercept)" = "(Intercept)",
             
             "pres.approval" = "Presidential approval rating",
             
             "govemploy_pct" = "Gov. employees",
             "over.sixty_pct" = "Over-sixty pop. share",
             "average.wage" = "Average wage",
             "logpop" = "Population",
             "ukr.lang.pct" = "Pct. Ukrainian language",
             "sample_size_east" = "N. precincts > 99 pro-East votes",
             "urban.pct" = "Urban pop. share",
             "east_oligcount" = "Pro-Eastern oligarch count",
             "east_oligcount:urban.pct" = "Pro-East oligarchs : Urban pct."
             ),
  gof_map = c("nobs", "r.squared", "adj.r.squared"),
  stars = c('*' = .1, '**' = .05, '***' = .01))  %>%
  # Add model numbers as column names
  #`colnames<-`(c("", "(1)", "(2)", "(3)", "(4)", "(5)", "(6)")) %>%
  # Add grouped headers for dependent variables
  add_header_above(c(" " = 1, "Pro-East turnout coef." = 1))


table3
```



```{r olig-fig, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Pro-eastern manipulation by urban vote-share and number of supportive oligarchs"}
plot_oligs_coef <- sjPlot::plot_model(model.coef_olig, type = "pred", terms = c("urban.pct", "east_oligcount [0, 3]")) + labs(x = "Urban pop. share", y = "Pro-East turnout coefficient", title = "Predicted values", color = "Pro-Eastern oligarchs \n(region-level)")
plot_oligs_coef + theme_bw()
```

Next, Table \@ref(tab:olig-models) reports the results of models that include measures of oligarchic influence. Figure \@ref(fig:olig-fig) shows the interaction effect from Model 8. Consistent with the prior results, these models show that pro-Eastern overall manipulation is improved by patronage consolidation: when more oligarchs are supporting the pro-Eastern party in more urbanized regions, the party's turnout coefficient improves. As Figure \@ref(fig:olig-fig) shows, this effect is most substantial in regions with a high concentration of politically active oligarchs and a very high degree of urbanization. The effect size for such regions is large; turnout coefficients practically double in highly urban regions when oligarchs get involved. The results support H2a and H2b.




```{r yearmodels, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
models <- list()
models[['Model 1']] <- model.coef_elex
models[['Model 2']] <- model.gs_elex


library(kableExtra)

reg_table_year <- modelsummary(models, output = "kableExtra",  stars = T, title = 'Table 3: OLS models of election manipulation by year', coef_map = c("(Intercept)" = "(Intercept)",
             "electionid2004_1_0" = "2004 (R3)",
             "electionid2006_0_0" = "2006",
             "electionid2007_0_0" = "2007",
             "electionid2010_1_0" = "2010 (R1)",
            "electionid2010_1_1" =  "2010 (R2)",
            "electionid2012_0_0" = "2012",
            "electionid2014_0_0" = "2014 (Leg.)",
            "electionid2014_1_0" =  "2014 (Pres.)",
             "gov.employ" = "Gov. employees",
             "over.sixty_pct" = "Over-sixty pop. share",
             "average.wage" = "Average wage",
             "population" = "Population",
                         "ukr.lang.pct" = "Pct. Ukrainian language",
             "sample_size_east" = "N. precincts > 99 pro-East votes"
             ))

reg_table_year



```





## Discussion and conclusion

Despite the difficulties of the case, Ukraine presents an important opportunity to test the implications of the model, especially in comparison with Russia. Russia shares many institutional and historical features with Ukraine (particularly in eastern parts of the country) as a result of a shared imperial and Soviet experience. However, patterns of patronage consolidation and local constraints vary considerably between the two states. Patronage consolidation has consistently been much lower in Ukraine than in Russia during the period of analysis. A single party of power largely dominates in Russia, while the party system in Ukraine is much more fragmented and fluid. Oligarchic interests were largely tamed during the early years of the first Putin administration; rival oligarchic clans continue to clash behind the scenes in Ukraine. The period in which Ukraine’s parliamentary-presidential constitution was in force helped establish a two-pyramid structure of patronage, in Hale’s (2014) terminology, while Russia continues to operate within a more consolidated vertical of power.

As a result, we should expect to see different patterns of manipulation occur in the two cases. First, while the direct comparison is not statistically tested here, widespread falsification does appear to be considerably less common in Ukraine than in Russia. In the Russian case, suspicious results were captured by the digits test in 24 percent of region-years, whereas in Ukraine the same figure is approximately five percent—roughly what would be expected due to random chance alone.^[Note that the digits test results differ from those of the vote-share test presented here; the mean number of suspicious precincts per region using the vote-share test is approximately 10 percent.] 


One peculiar feature of election management in Ukraine is an apparent emphasis on deflationary tactics; that is, efforts to reduce one’s opponents vote-share rather than (or in addition to) inflating one’s own. Ukrainian parties have attempted to influence election results by using manufactured parties and candidates to drain votes away from rivals and to gain additional influence on election commissions. Ukrainian elections have often been characterized by an abundance of so-called ‘technical parties,’ small parties artificially created by main parties in order to divert votes away from their opponents. One such technical party, which voiced strong opposition to Yanukovych and the Party of Regions and whose leader was reminiscent of former Orange-Revolution hero Yulia Tymoshenko, diverted as much as 300,000 votes from the nationalist party UDAR, depriving it of six to eight seats in parliament [@kovalov_electoral_2014]. 

The specifics of the Ukrainian case---robust but fragmented patronage networks combined with a high risk of exposure for illegal manipulation---may help explain this phenomenon. The high risk of exposure makes it difficult to recruit agents for attributable, but cost-effective, election fraud. Widespread voter mobilization and voter pressure campaigns can make up for this, in part, but are costly and unreliable tools not available to all patrons; they require large patronage networks and carry principal-agent risks. By contrast, cheaper deflationary tactics like clone candidates may offer a more predictable return on investment, without requiring anyone to expose themselves to legal risk.


However, the main benefit of the case in the present study is derived from the rapid fluctuation in patronage consolidation over time. Ukraine shifts from a fragmented patronage system after the Orange Revolution, to one with considerable consolidation around the Party of Regions under President Yanukovych, and subsequently back to a fragmented system after the 2014 crisis. Throughout these changes, risks to agents are largely structural---based on historical legacies, active civil society, political opposition, and unpredictable penetration of the criminal-legal system by elite factions. As a result, agent risk remains consistent across most regions over time, and is elevated compared to risk for pro-regime agents in many Russian regions. 

The results shown here indicate that outright falsification remains rare throughout these changes, while dispersed forms of electoral manipulation remain common. These more dispersed techniques, like voter pressure and vote-buying, survive in a higher-risk environment by being less attributable to agents. They are costly, however, and are most effective when Ukraine's patronage pyramids come into alignment. In conjunction with the Russian case in Chapter 4, this study shows that the micro-level predictions of the regime-strength model are borne out across varying levels of patronage consolidation and agent risk. In Chapter 6, we turn to the macro-level predictions of the theory regarding protest survival.



<!--
---
Current thoughts:

1) There is not a lot of extreme falsification in Ukraine. Rozenas spike score at the 90th percentile in Ukraine is 1.2%. In Russia it's 2.5%. Deviance is pretty similar to Russia though...at least for east parties.  Non-extreme distributions are pretty similar.
2) What do we know?
  A) Increasing pro-east Rada consolidation -> increased east-party vote-share in the Central regions, and a little in the East, but not in the West.
  B) Increasing pro-east overall turnout coefficient ratio -> eastern vote-share. (Ratio is needed because both parties can manipulate)
  C) Rozenas percent (i.e. spikes) is associated with increased east party vote-share in the East and the Center, but not the West.
  D) Increasing pro-east Rada consolidation leads to increased overall pro-East manipulation in the West and Center, but not the East.
  E) There is high east Rozenas score in the pre-war West region when the Rada is more pro-Western, declining as PoR rises. No change across other regions.
3) Putting this all together:






OPORA trained police officers in 2019: https://www.oporaua.org/en/vybory/17992-pid-chas-prezidentskikh-viboriv-treneri-opori-navchili-bilshe-1-5-tisiachi-politseiskikh-17992 

Court cases can be found here: https://reyestr.court.gov.ua/
Look for Criminal cases, under the criminal code, then violations of electoral rights. Then there are subcategories like bribery of voters (pidkup vybortsya) and falsification. Legal code is here: https://sherloc.unodc.org/cld/uploads/res/document/ukr/2001/criminal-code-of-the-republic-of-ukraine-en_html/Ukraine_Criminal_Code_as_of_2010_EN.pdf

Ухвала is decision, Постанова  is decree, Вирок is sentence.

This could be used to provide some qualitative support for the argument about risk.

-->