---
title: "Digits tests"
output: html_notebook
---

```{r setup}
library(tidyverse)
library(stringi)
library(here)
library(openxlsx)

filenames <- list.files(here("Russia_forensics", "Data"), pattern="*.xlsx", full.names=TRUE)
filenames_sub <- filenames[-10] #TODO: Fix file structure so that covariate data is not in the same folder


```

In the precinct election data, all KPRF results (legislative and presidential) are called 'kprf'. All regime results are coded as "yed_ro" (so, United Russia, Putin/Medvedev, and YES on the constitutional referendum.) NO on the referendum is still has its original name.  When the code generates NAs for region_id, those observations are polling places outside the RF.

```{r}
p<-as.vector(rbind(.1, .1, .1, .1, .1, .1, .1, .1, .1, .1)) # Expected uniform distribution
year.list <- c("2000", "2003", "2004", "2007", "2008", "2011", "2012", "2016", "2018", "2021") #Do referendum separately
fraud_scores <- data.frame(ur_chi = NA, kprf_chi = NA, region_id = NA, year = NA, ur_fraud = NA, kprf_fraud = NA)

for(i in 1:length(filenames_sub)){
  elex_data <- read.xlsx(filenames_sub[i])
  n.regions <- length(unique(elex_data$region_id))
  elex_data <- elex_data %>% mutate(kprf_digit = kprf %% 10)
  elex_data <- elex_data %>% mutate(ur_digit = yed_ro %% 10)
  regionids <- unique(elex_data$region_id)
  chivalues<-matrix(NA, nrow=n.regions, ncol=4) #Four columns: chi values for two parties, year, and regionid

  for (j in 1:length(regionids)){
    grouping_id <- regionids[j]
  group<-subset(elex_data, region_id==grouping_id)
  table1<-table(group$ur_digit)
  table2<-table(group$kprf_digit)
  
  if (nrow(table1) == 10 & nrow(table2) == 10){
  chi1<-chisq.test(x=table1, p=p)
  chi2<-chisq.test(x=table2, p=p)
  
  chivalues[j, ]<-rbind(chi1$p.value, chi2$p.value, grouping_id, year.list[i])  
  }
  if (nrow(table1) < 10 & nrow(table2) == 10) {
    chivalues[j, ]<-rbind(0, chi2$p.value, grouping_id, year.list[i])  
  }
   if (nrow(table1) == 10 & nrow(table2) < 10) {
    chivalues[j, ]<-rbind(chi1$p.value, 0, grouping_id, year.list[i])  
   }
  if (nrow(table1) < 10 & nrow(table2) < 10)  {
    chivalues[j, ]<-rbind(0, 0, grouping_id, year.list[i])  
  }
  
 }
 chivalues <- data.frame(chivalues) 
 chivalues <- chivalues %>% mutate(ur_fraud = ifelse(chivalues[,1] < .05, 1, 0))
 chivalues <- chivalues %>% mutate(kprf_fraud = ifelse(chivalues[,2] < .05, 1, 0))
fraud_scores <- bind_rows(fraud_scores, chivalues)
}
####

fraud_scores <- fraud_scores %>% select(-ur_chi, -kprf_chi, -region_id, -year)
fraud_scores <- fraud_scores %>% rename(ur_chi = X1, kprf_chi = X2, region_id = X3, year = X4)


write.csv(fraud_scores, here("Russia_forensics", "Data", "russia_elex_fraudscores.csv"))

```

The referendum:

```{r}
p<-as.vector(rbind(.1, .1, .1, .1, .1, .1, .1, .1, .1, .1)) # Expected uniform distribution
fraud_scores <- data.frame(ur_chi = NA, kprf_chi = NA, region_id = NA, year = NA, ur_fraud = NA, kprf_fraud = NA)

  elex_data <- read.xlsx(filenames[10])
  n.regions <- length(unique(elex_data$region_id))
  elex_data <- elex_data %>% mutate(ur_digit = yed_ro %% 10)
  regionids <- unique(elex_data$region_id)
  chivalues<-matrix(NA, nrow=n.regions, ncol=3) #Four columns: chi values for yes result, year, and regionid

  for (j in 1:length(regionids)){
    grouping_id <- regionids[j]
  group<-subset(elex_data, region_id==grouping_id)
  table1<-table(group$ur_digit)

  if (nrow(table1) == 10){
  chi1<-chisq.test(x=table1, p=p)

  chivalues[j, ]<-rbind(chi1$p.value, grouping_id, "2020")  
  }
  if (nrow(table1) < 10) {
    chivalues[j, ]<-rbind(0, grouping_id, "2020")  
  }
  
 }
 chivalues <- data.frame(chivalues) 
 chivalues <- chivalues %>% mutate(ur_fraud = ifelse(chivalues[,1] < .05, 1, 0))
fraud_scores <- bind_rows(fraud_scores, chivalues)

####

fraud_scores <- fraud_scores %>% select(-ur_chi, -kprf_chi, -region_id, -year, -kprf_fraud)
fraud_scores <- fraud_scores %>% rename(ur_chi = X1, region_id  = X2, year = X3)

write.csv(fraud_scores, here("Russia_forensics", "Data", "russia_2020ref_fraudscores.csv"))

```

